---
title: "Logistic Regression"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
  enabled: false
---

# Housekeeping

```{r}
#| warning: false

cat("\014")  # clear the console when run interactively
rm(list = ls())  # begin from a clean workspace

if (!requireNamespace("here", quietly = TRUE)) {
  stop("Install the 'here' package before rendering this document.")
}

suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Background

- Suppose you manage an actively selected equity portfolio.
- You want to identify companies at elevated risk of experiencing a bad earnings quarter.
- You use currently available financial ratios and sector information to estimate that probability.
- The predicted probability serves as a portfolio risk signal. Companies with sufficiently high predicted probabilities may receive additional research, be underweighted, or be excluded from the portfolio.
- We will estimate a simple logistic regression and then a model containing all available predictors.
- We will reserve a test sample for final evaluation. All data-dependent choices, including standardization and the definition of the target, will be based on the training sample.
- We assume that all predictors are available to the investor before the earnings quarter being predicted.

# Data

## Acquisition

Task: Load the data dictionary from the `Data_Dictionary` sheet of `earnings_ratio_data.xlsx`.

```{r}
data_dictionary <- readxl::read_excel(
  here("data", "?"),
  sheet = "?"
)

data_dictionary
```

Q: Before loading the data, what signs and relative predictive strength do you expect from the variables?

A: ?

Task: Load the data and inspect its structure.

```{r}
earnings_data <- readxl::read_excel(
  here("data", "?"),
  sheet = "?"
)

dplyr::glimpse(earnings_data)
```

## Clean

Task: Determine the dimensions of the dataset.

```{r}
?(earnings_data)
```

Task: Confirm the number of rows and unique companies.

```{r}
earnings_data %>%
  summarise(
    rows = n(),
    unique_companies = n_distinct(?),
    duplicated_company_ids = ?
  )
```

Task: Count the total number of missing cells.

```{r}
sum(?(earnings_data))
```

Task: Count the number of companies with and without at least one missing value.

```{r}
earnings_data %>%
  ?(
    companies_with_missing = sum(if_any(everything(), is.na)),
    companies_without_missing = sum(!if_any(everything(), is.na))
  )
```

Task: Retain a copy of the original data and drop companies with any missing values.

```{r}
earnings_data_raw <- earnings_data

earnings_data <- earnings_data %>%
  ?(!if_any(everything(), is.na))
```

Complete-case deletion is convenient for this exercise. In an applied project, we would first investigate why the values are missing and consider whether deletion could create bias.

Task: Determine the dimensions of the remaining dataset.

```{r}
?(earnings_data)
```

## Split the Sample

Q: Why might we not want to use the entire dataset to train our model?

A: ?

Task: Randomly assign 80% of the companies to the training sample and 20% to the test sample.

```{r}
set.seed(123)

train_companies <- earnings_data %>%
  distinct(Company_ID) %>%
  slice_sample(prop = ?) %>%
  pull(Company_ID)

train_data_raw <- earnings_data %>%
  filter(Company_ID %in% ?)

test_data_raw <- earnings_data %>%
  filter(!Company_ID %in% ?)

sample_sizes <- tibble(
  Sample = c("Training", "Test"),
  Rows = c(nrow(train_data_raw), nrow(test_data_raw)),
  Companies = c(
    n_distinct(train_data_raw$Company_ID),
    n_distinct(test_data_raw$Company_ID)
  )
)

knitr::kable(sample_sizes)
```

## Define “Bad” Earnings

Q: How might we define a bad earnings quarter?

A: ?

Task: Plot the training-sample distribution of quarterly earnings growth.

```{r}
ggplot(?, aes(x = ?)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Training-Sample Distribution of Quarterly Earnings Growth",
    x = "Quarterly Earnings Growth",
    y = "Count"
  ) +
  theme_minimal()
```

Task: Generate a detailed statistical summary of quarterly earnings growth in the training sample.

```{r}
earnings_summary <- train_data_raw %>%
  summarise(
    n = n(),
    min = min(Quarterly_Earnings_Growth),
    q10 = quantile(Quarterly_Earnings_Growth, 0.10),
    q25 = quantile(Quarterly_Earnings_Growth, 0.25),
    median = ?(Quarterly_Earnings_Growth),
    mean = mean(Quarterly_Earnings_Growth),
    stdev = sd(Quarterly_Earnings_Growth),
    q75 = quantile(Quarterly_Earnings_Growth, 0.75),
    q90 = quantile(Quarterly_Earnings_Growth, 0.90),
    max = max(Quarterly_Earnings_Growth)
  )

knitr::kable(earnings_summary, digits = 3)
```

Task: Let's use the 15%-quantile as the cut off for bad earnings. Create a `params` list containing the target quantile and a corresponding earnings-growth cutoff.

```{r}
params <- list(
  bad_quantile = ?,
  bad_threshold = as.numeric(
    quantile(
      train_data_raw$Quarterly_Earnings_Growth,
      probs = ?
    )
  )
)

params
```

Q: What does `bad_threshold` mean?

A: ?

Task: Create the observed target variable in both samples. Set it equal to 1 for a bad earnings quarter and 0 otherwise. Then tabulate the fraction of observations that are "Bad" in each sample.

```{r}
train_data_raw <- train_data_raw %>%
  mutate(
    Bad_Earnings = as.integer(
      Quarterly_Earnings_Growth <= ?
    )
  )

test_data_raw <- ? %>%
  mutate(
    Bad_Earnings = as.integer(
      Quarterly_Earnings_Growth <= ?
    )
  )

tibble(
  Sample = c("Training", "Test"),
  Bad_Earnings_Rate = c(
    mean(train_data_raw$Bad_Earnings),
    mean(test_data_raw$Bad_Earnings)
  )
) %>%
  knitr::kable(digits = 3)
```

Task: Add a classification-probability threshold of 65% to `params`.

```{r}
params$prob_threshold <- ?
params
```

Q: What is the conceptual difference between `bad_threshold` and `prob_threshold`?

A: ?

Q: What are other approaches to selecting the classification-probability threshold?

A: C?

## Select and Standardize the Predictors

Task: Create objects containing the names of all predictors, numeric predictors, and categorical predictors.

```{r}
excluded_names <- c(
  "Company_ID",
  "Quarterly_Earnings_Growth",
  "Bad_Earnings"
)

feature_names <- setdiff(
  names(train_data_raw),
  excluded_names
)

numeric_feature_names <- train_data_raw %>%
  select(all_of(feature_names)) %>%
  select(where(is.numeric)) %>%
  names()

categorical_feature_names <- ?(
  feature_names,
  numeric_feature_names
)

feature_names
numeric_feature_names
categorical_feature_names
```

Task: Compute each numeric predictor's mean and standard deviation using only the training sample.

```{r}
feature_means <- train_data_raw %>%
  ?(across(all_of(numeric_feature_names), mean)) %>%
  unlist(use.names = TRUE)

feature_sds <- train_data_raw %>%
  summarise(across(all_of(numeric_feature_names), ?)) %>%
  unlist(use.names = TRUE)

zero_variance_features <- names(feature_sds[feature_sds == 0])

if (length(zero_variance_features) > 0) {
  stop(
    "These predictors have zero variance: ",
    paste(zero_variance_features, collapse = ", ")
  )
}
```

Task: Standardize the numeric predictors in both samples using the training-sample means and standard deviations.

```{r}
standardize_predictors <- function(data) {
  data %>%
    mutate(
      across(
        all_of(numeric_feature_names),
        ~ (.x - feature_means[[cur_column()]]) /
          ?[[cur_column()]]
      )
    )
}

train_data <- standardize_predictors(train_data_raw)
test_data <- standardize_predictors(test_data_raw)
```

Q: Why don't we standardize Quarterly_Earnings_Growth?

A: ?

# Exploratory Data Analysis

The EDA below uses only the training sample. The test sample remains reserved for out-of-sample evaluation.

## Target Variable

Task: Create a bar chart showing the number of good and bad earnings quarters.

```{r}
train_data %>%
  dplyr::count(Bad_Earnings) %>%
  mutate(
    Earnings_Class = if_else(
      Bad_Earnings == ?,
      "Bad Earnings",
      "Good Earnings"
    )
  ) %>%
  ggplot(aes(x = Earnings_Class, y = n, fill = Earnings_Class)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Training-Sample Earnings Classes",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

Task: Construct a table containing counts and percentages for good and bad earnings quarters.

```{r}
earnings_proportions <- train_data %>%
  dplyr::count(Bad_Earnings) %>%
  mutate(
    Earnings_Class = if_else(
      Bad_Earnings == 1,
      "Bad Earnings",
      "Good Earnings"
    ),
    Percent = ?
  ) %>%
  select(Earnings_Class, Count = n, Percent)

knitr::kable(
  earnings_proportions,
  digits = 3,
  col.names = c("Earnings Class", "Count", "Percent")
)
```

Q: Any concerns?

A: ?

Task: Compute and plot the bad earnings rate by sector. Also produce a table with the Sector, Number of Companies, Bad Earnings Count, and Bad Earnings Rate.

```{r}
sector_bad_rates <- train_data %>%
  group_by(?) %>%
  summarise(
    Companies = n(),
    Bad_Earnings_Count = ?(Bad_Earnings),
    Bad_Earnings_Rate = ?(Bad_Earnings),
    .groups = "drop"
  ) %>%
  arrange(desc(Bad_Earnings_Rate))

knitr::kable(sector_bad_rates, digits = 3)

sector_bad_rates %>%
  ggplot(
    aes(
      x = reorder(Sector, Bad_Earnings_Rate),
      y = Bad_Earnings_Rate
    )
  ) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Bad Earnings Rate by Sector",
    x = NULL,
    y = "Bad Earnings Rate"
  ) +
  theme_minimal()
```

Q: How might the class imbalance by sector impact our analysis?

A: ?

## Predictors

Task: Compute and plot the correlations among the numeric predictors.

```{r}
feature_correlations <- train_data %>%
  select(all_of(numeric_feature_names)) %>%
  ?()

corrplot(?)
```

Q: What might the correlation matrix suggest about our modeling efforts?

A: ?

## Connect a Predictor to the Target

Let's use Gross Profit Margin (GPM) has a primary Predictor. Note: We're not suggesting this is the most important predictor, but I want to reduce the scope for pedagogical ease. We'll expand to more predictors later.

Task: For each sector, compare mean standardized GPM between good and bad earnings quarters. Also compute an exploratory two-sample t-test when each group has at least two observations.

```{r}
sector_gpm_comparison <- train_data %>%
  ?(Sector) %>%
  summarise(
    Good_Count = sum(Bad_Earnings == 0),
    Bad_Count = sum(?),
    Mean_GPM_Good = if (
      sum(Bad_Earnings == 0) > 0
    ) {
      mean(GPM[Bad_Earnings == 0])
    } else {
      NA_real_
    },
    Mean_GPM_Bad = if (
      sum(Bad_Earnings == 1) > 0
    ) {
      mean(GPM[Bad_Earnings == 1])
    } else {
      NA_real_
    },
    P_Value = if (
      sum(Bad_Earnings == 0) >= 2 &&
      sum(Bad_Earnings == 1) >= 2
    ) {
      ?(
        GPM[Bad_Earnings == 0],
        GPM[Bad_Earnings == 1]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(
    Difference = Mean_GPM_Good - Mean_GPM_Bad
  ) %>%
  select(
    Sector,
    Good_Count,
    Bad_Count,
    Mean_GPM_Good,
    Mean_GPM_Bad,
    Difference,
    P_Value
  ) %>%
  arrange(desc(abs(Difference)))

knitr::kable(sector_gpm_comparison, digits = 3)
```

Q: What's the implication of the Differences and PValues?

A: ?

Q: Why should we be cautious about these p-values?

A: ?

Task: Isolate the sector with the largest absolute difference and create side-by-side boxplots with individual observations.

```{r}
largest_difference_sector <- sector_gpm_comparison %>%
  filter(!is.na(Difference)) %>%
  slice_max(
    order_by = ?,
    n = 1,
    with_ties = FALSE
  ) %>%
  pull(Sector)

train_data %>%
  filter(Sector == ?) %>%
  mutate(
    Earnings_Class = if_else(
      Bad_Earnings == 1,
      "Bad Earnings",
      "Good Earnings"
    )
  ) %>%
  ggplot(
    aes(
      x = Earnings_Class,
      y = GPM,
      fill = Earnings_Class
    )
  ) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    show.legend = FALSE
  ) +
  geom_jitter(width = 0.08, alpha = 0.50) +
  labs(
    title = paste(
      "Gross Profit Margin by Earnings Class:",
      largest_difference_sector
    ),
    x = NULL,
    y = "Standardized Gross Profit Margin"
  ) +
  theme_minimal()
```

Q: Any concerns?

A: ?

# Base Model

## Estimate

Task: Estimate a logistic regression on the training sample using gross profit margin as the only predictor.

```{r}
#| warning: true

base_model <- ?(
  ? ~ ?,
  data = ?,
  family = binomial(link = "?")
)
```

## Interpret

Task: Display the model output.

```{r}
?(base_model)
```

Q: How interpret the coefficient on GPM?

A: ?

Q: Consider a Consumer Staples company. Suppose its starting probability of bad earnings is equal to the observed bad-earnings rate for Consumer Staples. How would a one-standard-deviation increase in GPM affect this probability?

A: ?

## Predict

Task: Predict the probability of bad earnings for each company in the test sample.

```{r}
test_data <- test_data %>%
  mutate(
    base_probability = ?(
      base_model,
      newdata = ?,
      type = "?"
    )
  )

head(
  test_data %>%
    select(Company_ID, Bad_Earnings, base_probability)
)
```

Task: Predict whether a company will have "bad earnings" for two hypothetical companies: one with GPM at the first quartile of the training sample and one with GPM at the third quartile.

```{r}
gross_margin_quantiles <- quantile(
  train_data$GPM,
  probs = ?
)

base_profiles <- tibble(
  Company = c(
    "First-Quartile Gross Margin",
    "Third-Quartile Gross Margin"
  ),
  GPM = as.numeric(gross_margin_quantiles)
) %>%
  mutate(
    Predicted_Probability = predict(
      base_model,
      newdata = .,
      type = "response"
    ),
    Predicted_Class = as.integer(
      Predicted_Probability >= params$prob_threshold
    )
  )

knitr::kable(base_profiles, digits = 3)
```

## Classify

Task: Convert the test-sample probabilities into predicted classes using the classification threshold defined earlier.

```{r}
test_data <- test_data %>%
  mutate(
    base_prediction = as.integer(
      ? >= ?
    )
  )

head(
  test_data %>%
    select(
      Company_ID,
      Bad_Earnings,
      base_probability,
      base_prediction
    )
)
```

Task: Construct a confusion matrix. Rows are actual classes and columns are predicted classes.

```{r}
base_confusion <- table(
  Actual = factor(test_data$Bad_Earnings, levels = c(0, 1)),
  Predicted = factor(?, levels = c(?, ?))
)

base_confusion
```

Q: How do we interpret this confusion matrix?

A: ?

Q: Define the following: accuracy, precision, sensitivity, specificity, false-positive rate, false-negative rate, and F1

A:

?

Q: Can you translate the metrics above within our case study?

A:

?

Q: Why F1 important in our dataset?

A: ?

Task: Define a function that computes the main classification metrics.

```{r}
classification_metrics <- function(actual, predicted) {
  confusion <- table(
    Actual = factor(actual, levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  )

  true_negative <- as.numeric(confusion["0", "0"])
  false_positive <- as.numeric(confusion["0", "1"])
  false_negative <- as.numeric(confusion["1", "0"])
  true_positive <- as.numeric(confusion["1", "1"])

  safe_divide <- function(numerator, denominator) {
    if (denominator == 0) {
      NA_real_
    } else {
      numerator / denominator
    }
  }

  accuracy <- safe_divide(
    ?,
    ?
  )
  precision <- safe_divide(
    true_positive,
    true_positive + false_positive
  )
  sensitivity <- safe_divide(
    true_positive,
    true_positive + false_negative
  )
  specificity <- safe_divide(
    true_negative,
    true_negative + false_positive
  )
  false_positive_rate <- safe_divide(
    false_positive,
    false_positive + true_negative
  )
  false_negative_rate <- safe_divide(
    false_negative,
    false_negative + true_positive
  )

  if (is.na(precision) || is.na(sensitivity)) {
    f1 <- NA_real_
  } else if (precision + sensitivity == 0) {
    f1 <- 0
  } else {
    f1 <- 2 * precision * sensitivity / (precision + sensitivity)
  }

  tibble(
    Metric = c(
      "Accuracy",
      "Precision",
      "Sensitivity (Recall)",
      "Specificity",
      "False Positive Rate (Type I)",
      "False Negative Rate (Type II)",
      "F1"
    ),
    Value = c(
      accuracy,
      precision,
      sensitivity,
      specificity,
      false_positive_rate,
      false_negative_rate,
      f1
    )
  )
}

get_metric <- function(metrics, metric_name) {
  metrics %>%
    filter(Metric == metric_name) %>%
    pull(Value)
}

format_rate <- function(value) {
  if (length(value) == 0 || is.na(value)) {
    "undefined"
  } else {
    scales::percent(value, accuracy = 0.1)
  }
}
```

Task: Compute accuracy, precision, sensitivity, specificity, false-positive rate, false-negative rate, and F1.

```{r}
base_metrics <- ?(
  test_data$Bad_Earnings,
  test_data$base_prediction
)

knitr::kable(base_metrics, digits = 3)

no_information_rate <- max(
  mean(test_data$Bad_Earnings == 0),
  mean(test_data$Bad_Earnings == 1)
)
```

Q: How do we interpret these metrics?

A: ?

Q: Is this a good classifier?

A: ?

Task: Construct the ROC curve.

```{r}
base_roc <- pROC::?(
  response = test_data$Bad_Earnings,
  predictor = ?,
  levels = c(0, 1),
  direction = "?",
  quiet = TRUE
)

pROC::ggroc(base_roc, legacy.axes = TRUE) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(
    title = "ROC Curve: Base Model",
    x = "False Positive Rate",
    y = "Sensitivity"
  ) +
  theme_minimal()
```

Q: How do we interpret the ROC curve?

A: ?

Task: Compute the area under the ROC curve.

```{r}
base_auc <- as.numeric(pROC::?(base_roc))
base_auc
```

Q: How do we interpret the AUC?

A: ?

# Alternative Model

## Estimate

Task: Estimate a logistic regression using all predictors.

```{r}
#| warning: true

alt_formula <- reformulate(
  termlabels = feature_names,
  response = "Bad_Earnings"
)

alt_model <- ?(
  ?,
  data = train_data,
  family = binomial(link = "logit")
)
```

## Interpret

Task: Display the coefficient on GPM in the alternative model.

```{r}
summary(?)
```

Q: What differences do you notice about the GPM coefficient. Why might that be the case?

A: ?

Task: Display all model coefficients, sorted by p-value.

```{r}
alt_coefficients <- as.data.frame(
  coef(summary(alt_model))
) %>%
  tibble::rownames_to_column("Predictor") %>%
  as_tibble() %>%
  rename(
    Standard_Error = `Std. Error`,
    z_value = `z value`,
    p_value = `Pr(>|z|)`
  ) %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(
    Odds_Ratio = exp(Estimate),
    Absolute_Estimate = abs(Estimate)
  ) %>%
  arrange(p_value)

knitr::kable(
  alt_coefficients %>%
    select(
      Predictor,
      Estimate,
      Standard_Error,
      z_value,
      p_value,
      Odds_Ratio
    ),
  digits = 3
)

top_alt_terms <- alt_coefficients %>%
  filter(!is.na(p_value)) %>%
  slice_head(n = 3) %>%
  pull(Predictor)
```

Q: Which model terms show the strongest conditional associations with bad earnings, and why is that not the same as definitive predictor importance?

A: ?

## Predict

Task: Predict the probability of bad earnings for each company in the test sample.

```{r}
test_data <- test_data %>%
  mutate(
    alt_probability = ?(
      ?,
      newdata = ?,
      type = "response"
    )
  )

head(
  test_data %>%
    select(Company_ID, Bad_Earnings, alt_probability)
)
```

Task: Predict bad earnings for two hypothetical companies: one with GPM at the first quartile and one with GPM at the third quartile. Hold all other numeric predictors at their training-sample means and all categorical predictors at their most common training-sample values.

```{r}
most_common_value <- function(x) {
  unique_values <- unique(x)
  unique_values[
    which.max(
      tabulate(match(x, unique_values))
    )
  ]
}

reference_numeric_profile <- train_data %>%
  summarise(
    across(all_of(numeric_feature_names), mean)
  )

reference_categorical_profile <- train_data %>%
  summarise(
    across(
      all_of(categorical_feature_names),
      most_common_value
    )
  )

reference_profile <- bind_cols(
  reference_numeric_profile,
  reference_categorical_profile
)

alt_profiles <- bind_rows(
  reference_profile,
  reference_profile
) %>%
  mutate(
    Company = c(
      "First-Quartile Gross Margin",
      "Third-Quartile Gross Margin"
    ),
    GPM = as.numeric(gross_margin_quantiles)
  ) %>%
  mutate(
    Predicted_Probability = predict(
      alt_model,
      newdata = .,
      type = "response"
    ),
    Predicted_Class = as.integer(
      Predicted_Probability ? params$prob_threshold
    )
  ) %>%
  relocate(Company)

knitr::kable(
  alt_profiles %>%
    select(
      Company,
      GPM,
      Predicted_Probability,
      Predicted_Class
    ),
  digits = 3
)
```

## Classify

Task: Convert the alternative-model probabilities into predicted classes using the classification threshold defined earlier.

```{r}
test_data <- test_data %>%
  mutate(
    alt_prediction = as.integer(
      alt_probability ? params$prob_threshold
    )
  )

head(
  test_data %>%
    select(
      Company_ID,
      Bad_Earnings,
      alt_probability,
      alt_prediction
    )
)
```

Task: Construct a confusion matrix. Rows are actual classes and columns are predicted classes.

```{r}
alt_confusion <- table(
  Actual = ?,
  Predicted = factor(test_data$alt_prediction, levels = c(0, 1))
)

alt_confusion
```

Q: How do we interpret this confusion matrix?

A: ?

Task: Compute the classification metrics.

```{r}
alt_metrics <- ?(
  test_data$Bad_Earnings,
  test_data$alt_prediction
)

knitr::kable(alt_metrics, digits = 3)
```

Q: How do we interpret these metrics?

A: ?

Task: Compute the AUC.

```{r}

alt_roc <- pROC::?(
  response = ?,
  predictor = test_data$alt_prediction,
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

alt_auc <- as.numeric(pROC::auc(alt_roc))
alt_auc
```

# Compare the Models

Task: Create a table comparing the classification metrics and ROC AUC of the two models.

```{r}
model_comparison <- base_metrics %>%
  rename(`Base Model` = Value) %>%
  left_join(
    alt_metrics %>%
      rename(`Alternative Model` = Value),
    by = "Metric"
  ) %>%
  bind_rows(
    tibble(
      Metric = "ROC AUC",
      `Base Model` = base_auc,
      `Alternative Model` = alt_auc
    )
  )

knitr::kable(model_comparison, digits = 3)
```

Task: Plot the two ROC curves together.

```{r}
roc_models <- list(
  `Base Model` = base_roc,
  `Alternative Model` = alt_roc
)

pROC::ggroc(roc_models, legacy.axes = TRUE) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(
    title = "Test-Sample ROC Curves",
    x = "False Positive Rate",
    y = "Sensitivity",
    color = "Model"
  ) +
  theme_minimal()
```

Q: Which model appears better?

A: ?

# Threshold Analysis

The analyses below use the holdout sample to illustrate how definitions and decision rules affect measured performance. In an applied project, do not use these tables to tune a threshold and then report performance on the same holdout sample as a final result. Select thresholds with a validation sample or cross-validation, and reserve a separate test sample for one final evaluation.

## Classification-Probability Threshold

Task: Create a vector of alternative classification-probability thresholds.

```{r}
prob_thresholds <- c(0.35, 0.45, 0.55, 0.65, 0.75, 0.85)
prob_thresholds
```

Task: Reclassify the alternative model's existing probabilities at each threshold. For each threshold, compute sensitivity, specificity, false-positive rate, and false-negative rate.

```{r}
prob_threshold_results <- lapply(
  prob_thresholds,
  function(threshold) {
    predicted_class <- as.integer(
      test_data$alt_probability ? threshold
    )

    metrics <- classification_metrics(
      test_data$Bad_Earnings,
      ?
    )

    tibble(
      Probability_Threshold = threshold,
      Predicted_Bad_Rate = mean(predicted_class),
      Sensitivity = get_metric(
        metrics,
        "Sensitivity (Recall)"
      ),
      Specificity = get_metric(
        metrics,
        "Specificity"
      ),
      False_Positive_Rate = get_metric(
        metrics,
        "False Positive Rate (Type I)"
      ),
      False_Negative_Rate = get_metric(
        metrics,
        "False Negative Rate (Type II)"
      )
    )
  }
) %>%
  bind_rows()

knitr::kable(prob_threshold_results, digits = 3)
```

Task: Plot sensitivity and specificity against the probability threshold.

```{r}
prob_threshold_results %>%
  select(
    Probability_Threshold,
    Sensitivity,
    Specificity
  ) %>%
  pivot_longer(
    cols = c(Sensitivity, Specificity),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  ggplot(
    aes(
      x = Probability_Threshold,
      y = Value,
      linetype = Metric
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Classification Performance by Probability Threshold",
    x = "Probability Threshold",
    y = "Rate",
    linetype = "Metric"
  ) +
  theme_minimal()
```

Q: Why do these metrics change with `prob_threshold`?

A: ?

Q: Why is this trade-off relevant for the lender?

A: ?

## Target-Definition Threshold

The values below are quantiles used to define the observed target. They are not probability thresholds.

Task: Create a vector of alternative lower-tail quantiles for defining bad earnings.

```{r}
bad_quantiles <- c(0.01, 0.05, 0.10, 0.15, 0.20, 0.25)
bad_quantiles
```

Task: For each quantile, compute the percent of the training sample that is bad, the percent for the test sample, the sensitivity and the specificity.

```{r}

#| warning: true

bad_quantile_results <- lapply(
  bad_quantiles,
  function(bad_quantile) {
    
    # Define bad earnings using the training-sample cutoff
    earnings_cutoff <- as.numeric(
      quantile(
        train_data$Quarterly_Earnings_Growth,
        probs = ?
      )
    )

    # Recreate the target in the training and test samples
    threshold_train <- train_data %>%
      mutate(
        Target_Bad_Earnings = as.integer(
          Quarterly_Earnings_Growth <= earnings_cutoff
        )
      )

    threshold_test <- test_data %>%
      mutate(
        Target_Bad_Earnings = as.integer(
          Quarterly_Earnings_Growth ? earnings_cutoff
        )
      )

    # Refit the logistic model
    threshold_formula <- reformulate(
      termlabels = feature_names,
      response = "Target_Bad_Earnings"
    )

    threshold_model <- glm(
      threshold_formula,
      data = threshold_train,
      family = binomial(link = "logit")
    )

    # Predict the test sample
    threshold_probability <- predict(
      threshold_model,
      newdata = threshold_test,
      type = "response"
    )

    threshold_prediction <- as.integer(
      threshold_probability >= params$prob_threshold
    )

    # Compute classification metrics
    metrics <- classification_metrics(
      threshold_test$Target_Bad_Earnings,
      threshold_prediction
    )

    tibble(
      Bad_Quantile = bad_quantile,
      Earnings_Growth_Cutoff = earnings_cutoff,
      Training_Bad_Rate = mean(
        threshold_train$Target_Bad_Earnings
      ),
      Test_Bad_Rate = mean(
        threshold_test$Target_Bad_Earnings
      ),
      Sensitivity = get_metric(
        metrics,
        "Sensitivity (Recall)"
      ),
      Specificity = get_metric(
        metrics,
        "Specificity"
      )
    )
  }
) %>%
  bind_rows()

# Display the results
bad_quantile_results %>%
  mutate(
    across(
      c(
        Training_Bad_Rate,
        Test_Bad_Rate,
        Sensitivity,
        Specificity
      ),
      ~ scales::percent(.x, accuracy = 0.1)
    )
  ) %>%
  knitr::kable()
```

Q: Why do the metrics change when the target-definition quantile changes?

A:

?

Q: Why is the target definition relevant for the portfolio manager?

A:

?