---
title: "Generative Models: Gaussian and KDE"
subtitle: "Machine Learning for Finance"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Code"
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
editor: source
---

## Background

A risk analyst at a private wealth manager wants to compute the range of potential outcomes for a client's retirement portfolio.  

Q: How might the analyst use Generative models like Gaussian parametric and nonparametric kernel density?

A:

## Housekeeping

```{r}
#| label: setup

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace

required_packages <- c("tidyverse", "scales", "knitr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace,
                                               quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_packages) > 0) {
  stop("Install these packages before rendering: ", paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(scales)
library(knitr)
library(here)

source(here("Topics/Generative", "ReturnSimulation-Helpers.R"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title.position = "plot"
    )
)

set.seed(20260905)

# Simulation settings. These are fixed constants; the exercise is small enough
# that the full run completes quickly.
N_PATHS <- 1000L
SIMULATION_DAYS <- 252L
ONE_DAY_DRAWS <- 100000L
KDE_QUANTILE_DRAWS <- 20000L

model_colors <- c("Historical" = "grey35", "Gaussian" = "#0072B2", "KDE" = "#D55E00")

asset_levels <- c("Asset 1", "Asset 2", "Asset 3", "Asset 4")
```

`N_PATHS` is the number of simulated one-year paths per asset and model.
`ONE_DAY_DRAWS` is the size of the large one-day sample used to compare fitted
marginal distributions. `KDE_QUANTILE_DRAWS` controls the reference sample used
for KDE interval and tail quantiles during validation.

## Data

Task: Load the simulated returns. The returns were generated by
[`SimulatedData-ReturnDistributions.R`](SimulatedData-ReturnDistributions.R)
and stored in `data/KDE-returns.csv`. 

```{r}
#| label: load-data

asset_returns <- read_csv(here("data", "KDE-returns.csv"),
                          show_col_types = FALSE) |>
  mutate(
    date = as.Date(date),
    asset_id = factor(asset_id, levels = asset_levels)
  )

sample_check <- asset_returns |>
  summarise(
    assets = n_distinct(asset_id),
    trading_days = n_distinct(date),
    observations = n(),
    first_date = min(date),
    last_date = max(date),
    missing_returns = sum(is.na(return)),
    minimum_observations_per_asset = min(table(asset_id)),
    maximum_observations_per_asset = max(table(asset_id))
  )

kable(sample_check, caption = "Synthetic sample integrity checks")
```

The dates are synthetic weekday labels. Exchange holidays are intentionally
outside the scope of the lesson.

### Calibration and validation samples

Task: Use the first four years to fit the models and reserve the fifth year for
validation.

```{r}
#| label: split-sample

all_dates <- sort(unique(asset_returns$date))
calibration_dates <- all_dates[1:1008]
validation_dates <- all_dates[1009:1260]

calibration <- asset_returns |>
  filter(date %in% calibration_dates)

validation <- asset_returns |>
  filter(date %in% validation_dates)

split_check <- bind_rows(
  calibration |> summarise(sample = "Calibration", start = min(date), end = max(date),
                           days = n_distinct(date), observations = n()),
  validation |> summarise(sample = "Validation", start = min(date), end = max(date),
                          days = n_distinct(date), observations = n())
)

kable(split_check, caption = "Chronological calibration and validation samples")
```

Q: Why should we not evaluate the models only on the data used to fit them?

A:

## Exploratory data analysis

All distributional EDA and model estimation below use the calibration sample so
that the validation year remains genuinely unseen.

```{r}
#| label: asset-summary

asset_summary <- calibration |>
  group_by(asset_id) |>
  summarise(distribution_summary(return), .groups = "drop")

kable(asset_summary, digits = 4,
      caption = "Calibration-period summaries for the four assets")
```


Task: Create time series plots of returns for each asset. 

```{r}
#| label: time-series
#| echo: true
#| fig-height: 7

calibration |>
  ggplot(aes(date, return)) +
  geom_line(linewidth = 0.3, color = "#3B6FB6") +
  facet_wrap(~ asset_id, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Calibration-period daily returns",
    x = NULL,
    y = "Daily return"
  )
```


Task: Create histograms of returns for each asset

```{r}
#| label: histograms
#| echo: true
#| fig-height: 6

calibration |>
  ggplot(aes(return)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45,
                 fill = "grey80", color = "white") +
  geom_density(color = "grey25", linewidth = 0.8) +
  facet_wrap(~ asset_id, scales = "free", ncol = 2) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "Return distributions for the four assets",
    x = "Daily return",
    y = "Density"
  )
```


Task: Create QQ plots of the 4 assets

```{r}
#| label: qq-plots
#| echo: true
#| fig-height: 6

calibration |>
  ggplot(aes(sample = return)) +
  stat_qq(alpha = 0.45, size = 0.8) +
  stat_qq_line(color = "#D55E00", linewidth = 0.8) +
  facet_wrap(~ asset_id, scales = "free", ncol = 2) +
  labs(
    title = "Gaussian Q–Q plots",
    x = "Theoretical Gaussian quantile",
    y = "Observed return quantile"
  )
```

Q: Which features would make a Gaussian model potentially inadequate?

A:

Task: For each asset, recommend either the Gaussian model or KDE. Justify the
recommendation using the histogram, Q–Q plot, skewness, kurtosis, and tail
quantiles.

## Gaussian parametric model

### Estimation

For asset $i$, estimate the calibration-period mean and standard deviation:

$$
\widehat{\mu}_i=\frac{1}{T}\sum_{t=1}^{T}r_{it},
\qquad
\widehat{\sigma}_i=
\sqrt{\frac{1}{T-1}\sum_{t=1}^{T}(r_{it}-\widehat{\mu}_i)^2}.
$$

The fitted model assumes:

$$
r_{i,t}^{\mathrm{sim}}\sim N\left(\widehat{\mu}_i,\widehat{\sigma}_i^2\right).
$$

Task: Summarize returns from the calibration set. 

```{r}
#| label: fit-gaussian

gaussian_estimates <- calibration |>
  group_by(?) |>
  ?(
    mu = ?,
    sigma = ?,
    .groups = "drop"
  )

kable(gaussian_estimates, digits = 5, caption = "Gaussian parameter estimates")
```

Q: If the Gaussian simulation matches an asset's mean and standard deviation, does it reproduce the entire distribution?

A:

### Simulation

Q: How can we use a Gaussian parametric simulation to achieve the risk manager's goal?

A:

Task: Create a temporary subset to focus on Asset 2

```{r}
#| label: demonstrate-gaussian-draw

set.seed(4127)

gaussian_example <- gaussian_estimates |>
  ?
tibble(simulated_return = draw_gaussian(10, gaussian_example$mu,
                                        gaussian_example$sigma)) |>
  mutate(simulated_return = percent(simulated_return, accuracy = 0.01)) |>
  kable(caption = "Ten illustrative Gaussian draws for Asset 2")
```

## KDE model

### KDE Intuition

Q: Can you describe the purpose and mechanics of KDE? 

A:

Q: How simulate from the KDE? 

A:

### Estimation

For asset $i$, the KDE is:

$$
\widehat f_i(r)=\frac{1}{T h_i}\sum_{t=1}^{T}
K\left(\frac{r-r_{it}}{h_i}\right),
$$

where $K$ is the kernel and $h_i$ is the bandwidth. We use a Gaussian kernel and
the normal-reference bandwidth (`bw.nrd0`) for the main comparison.

Task: Estimate the bandwidth for each asset within the calibration sample. 

```{r}
#| label: fit-kde

kde_estimates <- calibration |>
  group_by(?) |>
  summarise(
    bw_nrd0 = ?,
    .groups = "drop"
  )

kable(kde_estimates, digits = 5, caption = "KDE bandwidth estimates")
```

### Simulation

Q: How can we use KDE-based simulation to achieve the risk manager's goal?

A:

Task: Focus on Asset 3.  Draw 10 samples via draw_kde using this asset's estimated bandwidth. 

```{r}
#| label: demonstrate-kde-draw

example_id <- "Asset 3"
example_returns <- calibration |>
  filter(asset_id == ?) |>
  pull(return)
example_bandwidth <- ?

set.seed(6503)
example_kde_draws <- ?

tibble(simulated_return = example_kde_draws) |>
  mutate(simulated_return = percent(simulated_return, accuracy = 0.01)) |>
  kable(caption = paste("Ten illustrative KDE draws for", example_id))
```

Q: How does KDE simulation differ from simply resampling historical returns?

A:

### Bandwidth and the bias–variance tradeoff


Task: Focus on Asset 3.  Construct a plot of the empirical density with bandwidth 0.5. Repeat for 1 and 2. 

```{r}
#| label: bandwidth-adjust
#| fig-height: 5

adjust_values <- c(0.5, 1, 2)

bandwidth_curves <- map_dfr(adjust_values, function(adjust_value) {
  estimate <- density(?, bw = ?, adjust = adjust_value,
                      n = 1024)
  tibble(
    return = estimate$x,
    density = estimate$y,
    bandwidth = paste0("adjust = ", adjust_value)
  )
})


ggplot() +
  geom_histogram(
    data = tibble(return = example_returns),
    aes(return, y = after_stat(density)),
    bins = 45, fill = "grey85", color = "white"
  ) +
  geom_line(
    data = bandwidth_curves,
    aes(return, density, color = bandwidth),
    linewidth = 0.9
  ) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "The KDE changes substantially with bandwidth",
    subtitle = paste("Illustration using", example_id),
    x = "Daily return",
    y = "Density",
    color = NULL
  )

```

Q: What happens when we decrease or increase the KDE bandwidth?

A:

## In-sample fit

### In-sample fitted densities

Task: For each asset, identify the features that the Gaussian model misses and assess
whether the KDE appears to capture genuine structure or local sampling noise.

```{r}
#| label: in-sample-density-data
#| fig-height: 6

density_comparison <- calibration |>
  group_by(asset_id) |>
  group_modify(~ {
    x <- .x$return
    grid <- seq(min(x) - 0.15 * diff(range(x)),
                max(x) + 0.15 * diff(range(x)), length.out = 800)
    kde_fit <- density(?, bw = ?, from = ?, to = ?, n = 800)
    bind_rows(
      tibble(return = grid, density = dnorm(grid, mean(x), sd(x)), model = "Gaussian"),
      tibble(return = kde_fit$x, density = kde_fit$y, model = "KDE")
    )
  }) |>
  ungroup()


ggplot() +
  geom_histogram(
    data = calibration,
    aes(return, y = after_stat(density)),
    bins = 45, fill = "grey85", color = "white"
  ) +
  geom_line(
    data = density_comparison,
    aes(return, density, color = model),
    linewidth = 0.9
  ) +
  facet_wrap(~ asset_id, scales = "free", ncol = 2) +
  scale_color_manual(values = model_colors[c("Gaussian", "KDE")]) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "Gaussian and KDE fits to calibration returns",
    x = "Daily return",
    y = "Density",
    color = NULL
  )
```

Q: Does the closer-looking KDE fit establish that KDE is the better model?

A:

### Compare simulated one-day returns

Task: For each asset, draw a large one-day simulation sample from each fitted model and
compare its distribution with the calibration returns.

```{r}
#| label: one-day-simulation

set.seed(1401)

one_day_comparison <- map_dfr(asset_levels, function(id) {
  historical <- calibration |> filter(asset_id == id) |> pull(return)
  fitted <- fit_marginal_models(?)

  bind_rows(
    distribution_summary(historical) |> mutate(model = "Historical"),
    distribution_summary(draw_gaussian(ONE_DAY_DRAWS, fitted$mu, fitted$sigma)) |>
      mutate(model = "Gaussian"),
    distribution_summary(draw_kde(ONE_DAY_DRAWS, historical, fitted$bw_nrd0)) |>
      mutate(model = "KDE")
  ) |>
    mutate(asset_id = id, .before = 1)
})

one_day_comparison |>
  mutate(across(where(is.numeric), ~ round(.x, 5))) |>
  kable(caption = "Historical and simulated one-day return distributions")
```

Q: Why can KDE-simulated volatility be slightly higher than historical volatility?

A:

Q: Why can the Gaussian model match mean and volatility while producing different tail-risk estimates?

A:

## Out-of-sample

### Simulate one-year paths

Each path contains 252 simulated daily arithmetic returns. Starting with wealth
of 1, path $p$ for asset $i$ evolves as:

$$
W_{i,p,\tau}=\prod_{s=1}^{\tau}(1+r_{i,p,s}^{\mathrm{sim}}).
$$

Q: What does one simulated path represent?

A:

Task: Use the custom "simulate_paths" function across our assets for the pre-defined number of paths and simulations.  

```{r}
#| label: simulate-all-paths

path_summaries <- list()
path_results <- list()

for (i in seq_along(asset_levels)) {
  id <- asset_levels[i]
  x <- calibration |> filter(asset_id == id) |> pull(return)
  bandwidth <- bw.nrd0(x)

  for (model_name in c("Gaussian", "KDE")) {
    set.seed(500000 + i * 10 + match(model_name, c("Gaussian", "KDE")))

    result <- simulate_paths(
?
      keep_paths = TRUE
    )

    path_summaries[[paste(id, model_name, sep = "__")]] <-
      result$summary |> mutate(asset_id = id, .before = 1)
    path_results[[paste(id, model_name, sep = "__")]] <- result
  }
}

path_summary <- bind_rows(path_summaries) |>
  mutate(asset_id = factor(asset_id, levels = asset_levels))

kable(
  path_summary,
  digits = 4,
  caption = "One-year simulation summaries by asset and model"
)
```


Q: Interpret the median drawdown for these assets. 

A:

Task: Create a side by side plots of the paths for wealth distributions of the Gaussian and KDE estimators.  

```{r}
#| label: path-band-data
#| fig-height: 9

path_band_data <- imap_dfr(path_results, function(result, key) {
  pieces <- strsplit(key, "__", fixed = TRUE)[[1]]
  result$bands |>
    mutate(asset_id = pieces[1], model = pieces[2])
}) |>
  mutate(asset_id = factor(asset_id, levels = asset_levels))

ggplot(path_band_data, aes(day, q50, color = model, fill = model)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.10, color = NA) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.16, color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.25, color = NA) +
  geom_line(linewidth = 0.8) +
  facet_grid(asset_id ~ model, scales = "free_y") +
  scale_color_manual(values = model_colors[c("Gaussian", "KDE")]) +
  scale_fill_manual(values = model_colors[c("Gaussian", "KDE")]) +
  labs(
    title = "One-year simulated wealth distributions",
    subtitle = "Bands show the central 50%, 90%, and 95% of simulated wealth",
    x = "Trading day",
    y = "Wealth from an initial value of 1",
    color = NULL,
    fill = NULL
  ) +
  theme(legend.position = "none")
```

Q: How should we interpret a central 90% simulation band?

A:

Q: Why can terminal wealth be skewed even when daily returns are Gaussian?

A:

### Compare planning outcomes

Task: 

Compare the Gaussian and KDE models on the planning summaries that matter to
the risk manager:

- downside risk, measured by Value at Risk and Expected Shortfall;
- the probability that terminal wealth falls below its starting value; and
- the typical maximum drawdown.

Starting with wealth equal to 1, define the one-year loss on path $p$ fpr asset $i$ as

$$
L_{i,p}=1-W_{i,p,252}.
$$

A positive value of $L_{i,p}$ represents a loss relative to starting wealth.

The 95% Value at Risk is the 95th percentile of the simulated loss
distribution:

$$
\operatorname{VaR}_{0.95}(L_i)
=
q_{0.95}(L_i).
$$

It is the loss threshold exceeded by approximately the worst 5% of simulated
paths. Expected Shortfall is the average loss among those worst 5% of paths:

$$
\operatorname{ES}_{0.95}(L_i)
=
E\left[L_i\mid L_i\geq \operatorname{VaR}_{0.95}(L_i)\right].
$$

Expected Shortfall is therefore sensitive not only to how often severe losses
occur, but also to how severe those losses are once they occur.

```{r}
#| label: model-risk-comparison
#| fig-height: 4.5

model_risk_comparison <- path_summary |>
  select(
    asset_id,
    model,
    median_terminal_wealth,
    probability_below_one,
    loss_var_95,
    loss_es_95,
    median_max_drawdown
  ) |>
  mutate(model = factor(model, levels = c("Gaussian", "KDE"))) |>
  arrange(asset_id, model)

risk_differences <- model_risk_comparison |>
  select(asset_id, model, loss_var_95, loss_es_95) |>
  pivot_wider(
    names_from = model,
    values_from = c(loss_var_95, loss_es_95)
  ) |>
  transmute(
    asset_id,
    var_kde_minus_gaussian =
      loss_var_95_KDE - loss_var_95_Gaussian,
    es_kde_minus_gaussian =
      loss_es_95_KDE - loss_es_95_Gaussian
  )

model_risk_comparison |>
  mutate(
    probability_below_one =
      percent(probability_below_one, accuracy = 0.1),
    loss_var_95 =
      percent(loss_var_95, accuracy = 0.1),
    loss_es_95 =
      percent(loss_es_95, accuracy = 0.1),
    median_max_drawdown =
      percent(median_max_drawdown, accuracy = 0.1)
  ) |>
  kable(
    digits = 4,
    caption = "Planning outcomes by asset and model"
  )

risk_differences |>
  mutate(
    var_kde_minus_gaussian =
      percent(var_kde_minus_gaussian, accuracy = 0.1),
    es_kde_minus_gaussian =
      percent(es_kde_minus_gaussian, accuracy = 0.1)
  ) |>
  kable(
    caption =
      "KDE minus Gaussian downside risk; positive values indicate more severe KDE risk"
  )

downside_plot_data <- path_summary |>
  select(asset_id, model, loss_var_95, loss_es_95) |>
  pivot_longer(
    cols = c(loss_var_95, loss_es_95),
    names_to = "risk_measure",
    values_to = "loss"
  ) |>
  mutate(
    risk_measure = recode(
      risk_measure,
      loss_var_95 = "Value at Risk",
      loss_es_95 = "Expected Shortfall"
    ),
    risk_measure = factor(
      risk_measure,
      levels = c("Value at Risk", "Expected Shortfall")
    )
  )

ggplot(
  downside_plot_data,
  aes(asset_id, loss, fill = model)
) +
  geom_col(
    position = position_dodge(width = 0.7),
    width = 0.6
  ) +
  facet_wrap(~ risk_measure, nrow = 1) +
  scale_fill_manual(
    values = model_colors[c("Gaussian", "KDE")]
  ) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "One-year downside risk by asset and model",
    x = NULL,
    y = "Loss relative to starting wealth",
    fill = NULL
  )
```

Q: 

Interpret the planning results. Which assets show the most important
differences between the Gaussian and KDE models?

A:

### Validation

The models were calibrated using the first four years of returns. We now
evaluate them using the subsequent 252 daily returns that were not used during
estimation.

A useful validation strategy should examine more than one statistic.

$\underline{\text{Log Score}}$

The goal is to approximate the unknown density that generates future returns.
Call this unknown density $f_0$. Because $f_0$ is not observable, each fitted
model provides a proxy $\widehat f_m$ for that density.

Suppose $r_t$ is one realized return from the validation period. For a
continuous variable, the probability of observing exactly $r_t$ is zero.
Instead, $\widehat f_m(r_t)$ measures the local concentration of probability
around that realized return.

For model $m$, define the score for observation $r_t$ as

$$
\ell_{m,t}
=
\log\left[\widehat f_m(r_t)\right].
$$

The average validation log score is

$$
\operatorname{LogScore}_m
=
\frac{1}{n}
\sum_{t=1}^{n}
\log\left[\widehat f_m(r_t)\right].
$$

The validation average is an empirical estimate of

$$
E_{R\sim f_0}
\left[
\log\left(\widehat f_m(R)\right)
\right],
$$

the model's expected log density when evaluated at a future return generated
by the unknown process.

Consider a realized validation return of $r_t=-15\%$. Suppose the Gaussian
model assigns essentially no density to returns near $-15\%$:

$$
\widehat f_{\text{Gaussian}}(-0.15)=10^{-12}.
$$

Its score for this observation is therefore

$$
\log(10^{-12})\approx -27.63.
$$

Now suppose the calibration sample contains an extreme loss near $-15\%$.
The KDE places a kernel around that observation and assigns greater local
density to returns near $-15\%$:

$$
\widehat f_{\text{KDE}}(-0.15)=10^{-3}.
$$

The KDE receives the score

$$
\log(10^{-3})\approx -6.91.
$$

Thus, KDE receives a substantially better score for this realized return:

$$
-6.91-(-27.63)=20.72.
$$

The KDE is rewarded because it assigned more density to the region where the
return actually occurred. The Gaussian model is penalized because its fitted
density regarded that outcome as extremely unlikely. The numerical density
values above are illustrative; a density is not itself a probability and need
not lie between zero and one.

The log score repeats this comparison for every validation return. A model can
outperform another model because it assigns more density to important
observations, particularly observations that the competing model regards as
extremely unlikely. However, if KDE places too much density around unusual
calibration observations that do not recur, it may perform worse on the
remaining validation returns.

Higher average log scores are better. The logarithm strongly penalizes a model
when it assigns extremely low density to a realized return, so the score
rewards models that allocate density effectively across the full validation
sample, including its tails.


$\underline{Coverage}$

For a nominal predictive interval
$\left[\widehat q_{\alpha/2},\widehat q_{1-\alpha/2}\right]$, empirical
coverage is

$$
\operatorname{Coverage}_{1-\alpha}
=
\frac{1}{n}
\sum_{t=1}^{n}
\mathbf{1}
\left\{
\widehat q_{\alpha/2}
\leq r_t
\leq
\widehat q_{1-\alpha/2}
\right\}.
$$

For example, a well-calibrated 90% interval should contain approximately 90% of
the validation returns. We examine several interval levels because a model can
have reasonable central coverage while still misrepresenting the tails.


$ \underline{Tail Frequency}$

The realized 5% tail frequency is

$$
\operatorname{TailFrequency}_{0.05}
=
\frac{1}{n}
\sum_{t=1}^{n}
\mathbf{1}
\left\{
r_t\leq \widehat q_{0.05}
\right\}.
$$

This should be compared with 0.05:

- a value above 0.05 suggests that the model underestimates the frequency of
  sufficiently bad returns;
- a value below 0.05 suggests that the model predicts too many bad returns.

Because the validation sample contains only 252 observations, these measures
are subject to sampling variation. The validation evidence should therefore be
interpreted as suggestive rather than definitive.

Task: Use the custom validation_metrics function on the validation sample for each asset. Produce a table as well as bar chart of the log score for each asset under the Gaussian and KDE approaches. 

```{r}
#| label: validation-metrics
#| fig-height: 4.5

validation_results <- map_dfr(seq_along(asset_levels), function(i) {
  id <- asset_levels[i]

  train_x <- calibration |>
    filter(asset_id == id) |>
    pull(return)

  validation_x <- validation |>
    filter(asset_id == id) |>
    pull(return)

  set.seed(800000 + i)

  validation_metrics(
?
  ) |>
    mutate(asset_id = id, .before = 1)
}) |>
  mutate(
    asset_id = factor(asset_id, levels = asset_levels),
    model = factor(model, levels = c("Gaussian", "KDE"))
  )

validation_results |>
  mutate(
    coverage_90 = percent(coverage_90, accuracy = 0.1),
    coverage_95 = percent(coverage_95, accuracy = 0.1),
    coverage_98 = percent(coverage_98, accuracy = 0.1),
    predicted_q05 = percent(predicted_q05, accuracy = 0.1),
    realized_below_q05 =
      percent(realized_below_q05, accuracy = 0.1)
  ) |>
  kable(
    digits = 4,
    caption = "Out-of-sample validation by asset and model"
  )

validation_results |>
  ggplot(aes(asset_id, log_score, fill = model)) +
  geom_col(
    position = position_dodge(width = 0.7),
    width = 0.6
  ) +
  scale_fill_manual(
    values = model_colors[c("Gaussian", "KDE")]
  ) +
  labs(
    title = "Out-of-sample log score by asset and model",
    subtitle = "Higher values indicate better predictive density",
    x = NULL,
    y = "Mean validation log score",
    fill = NULL
  )
```

Q: 

Interpret the validation results. What do the log scores, coverage measures,
and tail frequencies suggest?

A:

Q: 

Why might KDE fit better in sample but produce a worse validation log score?

A:

