---
title: "Regression and the Cross-Section of Industry Returns"
subtitle: "A Fama-MacBeth exercise with 48 industry portfolios"
author: Mike Aguilar
format:
  html:
    toc: true
    toc-depth: 5
    toc-location: left
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: true
---

# Background

Q: Suppose you are an equity analyst. Your boss asking you to "run a factor analysis on these stocks".

- Explain the concept of a "factor" in the equity markets.
- Define the meaning of Mkt-RF, SMB, HML.
- How use for investment analysis?

A:

- A **factor** is a common source of variation in asset returns. Factor models seek to summarize the movements of many assets using a relatively small number of systematic return drivers.

- **Mkt−RF** is the return on the broad equity market in excess of the risk-free rate. It measures the market risk premium realized during a particular period.

- **SMB**, or “Small Minus Big,” is the return on a portfolio of small-cap stocks minus the return on a portfolio of large-cap stocks. It captures the performance of small firms relative to large firms.

- **HML**, or “High Minus Low,” is the return on a portfolio of high book-to-market—or value—stocks minus the return on a portfolio of low book-to-market—or growth—stocks. It captures the performance of value stocks relative to growth stocks.

- Different stocks and industries have different **factor exposures**: they respond differently when the market, size, and value factors rise or fall. A factor model can estimate these exposures and use them to explain and compare the sources of portfolio risk and return.

- Factor exposures can also support risk management. For example, an investor who expects value stocks to face unusually adverse conditions may be cautious about holding a portfolio with a large positive exposure to HML.

- Finally, the analyst can ask whether assets with greater exposure to a factor tend to earn higher average returns. This connects factor exposures to the cross section of expected returns and motivates the Fama–MacBeth procedure.

## Architecture and notation

We study the question using:

- the monthly **Ken French 48 Industry Portfolios** as test assets; and
- the monthly **Fama-French three factors**: market excess return, size, and value.

The exercise has two main steps.

1.  **Time-series regressions:** estimate each industry's factor exposures.
2.  **Cross-sectional regressions:** estimate the return associated with each unit of factor exposure.

This is an **asset-pricing exercise**, not a trading backtest. We use the full sample to estimate first-pass betas so that the mechanics are transparent. A genuinely investable exercise would estimate betas using only information available at each portfolio-formation date.

The first-pass regression for industry $i$ is

$$
R_{i,t}-R_{f,t}
=
\alpha_i
+\beta_{i,MKT}(R_{M,t}-R_{f,t})
+\beta_{i,SMB}SMB_t
+\beta_{i,HML}HML_t
+\varepsilon_{i,t}.
$$

The first pass uses variation **through time** to estimate one set of exposures for each industry.

The second-pass regression for month $t$ is

$$
R_{i,t}-R_{f,t}
=
\lambda_{0,t}
+\lambda_{MKT,t}\widehat\beta_{i,MKT}
+\lambda_{SMB,t}\widehat\beta_{i,SMB}
+\lambda_{HML,t}\widehat\beta_{i,HML}
+u_{i,t}.
$$

The second pass uses variation **across industries**. Repeating it each month produces a time series of estimated prices of risk, $\widehat\lambda_{k,t}$.

| Quantity | Varies across | Interpretation |
|------------------------|------------------------|------------------------|
| Factor return, $F_{k,t}$ | Months | The realized factor payoff in month $t$ |
| Factor beta, $\beta_{i,k}$ | Industries | Industry $i$'s sensitivity to factor $k$ |
| Factor price, $\lambda_{k,t}$ | Months | Cross-sectional return associated with one unit of beta exposure in month $t$ |

Q: What is the meaning of $\beta$?

A: $\beta_{i,k}$ represents the sensitivity of industry $i$'s returns to factor $k$. It measures how much industry $i$'s returns are expected to change for a one-unit change in the return of factor $k$, holding other factors constant; aka Risk Exposure.

Q: What is the meaning of $\lambda$?

A: The $\lambda$ represents the cross-sectional return associated with one unit of beta exposure in month $t$. It is the estimated price of risk for each factor in a given month; aka Risk Premium

# Housekeeping

Task: Reset the R session, load the required packages, define the analysis settings, and create small formatting and regression helpers. Let's define "Banks" as the "focus" industry; i.e. the industry that we will use in this exercise to zoom in on and explore in more depth.

```{r}
#| label: setup

cat("\014")
rm(list = ls())

required_packages <- c(
  "tidyverse", "lubridate", "broom", "sandwich", "lmtest",
  "gt", "scales", "zoo", "ggrepel"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(broom)
  library(sandwich)
  library(lmtest)
  library(gt)
  library(scales)
  library(zoo)
  library(ggrepel)
})

theme_set(theme_minimal(base_size = 12))

settings <- list(
  start_date = as.Date("1970-01-01"),
  end_date = Sys.Date(),
  focus_industry = "Banks",
  nw_lag = 6,
  refresh_data = FALSE,
  data_dir = "data"
)

dir.create(settings$data_dir, showWarnings = FALSE, recursive = TRUE)

pct <- function(x, accuracy = 0.01) {
  scales::percent(x, accuracy = accuracy)
}

ppts <- function(x, digits = 2) {
  paste0(format(round(100 * x, digits), nsmall = digits), " ppts")
}

gt_compact <- function(data, title = NULL, subtitle = NULL) {
  out <- gt(data)
  if (!is.null(title)) {
    out <- tab_header(out, title = title, subtitle = subtitle)
  }
  out |>
    sub_missing(missing_text = "—") |>
    cols_align(align = "center") |>
    opt_row_striping()
}

robust_tidy <- function(model, vcov_matrix = sandwich::vcovHC(model, type = "HC1")) {
  result <- broom::tidy(lmtest::coeftest(model, vcov. = vcov_matrix))
  critical_value <- qt(0.975, df = df.residual(model))
  result |>
    mutate(
      conf.low = estimate - critical_value * std.error,
      conf.high = estimate + critical_value * std.error
    )
}

calculate_vif <- function(data, regressors) {
  map_dfr(regressors, function(target) {
    other_regressors <- setdiff(regressors, target)
    auxiliary_model <- lm(
      reformulate(other_regressors, response = target),
      data = data
    )
    auxiliary_r_squared <- summary(auxiliary_model)$r.squared

    tibble(
      variable = target,
      auxiliary_r_squared = auxiliary_r_squared,
      vif = 1 / (1 - auxiliary_r_squared)
    )
  })
}
```

# Data

## Download and parse

The data come from the [Kenneth R. French Data Library](https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html). The source files report returns in percentage points; this exercise converts them to decimal returns.

Task: Define helpers that download and extract the first monthly data block from a Ken French ZIP file.

```{r}
#| label: data-helpers

download_if_needed <- function(url, destination, refresh = FALSE) {
  if (refresh || !file.exists(destination)) {
    download.file(url, destination, mode = "wb", quiet = TRUE)
  }
  invisible(destination)
}

read_first_monthly_block <- function(zip_path) {
  member <- unzip(zip_path, list = TRUE)$Name[1]
  connection <- unz(zip_path, member, open = "rt")
  on.exit(close(connection), add = TRUE)
  lines <- readLines(connection, warn = FALSE)

  first_data_row <- which(str_detect(lines, "^\\s*\\d{6},"))[1]
  header_candidates <- which(
    seq_along(lines) < first_data_row &
      str_detect(lines, "^\\s*,")
  )
  header_row <- max(header_candidates)

  candidate_rows <- seq(first_data_row, length(lines))
  first_nonmonthly <- which(
    !str_detect(lines[candidate_rows], "^\\s*\\d{6},")
  )[1]

  last_data_row <- if (is.na(first_nonmonthly)) {
    length(lines)
  } else {
    candidate_rows[first_nonmonthly] - 1
  }

  csv_text <- paste(
    c(lines[header_row], lines[first_data_row:last_data_row]),
    collapse = "\n"
  )

  readr::read_csv(
    I(csv_text),
    show_col_types = FALSE,
    trim_ws = TRUE,
    name_repair = "minimal"
  )
}
```

Task: Download the monthly Fama-French three-factor and 48-industry files, then extract their monthly value-weighted returns.

```{r}
#| label: download-data

factor_url <- paste0(
  "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/",
  "F-F_Research_Data_Factors_CSV.zip"
)

industry_url <- paste0(
  "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/",
  "48_Industry_Portfolios_CSV.zip"
)

factor_zip <- file.path(settings$data_dir, "F-F_Research_Data_Factors_CSV.zip")
industry_zip <- file.path(settings$data_dir, "48_Industry_Portfolios_CSV.zip")

download_if_needed(factor_url, factor_zip, settings$refresh_data)
download_if_needed(industry_url, industry_zip, settings$refresh_data)

factor_raw <- read_first_monthly_block(factor_zip)
industry_raw <- read_first_monthly_block(industry_zip)
```

Task: Clean the two data sets, convert percentage-point returns to decimals, and join them into an industry-month panel.

```{r}
#| label: clean-data

names(factor_raw) <- str_trim(names(factor_raw))
names(industry_raw) <- str_trim(names(industry_raw))

factors <- factor_raw |>
  rename(
    date_code = 1,
    MKT_RF = `Mkt-RF`
  ) |>
  mutate(
    date = as.Date(paste0(date_code, "01"), format = "%Y%m%d"),
    across(c(MKT_RF, SMB, HML, RF), ~ as.numeric(.x) / 100)
  ) |>
  select(date, MKT_RF, SMB, HML, RF) |>
  filter(between(date, settings$start_date, settings$end_date))

industries <- industry_raw |>
  rename(date_code = 1) |>
  mutate(date = as.Date(paste0(date_code, "01"), format = "%Y%m%d")) |>
  select(-date_code) |>
  pivot_longer(
    cols = -date,
    names_to = "industry",
    values_to = "industry_return"
  ) |>
  mutate(
    industry = str_trim(industry),
    industry_return = as.numeric(industry_return),
    industry_return = na_if(industry_return, -99.99),
    industry_return = na_if(industry_return, -999),
    industry_return = industry_return / 100
  ) |>
  filter(between(date, settings$start_date, settings$end_date))

panel <- industries |>
  inner_join(factors, by = "date") |>
  mutate(excess_return = industry_return - RF) |>
  drop_na(excess_return, MKT_RF, SMB, HML)

data_overview <- tibble(
  Item = c("First month", "Last month", "Months", "Industries", "Panel observations"),
  Value = c(
    format(min(panel$date), "%B %Y"),
    format(max(panel$date), "%B %Y"),
    n_distinct(panel$date),
    n_distinct(panel$industry),
    nrow(panel)
  )
)

data_overview |>
  gt_compact(title = "Analysis sample")
```

## Validate

Task: Check for missing observations, duplicate industry-months, and an unbalanced panel.

```{r}
#| label: validate-data

expected_months <- n_distinct(panel$date)

validation <- bind_rows(
  tibble(
    check = "No missing model variables",
    passed = !anyNA(panel |> select(excess_return, MKT_RF, SMB, HML)),
    detail = paste(sum(!complete.cases(panel |> select(excess_return, MKT_RF, SMB, HML))), "incomplete rows")
  ),
  tibble(
    check = "One observation per industry-month",
    passed = !any(duplicated(panel[c("industry", "date")])),
    detail = paste(sum(duplicated(panel[c("industry", "date")])), "duplicates")
  ),
  panel |>
    dplyr::count(industry, name = "months") |>
    summarise(
      check = "Balanced industry panel",
      passed = all(months == expected_months),
      detail = paste(min(months), "to", max(months), "months per industry")
    )
)

if (!all(validation$passed)) {
  stop("At least one data-validation check failed.")
}

validation_table <- validation |>
  mutate(Status = if_else(passed, "PASS", "FAIL")) |>
  select(Check = check, Status, Detail = detail) |>
  gt_compact(title = "Data validation")

validation_table
```

Q: Why is a balanced panel especially useful when we compare the historical-average cross-sectional regression with the average of the monthly cross-sectional regressions?

A:

With the same industries and the same fixed beta regressors every month, OLS is linear in the dependent variable. Consequently, the coefficients from a regression of historical average returns on betas equal the time-series averages of the monthly cross-sectional coefficients, apart from numerical rounding. Missing industries would change the cross-sectional design matrix across months and break this exact equivalence.

# Exploratory data analysis

## Return distributions

Task: Summarize the monthly factors and industry excess returns.

```{r}
#| label: summary-statistics

factor_summary <- factors |>
  select(MKT_RF, SMB, HML, RF) |>
  pivot_longer(everything(), names_to = "series", values_to = "return") |>
  group_by(series) |>
  summarise(
    observations = n(),
    mean = mean(return),
    volatility = sd(return),
    minimum = min(return),
    maximum = max(return),
    .groups = "drop"
  )

industry_summary <- panel |>
  group_by(industry) |>
  summarise(
    mean_excess_return = mean(excess_return),
    volatility = sd(excess_return),
    .groups = "drop"
  )

factor_summary_table <- factor_summary |>
  mutate(across(c(mean, volatility, minimum, maximum), ~ pct(.x))) |>
  rename(
    Series = series,
    Observations = observations,
    Mean = mean,
    Volatility = volatility,
    Minimum = minimum,
    Maximum = maximum
  ) |>
  gt_compact(title = "Monthly factor summary")

factor_summary_table
```

## Correlations among first-pass regressors

Task: Compute and visualize correlations among the three factor returns used as first-pass regressors.

```{r}
#| label: factor-correlations

factor_correlations <- factors |>
  select(MKT_RF, SMB, HML) |>
  cor(use = "complete.obs")

factor_correlation_table <- factor_correlations |>
  as.data.frame() |>
  rownames_to_column("Factor") |>
  gt_compact(title = "Correlations among factor returns") |>
  fmt_number(columns = where(is.numeric), decimals = 2)

factor_correlation_plot <- factor_correlations |>
  as.data.frame() |>
  rownames_to_column("factor_x") |>
  pivot_longer(-factor_x, names_to = "factor_y", values_to = "correlation") |>
  ggplot(aes(factor_x, factor_y, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = number(correlation, accuracy = 0.01)), size = 4) +
  scale_fill_gradient2(
    low = "#b2182b", mid = "white", high = "#2166ac",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_equal()

factor_correlation_table

factor_correlation_plot
```

Q: Interpret as it relates to our case study.

A:

The factors are relatively uncorrelated. This might help us isolate and understand their "partial" effects on returns.

## Variance inflation factors

Task: Compute a VIF for each first-pass regressor. For each factor, regress it on the other two factors and use $VIF_k=1/(1-R_k^2)$.

```{r}
#| label: factor-vif

factor_vif <- calculate_vif(
  factors,
  regressors = c("MKT_RF", "SMB", "HML")
)

factor_vif_table <- factor_vif |>
  mutate(
    interpretation = case_when(
      vif > 10 ~ "Severe by the conventional rule of thumb",
      vif > 5 ~ "Potentially concerning",
      TRUE ~ "Limited inflation"
    )
  ) |>
  rename(
    Factor = variable,
    `Auxiliary R-squared` = auxiliary_r_squared,
    VIF = vif,
    Interpretation = interpretation
  ) |>
  gt_compact(title = "First-pass variance inflation factors") |>
  fmt_number(columns = c(`Auxiliary R-squared`, VIF), decimals = 2)

factor_vif_table
```

Q: Any concerns?

A:

No. Each of the VIFs are below the 2.5 rule of thumb, suggesting there is little evidence of multi-collinearity. Of course, this is just a diagnostic and not a set rule.

# Step 1: Estimate factor exposures

## Fit 48 time-series regressions

Task: Estimate one Fama-French three-factor time-series regression for each industry. Store each fitted model and compute heteroscedasticity-robust coefficient estimates.

```{r}
#| label: first-pass-models

industry_models <- panel |>
  select(industry, date, excess_return, MKT_RF, SMB, HML) |>
  group_nest(industry) |>
  mutate(
    model = purrr::map(
      data,
      ~ lm(excess_return ~ MKT_RF + SMB + HML, data = .x)
    ),
    coefficients = purrr::map(model, robust_tidy),
    fit = purrr::map(model, broom::glance)
  )

first_pass_coefficients <- industry_models |>
  select(industry, coefficients) |>
  unnest(coefficients)

first_pass_fit <- industry_models |>
  select(industry, fit) |>
  unnest(fit)
```

## Interpret one industry regression

Task: Display conventional and heteroscedasticity-robust inference for the focus industry.

```{r}
#| label: focus-industry-inference

if (!settings$focus_industry %in% industry_models$industry) {
  settings$focus_industry <- sort(industry_models$industry)[1]
}

focus_model <- industry_models |>
  filter(industry == settings$focus_industry) |>
  pull(model) |>
  pluck(1)

focus_conventional <- broom::tidy(focus_model) |>
  select(term, conventional_se = std.error, conventional_p = p.value)

focus_robust <- robust_tidy(focus_model) |>
  select(term, estimate, robust_se = std.error, robust_p = p.value, conf.low, conf.high)

focus_inference <- focus_robust |>
  left_join(focus_conventional, by = "term") |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Alpha",
      MKT_RF = "Market beta",
      SMB = "SMB beta",
      HML = "HML beta"
    )
  )

focus_inference_table <- focus_inference |>
  mutate(
    across(
      c(estimate, conventional_se, robust_se, conf.low, conf.high),
      ~ if_else(term == "Alpha", 100 * .x, .x)
    )
  ) |>
  rename(
    Term = term,
    Estimate = estimate,
    `Conventional SE` = conventional_se,
    `Conventional p-value` = conventional_p,
    `Robust SE` = robust_se,
    `Robust p-value` = robust_p,
    `Robust CI low` = conf.low,
    `Robust CI high` = conf.high
  ) |>
  gt_compact(
    title = paste(settings$focus_industry, "first-pass regression"),
    subtitle = "Alpha is shown in monthly percentage points; betas are unitless"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

focus_inference_table
```

Q: What is meant by the "robust" inference metrics?

A:

- Conventional OLS inference assumes **homoskedasticity**:

  $$
  \operatorname{Var}(\varepsilon_t \mid X) = \sigma^2
  $$

  That is, the variance of the regression errors is assumed to be constant across observations.

- Financial-market volatility can change over time, so this assumption may be unrealistic.

- If the errors are heteroskedastic, the OLS coefficient estimates remain valid, but the conventional standard errors, $p$-values, and confidence intervals may be misleading.

- HC1 robust inference allows the error variance to differ across observations:

  $$
  \operatorname{Var}(\varepsilon_t \mid X) = \sigma_t^2
  $$

- The HC1 robust covariance matrix is estimated as

  $$
  \widehat{\operatorname{Var}}_{\mathrm{HC1}}
  \left(\hat{\boldsymbol{\beta}}\right)
  =
  \frac{n}{n-k}
  (X'X)^{-1}
  X'
  \operatorname{diag}
  \left(
  \hat{\varepsilon}_1^2,\ldots,\hat{\varepsilon}_n^2
  \right)
  X
  (X'X)^{-1},
  $$

  where $n$ is the number of observations, $k$ is the number of estimated coefficients, and $\hat{\varepsilon}_t$ is the OLS residual for observation $t$.

- Each robust standard error is the square root of the corresponding diagonal element of this covariance matrix:

  $$
  SE_{\mathrm{robust}}(\hat{\beta}_j)
  =
  \sqrt{
  \left[
  \widehat{\operatorname{Var}}_{\mathrm{HC1}}
  \left(\hat{\boldsymbol{\beta}}\right)
  \right]_{jj}
  }.
  $$

- The robust test statistic and 95% confidence interval are then calculated as

  $$
  t_j
  =
  \frac{\hat{\beta}_j}
  {SE_{\mathrm{robust}}(\hat{\beta}_j)}
  $$

  and

  $$
  CI_{95\%}
  =
  \hat{\beta}_j
  \pm
  t_{0.975,n-k}
  SE_{\mathrm{robust}}(\hat{\beta}_j).
  $$

- The code implements these calculations in the following sequence:

  - `lm()` estimates the OLS coefficients and residuals.
  - `sandwich::vcovHC(model, type = "HC1")` estimates the HC1 robust covariance matrix.
  - `lmtest::coeftest()` uses that covariance matrix to calculate the robust standard errors, test statistics, and $p$-values.
  - `robust_tidy()` calculates and adds the robust 95% confidence intervals.
  - `left_join()` places the conventional and robust inference metrics beside one another for comparison.

- "Robust" does **not** mean that the coefficient estimates change. The OLS coefficient estimates remain the same; only the standard errors, $p$-values, and confidence intervals change.

- HC1 inference is robust to heteroskedasticity, but not to serial correlation. If the regression errors are correlated over time, an approach such as Newey--West inference would be needed.

Task: Report the focus regression's sample size, $R^2$, adjusted $R^2$, residual standard error, and overall F-statistic.

```{r}
#| label: focus-industry-fit

focus_fit <- broom::glance(focus_model) |>
  transmute(
    Observations = nobs,
    `R-squared` = r.squared,
    `Adjusted R-squared` = adj.r.squared,
    `Residual standard error` = sigma,
    `F-statistic` = statistic,
    `F-test p-value` = p.value
  )

focus_fit_table <- focus_fit |>
  gt_compact(title = paste(settings$focus_industry, "model fit")) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

focus_fit_table
```

Q: Comment on the model's goodness of fit.

A:

- $R^2$ is the fraction of the focus industry's sample variation in monthly excess returns explained by the fitted three-factor model.
- The adjusted $R^{2}$ is smaller. Note that adding a regressor cannot reduce ordinary $R^2$, even if that regressor contributes little. Adjusted $R^2$ imposes a degrees-of-freedom penalty, so it is weakly smaller and can decline when an unhelpful regressor is added.
- The F test examines whether all of the factors are jointly equal to 0. The small p value suggests we fail to reject, implying that at least one of the factors are statistically different than zero.
- Overall, the model fits rather well.

Q: Interpret each coefficient in the focus-industry regression.

A:

- **Alpha:** the industry's average monthly excess return left unexplained by its three factor exposures. Under the model, the null hypothesis is $H_0:\alpha_i=0$.
- **Market beta:** holding SMB and HML fixed, a 1 percentage-point increase in the market excess return is associated with an estimated $\widehat\beta_{i,MKT}$ percentage-point change in the industry's excess return.
- **SMB beta:** holding the other factors fixed, a positive coefficient indicates a small-stock tilt and a negative coefficient indicates a large-stock tilt.
  - The negative SMB coefficient does not mean that banks themselves are necessarily large firms. It means that bank-stock returns tend to move in the opposite direction of the SMB factor; that is, they tend to perform relatively better when large stocks outperform small stocks.
- **HML beta:** holding the other factors fixed, a positive coefficient indicates a value tilt and a negative coefficient indicates a growth tilt.
  - Note that the HML coefficient is positive. This does NOT mean that banks are "value". It means that bank-stock returns tend to move in the same direction as the HML factor; that is, they tend to perform relatively better when value stocks outperform growth stocks.

## Heteroscedasticity diagnostic

Task: Use the Breusch-Pagan test to examine whether the "focus" regression's residual variance is related to the factor regressors.

```{r}
#| label: breusch-pagan

focus_bp_test <- lmtest::bptest(focus_model)

focus_bp_table <- broom::tidy(focus_bp_test) |>
  select(statistic, parameter, p.value) |>
  rename(
    `BP statistic` = statistic,
    `Degrees of freedom` = parameter,
    `p-value` = p.value
  ) |>
  gt_compact(title = "Breusch-Pagan heteroscedasticity test") |>
  fmt_number(columns = where(is.numeric), decimals = 4)

focus_bp_table
```

Q: State the Breusch-Pagan null hypothesis. Evaluate. Why might we still prefer robust inference if the test does not reject?

A:

The null is that the conditional residual variance is constant.

Given the large pvalue, we fail to reject, suggesting homoscedasticity.

Failure to reject is not proof of homoscedasticity; the test may have limited power or miss forms of heteroscedasticity outside its auxiliary specification. Robust standard errors provide a modest safeguard without changing the OLS coefficients.

## Individual and joint tests

Task: Conduct a robust joint test of whether all three factor slopes equal zero for the focus industry.

```{r}
#| label: joint-test

focus_intercept_model <- lm(excess_return ~ 1, data = model.frame(focus_model))

focus_joint_test <- lmtest::waldtest(
  focus_intercept_model,
  focus_model,
  vcov = sandwich::vcovHC(focus_model, type = "HC1"),
  test = "Chisq"
)

focus_joint_table <- broom::tidy(focus_joint_test) |>
  gt_compact(
    title = "Joint test of the three factor slopes",
    subtitle = "H0: beta_MKT = beta_SMB = beta_HML = 0"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 4)

focus_joint_table
```

Q: Interpret the results.

A:

The p value is quite small, suggesting that we reject the null. i.e. there is evidence that the betas are not jointly equal to 0, which in our case implies that at least one of these factors helps to explain the cross section of excess returns.

Q: How does the joint test differ from the three individual slope tests?

A:

Each individual test asks whether one slope is zero while leaving the others unrestricted. The joint test asks whether all three restrictions hold simultaneously and accounts for covariance among the estimated slopes. It can reject even when some individual tests do not.

## Fitted values, residuals, and regression plots

Task: Create a time-series plot of actual and fitted excess returns and a residual-versus-fitted plot for the focus industry.

```{r}
#| label: focus-regression-plots

focus_diagnostics <- industry_models |>
  filter(industry == settings$focus_industry) |>
  transmute(
    diagnostics = map2(
      model,
      data,
      ~ broom::augment(.x, data = .y)
    )
  ) |>
  unnest(diagnostics)

actual_fitted_plot <- focus_diagnostics |>
  select(date, Actual = excess_return, Fitted = .fitted) |>
  pivot_longer(c(Actual, Fitted), names_to = "series", values_to = "return") |>
  ggplot(aes(date, return, color = series)) +
  geom_line(linewidth = 0.45, alpha = 0.85) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_color_manual(values = c(Actual = "black", Fitted = "#2166ac")) +
  labs(
    title = paste(settings$focus_industry, "actual and fitted excess returns"),
    x = NULL, y = "Monthly excess return", color = NULL
  )

residual_fitted_plot <- focus_diagnostics |>
  ggplot(aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.45, color = "#2166ac") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "#b2182b") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = paste(settings$focus_industry, "residuals versus fitted values"),
    x = "Fitted excess return", y = "Residual"
  )

actual_fitted_plot
residual_fitted_plot
```

Q: What do the plots suggest?

A:

- The fitted returns generally track the direction and broad variation in actual excess returns, although the model does not fully capture some of the largest movements.

- The residuals are centered around zero and do not exhibit a strong, systematic relationship with the fitted excess returns.

- The smooth red line is fairly flat near zero across most fitted values, although its slight curvature at the extremes suggests some modest nonlinearity that the model may not capture.

- There is no obvious funnel-shaped pattern, so the plot provides limited visual evidence of severe heteroskedasticity.

- Several large residuals remain, indicating months in which realized excess returns differed substantially from the returns predicted by the factor model.

- Overall, the linear specification appears broadly reasonable, but the model does not explain every return movement; particularly extreme outcomes.

## Influential observations

Task: Identify the five focus-industry months with the largest Cook's distances and label them on a diagnostic plot.

```{r}
#| label: influence-diagnostics

influential_months <- focus_diagnostics |>
  arrange(desc(.cooksd)) |>
  slice_head(n = 5) |>
  transmute(
    Date = date,
    `Actual excess return` = excess_return,
    `Studentized residual` = .std.resid,
    Leverage = .hat,
    `Cook's distance` = .cooksd
  )

influence_table <- influential_months |>
  mutate(`Actual excess return` = pct(`Actual excess return`)) |>
  gt_compact(title = paste("Most influential", settings$focus_industry, "months")) |>
  fmt_number(columns = c(`Studentized residual`, Leverage, `Cook's distance`), decimals = 3)

influence_plot <- focus_diagnostics |>
  mutate(label = if_else(
    date %in% influential_months$Date,
    format(date, "%b %Y"),
    NA_character_
  )) |>
  ggplot(aes(.hat, .std.resid, size = .cooksd)) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c(2, 1, 2), color = "grey60") +
  geom_point(alpha = 0.45, color = "#2166ac") +
  geom_text_repel(aes(label = label), na.rm = TRUE, size = 3, max.overlaps = Inf) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title = paste(settings$focus_industry, "influence diagnostics"),
    x = "Leverage", y = "Studentized residual", size = "Cook's distance"
  )

influence_table
influence_plot
```

Q: What does the Cook's distance measure?

A:

- Cook's distance measures the extent to which a single data point influences the results of the regression model.

- It asks how much the model's fitted values would change if observation $i$ were removed and the regression were re-estimated.

- Cook's distance is commonly defined as

  $$
  D_i
  =
  \frac{
  \sum_{j=1}^{n}
  \left(
  \hat{y}_j-\hat{y}_{j(-i)}
  \right)^2
  }{
  k\widehat{\sigma}^2
  },
  $$

  where $\hat{y}_j$ is the fitted value from the full model, $\hat{y}_{j(-i)}$ is the fitted value after removing observation $i$, $k$ is the number of estimated coefficients including the intercept, and $\widehat{\sigma}^2$ is the regression mean squared error.

- Cook's distance can also be calculated from the observation's residual and leverage:

  $$
  D_i
  =
  \frac{\hat{\varepsilon}_i^2}
       {k\widehat{\sigma}^2}
  \frac{h_{ii}}
       {(1-h_{ii})^2}.
  $$

- The leverage of observation $i$ is

  $$
  h_{ii}
  =
  \mathbf{x}_i'
  (X'X)^{-1}
  \mathbf{x}_i,
  $$

  where $\mathbf{x}_i$ contains the explanatory-variable values for observation $i$, including the intercept.

- Leverage measures how unusual an observation's explanatory-variable values are relative to the rest of the sample.

- Average leverage equals

  $$
  \bar{h}=\frac{k}{n}.
  $$

  Values substantially above the average—often values greater than $2k/n$ or $3k/n$—may be considered high leverage.

- An observation tends to have a large Cook's distance when it has:

  - a large residual, meaning its outcome is poorly predicted by the model;
  - high leverage, meaning its explanatory-variable values are unusual; or
  - a combination of both.

- A common screening guideline flags observations for which

  $$
  D_i>\frac{4}{n}.
  $$

  This is a diagnostic guideline rather than a formal statistical test.

Q: Should we delete the month with the largest Cook's distance?

A:

Not automatically. An influential observation may be a data error, but it may also be a genuine crisis or boom that is central to financial risk. Investigate the observation, confirm the data, explain its economic context, and report sensitivity if it materially changes the conclusion.

## Compare exposures across industries

Task: Plot the estimated factor exposures and robust 95% confidence intervals for all 48 industries.

```{r}
#| label: beta-plot

beta_plot <- first_pass_coefficients |>
  filter(term != "(Intercept)") |>
  mutate(
    factor = recode(term, MKT_RF = "Market", SMB = "Size", HML = "Value"),
    industry = fct_reorder(industry, estimate)
  ) |>
  ggplot(aes(estimate, industry)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    width = 0,
    color = "grey55",
    orientation = "y"
  ) +
  geom_point(color = "#2166ac", size = 1.6) +
  facet_wrap(~ factor, scales = "free_x", ncol = 3) +
  labs(
    title = "First-pass factor exposures by industry",
    subtitle = "HC1 robust 95% confidence intervals",
    x = "Estimated beta", y = NULL
  ) +
  theme(axis.text.y = element_text(size = 6))

plotly::ggplotly(beta_plot)
```

Q: What do you notice about the range of exposures for the market risk factor in contrast to size and value? What's the implication?

A:

All of the industries have decisively positive market risk exposures. By contrast, the betas for SMB and HML range in the positive and negative domains.

Diversifying across industries can't eliminate market risk.

But industry selection might be able to offset size and value exposures.

# Step 2A: Historical-average cross section

## Estimate average prices of risk

We first run one cross-sectional regression on the averages:

$$
\overline{R_i-R_f}
=
\lambda_0
+\lambda_{MKT}\widehat\beta_{i,MKT}
+\lambda_{SMB}\widehat\beta_{i,SMB}
+\lambda_{HML}\widehat\beta_{i,HML}
+\overline u_i.
$$

Task: Combine each industry's average excess return with its first-pass betas and estimate the historical-average cross-sectional regression.

```{r}
#| label: average-cross-section

betas <- first_pass_coefficients |>
  filter(term != "(Intercept)") |>
  transmute(
    industry,
    term = recode(term, MKT_RF = "beta_mkt", SMB = "beta_smb", HML = "beta_hml"),
    estimate
  ) |>
  pivot_wider(names_from = term, values_from = estimate)

average_cross_section <- panel |>
  group_by(industry) |>
  summarise(average_excess_return = mean(excess_return), .groups = "drop") |>
  left_join(betas, by = "industry")

average_pricing_model <- lm(
  average_excess_return ~ beta_mkt + beta_smb + beta_hml,
  data = average_cross_section
)

average_pricing_inference <- robust_tidy(average_pricing_model)
```

Task: Display the estimated historical-average prices of risk with HC1 robust inference.

```{r}
#| label: average-prices-table

average_pricing_table <- average_pricing_inference |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market",
      beta_smb = "Size",
      beta_hml = "Value"
    ),
    across(c(estimate, std.error, conf.low, conf.high), ~ 100 * .x)
  ) |>
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high) |>
  rename(
    Term = term,
    `Estimate (monthly ppts)` = estimate,
    `Robust SE` = std.error,
    `t-statistic` = statistic,
    `p-value` = p.value,
    `CI low` = conf.low,
    `CI high` = conf.high
  ) |>
  gt_compact(
    title = "Historical-average cross-sectional regression",
    subtitle = "48 industry portfolios; HC1 robust inference"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

average_pricing_table
```

Q: Interpret the coefficient on size.

A:

Holding market and value betas fixed, an industry with an SMB beta one unit higher has an estimated average monthly excess return that is $\widehat\lambda_{SMB}$ percentage points higher.

This is a cross-sectional association between exposure and average return,not the first-pass sensitivity of an industry's return to the realized SMB factor.

Q: What does a statistically insignificant price of risk mean?

A:

- Notice that Value is statistically insignificant, and market is marginally significant.
- The implication is that the market is not rewarding investors for these risk exposures.\
- That does not automatically mean that we ignore those factors.
- It means the exercise does not provide sufficiently precise evidence that average returns differ with that beta, conditional on the other betas. It does not prove that the factor is useless, that the true price is exactly zero, or that the factor does not explain time-series variation.

## Correlations and VIFs among estimated betas

Task: Examine correlations and VIFs among the estimated betas used as second-pass regressors.

```{r}
#| label: beta-collinearity

beta_correlations <- average_cross_section |>
  select(beta_mkt, beta_smb, beta_hml) |>
  cor()

beta_correlation_table <- beta_correlations |>
  as.data.frame() |>
  rownames_to_column("Beta") |>
  gt_compact(title = "Correlations among estimated industry betas") |>
  fmt_number(columns = where(is.numeric), decimals = 2)

beta_vif <- calculate_vif(
  average_cross_section,
  regressors = c("beta_mkt", "beta_smb", "beta_hml")
)

beta_vif_table <- beta_vif |>
  rename(
    Beta = variable,
    `Auxiliary R-squared` = auxiliary_r_squared,
    VIF = vif
  ) |>
  gt_compact(title = "Second-pass variance inflation factors") |>
  fmt_number(columns = c(`Auxiliary R-squared`, VIF), decimals = 2)

beta_correlation_table
beta_vif_table
```

Q: Why do we calculate a second set of correlations and VIFs?

A:

The two regression steps use different regressors. Factor-return correlations govern first-pass collinearity through time. Beta correlations govern second-pass collinearity across industries. Low collinearity in one step does not guarantee low collinearity in the other.

Q: Interpret the results.

A:

- The market and size betas have a moderate positive correlation of $0.55$.

- Thus, industry portfolios with higher estimated exposure to the market factor factor tend to have higher exposure to the SMB factor.

- The HML beta has little correlation with either the market beta or the size beta.

- The VIFs range from $1.03$ to $1.45$, well below commonly used thresholds for concern.

- Therefore, the estimated betas contain some shared information, particularly the market and size betas, but not enough to create a meaningful multicollinearity problem in the second-pass regression.

- If the estimated betas were highly correlated across industries, the factor exposures would be difficult to separate. For example, if industries with high market betas also had high SMB betas, it would be difficult to change a portfolio’s size exposure without also changing its market exposure.

- In the second-pass Fama–MacBeth regression, highly correlated betas would also create multicollinearity, making the estimated factor risk premiums less precise and their individual effects harder to distinguish.

## Plot the cross-sectional regression

With three beta regressors, the fitted relation is a hyperplane rather than a line. We therefore plot fitted against actual average returns.

Task: Plot each industry's actual average excess return against its fitted value and label the most poorly priced industries.

```{r}
#| label: cross-sectional-plot

average_diagnostics <- broom::augment(
  average_pricing_model,
  data = average_cross_section
) |>
  mutate(
    absolute_pricing_error = abs(.resid),
    label = if_else(
      min_rank(desc(absolute_pricing_error)) <= 6,
      industry,
      NA_character_
    )
  )

# coord_equal() below scales the panel so that fitted returns
# (a narrow range) and actual returns (a wider range) share one axis
# unit; giving both scales the same limits keeps the panel from
# collapsing into a tall, narrow strip.
pricing_error_range <- range(
  c(average_diagnostics$.fitted, average_diagnostics$average_excess_return)
)

average_cross_section_plot <- ggplot(
  average_diagnostics,
  aes(.fitted, average_excess_return)
) +
  geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = 2) +
  geom_point(color = "#2166ac", size = 2) +
  geom_text_repel(aes(label = label), na.rm = TRUE, size = 3, max.overlaps = Inf) +
  scale_x_continuous(limits = pricing_error_range, labels = label_percent(accuracy = 0.01)) +
  scale_y_continuous(limits = pricing_error_range, labels = label_percent(accuracy = 0.01)) +
  labs(
    title = "Actual versus model-implied average excess returns",
    subtitle = "Labels identify the six largest absolute pricing errors",
    x = "Fitted average monthly excess return",
    y = "Actual average monthly excess return"
  ) +
  coord_equal()

average_cross_section_plot
```

Q: What does a point above the 45-degree line represent?

A:

The industry's actual average excess return exceeds the return implied by its estimated factor exposures. Its cross-sectional residual, or pricing error, is positive.

This might suggest that we need to search for omitted factors that can assist with pricing.

# Step 2B: Month-by-month Fama-MacBeth regressions

## Estimate monthly prices of risk

Task: Attach the fixed first-pass betas to every industry-month and estimate one cross-sectional regression in each month.

```{r}
#| label: monthly-cross-sections

monthly_pricing_data <- panel |>
  select(date, industry, excess_return) |>
  left_join(betas, by = "industry")

monthly_pricing_models <- monthly_pricing_data |>
  group_nest(date) |>
  mutate(
    model = purrr::map(
      data,
      ~ lm(excess_return ~ beta_mkt + beta_smb + beta_hml, data = .x)
    ),
    coefficients = purrr::map(model, broom::tidy),
    fit = purrr::map(model, broom::glance)
  )

lambda_time_series <- monthly_pricing_models |>
  select(date, coefficients) |>
  unnest(coefficients) |>
  select(date, term, estimate)
```

Task: For each coefficient, estimate its time-series mean and a Newey-West standard error for that mean.

```{r}
#| label: fama-macbeth-inference

fama_macbeth_summary <- lambda_time_series |>
  group_by(term) |>
  group_modify(~ {
    mean_model <- lm(estimate ~ 1, data = .x)
    nw_variance <- sandwich::NeweyWest(
      mean_model,
      lag = settings$nw_lag,
      prewhite = FALSE,
      adjust = TRUE
    )
    mean_estimate <- unname(coef(mean_model)[1])
    standard_error <- sqrt(nw_variance[1, 1])
    t_statistic <- mean_estimate / standard_error

    tibble(
      months = nrow(.x),
      mean = mean_estimate,
      nw_se = standard_error,
      t_statistic = t_statistic,
      p_value = 2 * pt(-abs(t_statistic), df = nrow(.x) - 1),
      conf_low = mean_estimate - qt(0.975, df = nrow(.x) - 1) * standard_error,
      conf_high = mean_estimate + qt(0.975, df = nrow(.x) - 1) * standard_error
    )
  }) |>
  ungroup()

fama_macbeth_table <- fama_macbeth_summary |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    ),
    across(c(mean, nw_se, conf_low, conf_high), ~ 100 * .x)
  ) |>
  rename(
    Term = term,
    Months = months,
    `Mean monthly price (ppts)` = mean,
    `Newey-West SE` = nw_se,
    `t-statistic` = t_statistic,
    `p-value` = p_value,
    `CI low` = conf_low,
    `CI high` = conf_high
  ) |>
  gt_compact(
    title = "Fama-MacBeth estimates",
    subtitle = paste("Newey-West inference for mean monthly prices; lag =", settings$nw_lag)
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

fama_macbeth_table
```

Q: What is being tested in the Fama-MacBeth table? Interpret the results.

A:

For each beta characteristic, the null is that its expected monthly cross-sectional price equals zero: $H_0:E[\lambda_{k,t}]=0$. The test uses the time-series variation in the monthly estimates. The Newey-West adjustment allows those estimated prices to be heteroscedastic and serially correlated through the selected lag.

All of the factor lambdas are statistically significant, implying that on average exposure to the Market, Size, and Value factors are rewarded.

## Compare the two second-pass approaches

Task: Compare coefficients from the historical-average cross section with the means of the monthly cross-sectional coefficients.

```{r}
#| label: compare-second-pass

second_pass_comparison <- average_pricing_inference |>
  select(term, historical_average_coefficient = estimate) |>
  left_join(
    fama_macbeth_summary |>
      select(term, mean_monthly_coefficient = mean),
    by = "term"
  ) |>
  mutate(
    difference = historical_average_coefficient - mean_monthly_coefficient,
    across(c(historical_average_coefficient, mean_monthly_coefficient, difference), ~ 100 * .x),
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    )
  )

second_pass_comparison_table <- second_pass_comparison |>
  rename(
    Term = term,
    `Average-return regression (ppts)` = historical_average_coefficient,
    `Mean monthly coefficient (ppts)` = mean_monthly_coefficient,
    `Difference (ppts)` = difference
  ) |>
  gt_compact(title = "Two ways to obtain average prices of risk") |>
  fmt_number(columns = where(is.numeric), decimals = 8)

second_pass_comparison_table
```

Q: If the point estimates are effectively identical, why run the month-by-month regressions?

A:

The monthly regressions reveal how prices of risk vary through time and supply the time series used for Fama-MacBeth inference. The historical-average regression instead treats the 48 industries as its observations and uses cross-sectional residual variation for its standard errors. Identical point estimates do not imply identical uncertainty estimates.

## Plot prices of risk through time

Task: Plot each monthly price of risk and its trailing 12-month average.

```{r}
#| label: lambda-time-series-plot

lambda_plot_data <- lambda_time_series |>
  group_by(term) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    trailing_12_month_mean = zoo::rollmean(
      estimate,
      k = 12,
      fill = NA,
      align = "right"
    ),
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    )
  ) |>
  ungroup()

lambda_time_series_plot <- ggplot(lambda_plot_data, aes(date, estimate)) +
  geom_hline(yintercept = 0, color = "grey55") +
  geom_line(alpha = 0.22, linewidth = 0.35, color = "grey35") +
  geom_line(aes(y = trailing_12_month_mean), color = "#2166ac", linewidth = 0.7) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Monthly prices of risk",
    subtitle = "Grey: monthly estimate; blue: trailing 12-month mean",
    x = NULL, y = "Monthly price of risk"
  )

lambda_time_series_plot
```

Q: Why can a factor have a positive average price of risk even though its estimated monthly price is often negative?

A:

A positive mean is a statement about the center of a time-varying distribution, not a claim that every realization is positive. Risk prices can be noisy and state-dependent. Investors may earn compensation on average while experiencing many months in which the cross-sectional relation reverses.

# Extension: Do prices of risk change in recessions?

We use the monthly NBER recession indicator distributed by FRED as `USREC`. For each estimated price of risk, we fit

$$
\widehat\lambda_{k,t}=\delta_{0,k}+\delta_{1,k}Recession_t+v_{k,t}.
$$

Q: How interpret the $\delta_{0,k}$ and $\delta_{1,k}$?

A:

$\delta_{0,k}$ is the mean price during expansions

$\delta_{1,k}$ is the recession-minus-expansion difference

Task: Download the monthly recession indicator from FRED and join it to the monthly prices of risk.

```{r}
#| label: recession-data

recession_file <- file.path(settings$data_dir, "USREC.csv")
recession_url <- "https://fred.stlouisfed.org/graph/fredgraph.csv?id=USREC"

download_if_needed(recession_url, recession_file, settings$refresh_data)

recessions <- read_csv(recession_file, show_col_types = FALSE) |>
  transmute(
    date = as.Date(observation_date),
    recession = as.integer(USREC)
  )

lambda_events <- lambda_time_series |>
  inner_join(recessions, by = "date") |>
  mutate(state = if_else(recession == 1, "Recession", "Expansion"))
```

Q: Why can we not add `recession` directly to each month's cross-sectional regression?

A:

Within a given month, every industry has the same recession value. The dummy is therefore constant across the 48 observations and perfectly collinear with that month's intercept. It contains no within-month cross-sectional variation. After we create a time series of monthly lambdas, the indicator varies through time and can be used as a time-series regressor.

Task: Estimate the expansion mean and recession-minus-expansion difference for each price of risk using Newey-West inference.

```{r}
#| label: recession-models

recession_models <- lambda_events |>
  group_nest(term) |>
  mutate(
    model = purrr::map(data, ~ lm(estimate ~ recession, data = .x)),
    inference = purrr::map(
      model,
      ~ robust_tidy(
          .x,
          sandwich::NeweyWest(
            .x,
            lag = settings$nw_lag,
            prewhite = FALSE,
            adjust = TRUE
          )
        ) |>
        rename(parameter = term)
    )
  )

recession_inference <- recession_models |>
  select(term, inference) |>
  unnest(inference) |>
  mutate(
    price = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    ),
    parameter = recode(
      parameter,
      `(Intercept)` = "Expansion mean",
      recession = "Recession minus expansion"
    )
  )
```

Task: Display the recession-regression estimates.

```{r}
#| label: recession-table

recession_table <- recession_inference |>
  mutate(across(c(estimate, std.error, conf.low, conf.high), ~ 100 * .x)) |>
  select(price, parameter, estimate, std.error, statistic, p.value, conf.low, conf.high) |>
  rename(
    Price = price,
    Parameter = parameter,
    `Estimate (monthly ppts)` = estimate,
    `Newey-West SE` = std.error,
    `t-statistic` = statistic,
    `p-value` = p.value,
    `CI low` = conf.low,
    `CI high` = conf.high
  ) |>
  gt_compact(
    title = "Prices of risk in expansions and recessions",
    subtitle = paste("Newey-West inference; lag =", settings$nw_lag)
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

recession_table
```

Task: Plot the distribution of each estimated price of risk in expansions and recessions.

```{r}
#| label: recession-plot

recession_plot <- lambda_events |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    )
  ) |>
  ggplot(aes(state, estimate, fill = state)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c(Expansion = "#9ecae1", Recession = "#de2d26")) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Monthly prices of risk by economic state",
    x = NULL, y = "Monthly price of risk", fill = NULL
  ) +
  theme(legend.position = "none")

recession_plot
```

Q: Interpret a positive and statistically significant recession coefficient for the market price of risk.

A:

The average monthly cross-sectional return associated with one unit of market beta is estimated to be higher in NBER recession months than in expansion months, and the difference is statistically distinguishable from zero under the model and Newey-West inference. This is an association with recession timing, not by itself a causal effect of recessions.

# Synthesis

Q: Can the first-pass market beta be statistically significant for nearly every industry while the second-pass market price of risk is statistically insignificant?

A:

Yes. The first pass asks whether industry returns co-move with monthly market shocks. The second pass asks whether industries with larger estimated market betas earn systematically different average returns. A factor can explain substantial time-series variation without explaining the cross section of average returns precisely.

Q: What are the main limitations of this exercise?

A:

- The first-pass betas are estimated rather than observed, so the second pass has an errors-in-variables problem. The reported exercise does not implement a Shanken correction.
- Full-sample betas use future information relative to earlier months. This is acceptable for a descriptive asset-pricing exercise but not for an investable backtest.
- Forty-eight industries provide a small and highly aggregated cross section.
- Betas may vary through time, but the main exercise treats them as constant.
- Statistical significance does not establish that a relation is economically large, stable, or exploitable after costs.

## Investment Implications

Q: How use these results to inform your investment decisions?

A:

- Identify rewarded factor exposures. The estimated lambdas indicate which factor exposures have historically been associated with higher average returns across industries. Investors should consider each lambda’s sign, economic magnitude, statistical significance, and stability, not simply select the largest estimate.
- Identify industries with the desired exposures. The estimated betas show which industries are most sensitive to each factor. If an investor expects a factor’s positive risk premium to persist, industries with larger positive betas provide greater exposure to that factor.
- Estimate factor-implied required returns. For industry (i), the model-implied expected excess return is

$$
\widehat{\mathbb{E}\!\left[R_i - R_f\right]}
=
\widehat{\lambda}_0
+
\widehat{\beta}_{i,\mathrm{MKT}}\widehat{\lambda}_{\mathrm{MKT}}
+
\widehat{\beta}_{i,\mathrm{SMB}}\widehat{\lambda}_{\mathrm{SMB}}
+
\widehat{\beta}_{i,\mathrm{HML}}\widehat{\lambda}_{\mathrm{HML}}.
$$ - Compare model-implied and subjective expected returns. If the investor’s forecast exceeds the factor-implied required return, the industry may appear relatively attractive; if it falls below the required return, it may appear relatively unattractive. This comparison is a screening tool rather than definitive evidence of mispricing. - Recognize that higher expected return accompanies higher risk exposure. A high beta to a rewarded factor does not provide a free return. The higher expected return is interpreted as compensation for bearing more systematic risk.