---
title: "Regression and the Cross-Section of Industry Returns"
subtitle: "A Fama-MacBeth exercise with 48 industry portfolios"
author: Mike Aguilar
format:
  html:
    toc: true
    toc-depth: 5
    toc-location: left
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: true
---

# Background

Q: Suppose you are an equity analyst. Your boss asking you to "run a factor analysis on these stocks".

- Explain the concept of a "factor" in the equity markets.
- Define the meaning of Mkt-RF, SMB, HML.
- How use for investment analysis?

A:

## Architecture and notation

We study the question using:

- the monthly **Ken French 48 Industry Portfolios** as test assets; and
- the monthly **Fama-French three factors**: market excess return, size, and value.

The exercise has two main steps.

1.  **Time-series regressions:** estimate each industry's factor exposures.
2.  **Cross-sectional regressions:** estimate the return associated with each unit of factor exposure.

This is an **asset-pricing exercise**, not a trading backtest. We use the full sample to estimate first-pass betas so that the mechanics are transparent. A genuinely investable exercise would estimate betas using only information available at each portfolio-formation date.

The first-pass regression for industry $i$ is

$$
R_{i,t}-R_{f,t}
=
\alpha_i
+\beta_{i,MKT}(R_{M,t}-R_{f,t})
+\beta_{i,SMB}SMB_t
+\beta_{i,HML}HML_t
+\varepsilon_{i,t}.
$$

The first pass uses variation **through time** to estimate one set of exposures for each industry.

The second-pass regression for month $t$ is

$$
R_{i,t}-R_{f,t}
=
\lambda_{0,t}
+\lambda_{MKT,t}\widehat\beta_{i,MKT}
+\lambda_{SMB,t}\widehat\beta_{i,SMB}
+\lambda_{HML,t}\widehat\beta_{i,HML}
+u_{i,t}.
$$

The second pass uses variation **across industries**. Repeating it each month produces a time series of estimated prices of risk, $\widehat\lambda_{k,t}$.

| Quantity | Varies across | Interpretation |
|------------------------|------------------------|------------------------|
| Factor return, $F_{k,t}$ | Months | The realized factor payoff in month $t$ |
| Factor beta, $\beta_{i,k}$ | Industries | Industry $i$'s sensitivity to factor $k$ |
| Factor price, $\lambda_{k,t}$ | Months | Cross-sectional return associated with one unit of beta exposure in month $t$ |

Q: What is the meaning of $\beta$?

A:

Q: What is the meaning of $\lambda$?

A:

# Housekeeping

Task: Reset the R session, load the required packages, define the analysis settings, and create small formatting and regression helpers. Let's define "Banks" as the "focus" industry; i.e. the industry that we will use in this exercise to zoom in on and explore in more depth.

```{r}
#| label: setup

cat("\014")
rm(list = ls())

required_packages <- c(
  "tidyverse", "lubridate", "broom", "sandwich", "lmtest",
  "gt", "scales", "zoo", "ggrepel"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(broom)
  library(sandwich)
  library(lmtest)
  library(gt)
  library(scales)
  library(zoo)
  library(ggrepel)
})

theme_set(theme_minimal(base_size = 12))

settings <- list(
  start_date = as.Date("1970-01-01"),
  end_date = Sys.Date(),
  focus_industry = "Banks",
  nw_lag = 6,
  refresh_data = FALSE,
  data_dir = "data"
)

dir.create(settings$data_dir, showWarnings = FALSE, recursive = TRUE)

pct <- function(x, accuracy = 0.01) {
  scales::percent(x, accuracy = accuracy)
}

ppts <- function(x, digits = 2) {
  paste0(format(round(100 * x, digits), nsmall = digits), " ppts")
}

gt_compact <- function(data, title = NULL, subtitle = NULL) {
  out <- gt(data)
  if (!is.null(title)) {
    out <- tab_header(out, title = title, subtitle = subtitle)
  }
  out |>
    sub_missing(missing_text = "—") |>
    cols_align(align = "center") |>
    opt_row_striping()
}

robust_tidy <- function(model, vcov_matrix = sandwich::vcovHC(model, type = "HC1")) {
  result <- broom::tidy(lmtest::coeftest(model, vcov. = vcov_matrix))
  critical_value <- qt(0.975, df = df.residual(model))
  result |>
    mutate(
      conf.low = estimate - critical_value * std.error,
      conf.high = estimate + critical_value * std.error
    )
}

calculate_vif <- function(data, regressors) {
  map_dfr(regressors, function(target) {
    other_regressors <- setdiff(regressors, target)
    auxiliary_model <- lm(
      reformulate(other_regressors, response = target),
      data = data
    )
    auxiliary_r_squared <- summary(auxiliary_model)$r.squared

    tibble(
      variable = target,
      auxiliary_r_squared = auxiliary_r_squared,
      vif = 1 / (1 - auxiliary_r_squared)
    )
  })
}
```

# Data

## Download and parse

The data come from the [Kenneth R. French Data Library](https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html). The source files report returns in percentage points; this exercise converts them to decimal returns.

Task: Define helpers that download and extract the first monthly data block from a Ken French ZIP file.

```{r}
#| label: data-helpers

download_if_needed <- function(url, destination, refresh = FALSE) {
  if (refresh || !file.exists(destination)) {
    download.file(url, destination, mode = "wb", quiet = TRUE)
  }
  invisible(destination)
}

read_first_monthly_block <- function(zip_path) {
  member <- unzip(zip_path, list = TRUE)$Name[1]
  connection <- unz(zip_path, member, open = "rt")
  on.exit(close(connection), add = TRUE)
  lines <- readLines(connection, warn = FALSE)

  first_data_row <- which(str_detect(lines, "^\\s*\\d{6},"))[1]
  header_candidates <- which(
    seq_along(lines) < first_data_row &
      str_detect(lines, "^\\s*,")
  )
  header_row <- max(header_candidates)

  candidate_rows <- seq(first_data_row, length(lines))
  first_nonmonthly <- which(
    !str_detect(lines[candidate_rows], "^\\s*\\d{6},")
  )[1]

  last_data_row <- if (is.na(first_nonmonthly)) {
    length(lines)
  } else {
    candidate_rows[first_nonmonthly] - 1
  }

  csv_text <- paste(
    c(lines[header_row], lines[first_data_row:last_data_row]),
    collapse = "\n"
  )

  readr::read_csv(
    I(csv_text),
    show_col_types = FALSE,
    trim_ws = TRUE,
    name_repair = "minimal"
  )
}
```

Task: Download the monthly Fama-French three-factor and 48-industry files, then extract their monthly value-weighted returns.

```{r}
#| label: download-data

factor_url <- paste0(
  "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/",
  "F-F_Research_Data_Factors_CSV.zip"
)

industry_url <- paste0(
  "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/",
  "48_Industry_Portfolios_CSV.zip"
)

factor_zip <- file.path(settings$data_dir, "F-F_Research_Data_Factors_CSV.zip")
industry_zip <- file.path(settings$data_dir, "48_Industry_Portfolios_CSV.zip")

download_if_needed(factor_url, factor_zip, settings$refresh_data)
download_if_needed(industry_url, industry_zip, settings$refresh_data)

factor_raw <- read_first_monthly_block(factor_zip)
industry_raw <- read_first_monthly_block(industry_zip)
```

Task: Clean the two data sets, convert percentage-point returns to decimals, and join them into an industry-month panel.

```{r}
#| label: clean-data

names(factor_raw) <- str_trim(names(factor_raw))
names(industry_raw) <- str_trim(names(industry_raw))

factors <- factor_raw |>
  rename(
    date_code = 1,
    MKT_RF = `Mkt-RF`
  ) |>
  mutate(
    date = as.Date(paste0(date_code, "01"), format = "%Y%m%d"),
    across(c(MKT_RF, SMB, HML, RF), ~ as.numeric(.x) / 100)
  ) |>
  select(date, MKT_RF, SMB, HML, RF) |>
  filter(between(date, settings$start_date, settings$end_date))

industries <- industry_raw |>
  rename(date_code = 1) |>
  mutate(date = as.Date(paste0(date_code, "01"), format = "%Y%m%d")) |>
  select(-date_code) |>
  pivot_longer(
    cols = -date,
    names_to = "industry",
    values_to = "industry_return"
  ) |>
  mutate(
    industry = str_trim(industry),
    industry_return = as.numeric(industry_return),
    industry_return = na_if(industry_return, -99.99),
    industry_return = na_if(industry_return, -999),
    industry_return = industry_return / 100
  ) |>
  filter(between(date, settings$start_date, settings$end_date))

panel <- industries |>
  inner_join(factors, by = "date") |>
  mutate(excess_return = industry_return - RF) |>
  drop_na(excess_return, MKT_RF, SMB, HML)

data_overview <- tibble(
  Item = c("First month", "Last month", "Months", "Industries", "Panel observations"),
  Value = c(
    format(min(panel$date), "%B %Y"),
    format(max(panel$date), "%B %Y"),
    n_distinct(panel$date),
    n_distinct(panel$industry),
    nrow(panel)
  )
)

data_overview |>
  gt_compact(title = "Analysis sample")
```

## Validate

Task: Check for missing observations, duplicate industry-months, and an unbalanced panel.

```{r}
#| label: validate-data

expected_months <- n_distinct(panel$date)

validation <- bind_rows(
  tibble(
    check = "No missing model variables",
    passed = !anyNA(panel |> select(excess_return, MKT_RF, SMB, HML)),
    detail = paste(sum(!complete.cases(panel |> select(excess_return, MKT_RF, SMB, HML))), "incomplete rows")
  ),
  tibble(
    check = "One observation per industry-month",
    passed = !any(duplicated(panel[c("industry", "date")])),
    detail = paste(sum(duplicated(panel[c("industry", "date")])), "duplicates")
  ),
  panel |>
    dplyr::count(industry, name = "months") |>
    summarise(
      check = "Balanced industry panel",
      passed = all(months == expected_months),
      detail = paste(min(months), "to", max(months), "months per industry")
    )
)

if (!all(validation$passed)) {
  stop("At least one data-validation check failed.")
}

validation_table <- validation |>
  mutate(Status = if_else(passed, "PASS", "FAIL")) |>
  select(Check = check, Status, Detail = detail) |>
  gt_compact(title = "Data validation")

validation_table
```

Q: Why is a balanced panel especially useful when we compare the historical-average cross-sectional regression with the average of the monthly cross-sectional regressions?

A:

# Exploratory data analysis

## Return distributions

Task: Summarize the monthly factors and industry excess returns.

```{r}
#| label: summary-statistics

factor_summary <- factors |>
  select(MKT_RF, SMB, HML, RF) |>
  pivot_longer(everything(), names_to = "series", values_to = "return") |>
  group_by(series) |>
  summarise(
    observations = n(),
    mean = mean(return),
    volatility = sd(return),
    minimum = min(return),
    maximum = max(return),
    .groups = "drop"
  )

industry_summary <- panel |>
  group_by(industry) |>
  summarise(
    mean_excess_return = mean(excess_return),
    volatility = sd(excess_return),
    .groups = "drop"
  )

factor_summary_table <- factor_summary |>
  mutate(across(c(mean, volatility, minimum, maximum), ~ pct(.x))) |>
  rename(
    Series = series,
    Observations = observations,
    Mean = mean,
    Volatility = volatility,
    Minimum = minimum,
    Maximum = maximum
  ) |>
  gt_compact(title = "Monthly factor summary")

factor_summary_table
```

## Correlations among first-pass regressors

Task: Compute and visualize correlations among the three factor returns used as first-pass regressors.

```{r}
#| label: factor-correlations

factor_correlations <- factors |>
  select(MKT_RF, SMB, HML) |>
  cor(use = "complete.obs")

factor_correlation_table <- factor_correlations |>
  as.data.frame() |>
  rownames_to_column("Factor") |>
  gt_compact(title = "Correlations among factor returns") |>
  fmt_number(columns = where(is.numeric), decimals = 2)

factor_correlation_plot <- factor_correlations |>
  as.data.frame() |>
  rownames_to_column("factor_x") |>
  pivot_longer(-factor_x, names_to = "factor_y", values_to = "correlation") |>
  ggplot(aes(factor_x, factor_y, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = number(correlation, accuracy = 0.01)), size = 4) +
  scale_fill_gradient2(
    low = "#b2182b", mid = "white", high = "#2166ac",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(x = NULL, y = NULL, fill = "Correlation") +
  coord_equal()

factor_correlation_table

factor_correlation_plot
```

Q: Interpret as it relates to our case study.

A:

## Variance inflation factors

Task: Compute a VIF for each first-pass regressor. For each factor, regress it on the other two factors and use $VIF_k=1/(1-R_k^2)$.

```{r}
#| label: factor-vif

factor_vif <- calculate_vif(
  factors,
  regressors = c("MKT_RF", "SMB", "HML")
)

factor_vif_table <- factor_vif |>
  mutate(
    interpretation = case_when(
      vif > 10 ~ "Severe by the conventional rule of thumb",
      vif > 5 ~ "Potentially concerning",
      TRUE ~ "Limited inflation"
    )
  ) |>
  rename(
    Factor = variable,
    `Auxiliary R-squared` = auxiliary_r_squared,
    VIF = vif,
    Interpretation = interpretation
  ) |>
  gt_compact(title = "First-pass variance inflation factors") |>
  fmt_number(columns = c(`Auxiliary R-squared`, VIF), decimals = 2)

factor_vif_table
```

Q: Any concerns?

A:

# Step 1: Estimate factor exposures

## Fit 48 time-series regressions

Task: Estimate one Fama-French three-factor time-series regression for each industry. Store each fitted model and compute heteroscedasticity-robust coefficient estimates.

```{r}
#| label: first-pass-models

industry_models <- panel |>
  select(?) |>
  group_nest(industry) |>
  mutate(
    model = purrr::map(
      data,
      ~ ?, data = .x)
    ),
    coefficients = purrr::map(model, robust_tidy),
    fit = purrr::map(model, broom::glance)
  )

first_pass_coefficients <- industry_models |>
  select(industry, coefficients) |>
  unnest(coefficients)

first_pass_fit <- industry_models |>
  select(industry, fit) |>
  unnest(fit)
```

## Interpret one industry regression

Task: Display conventional and heteroscedasticity-robust inference for the focus industry.

```{r}
#| label: focus-industry-inference

if (!settings$focus_industry %in% industry_models$industry) {
  settings$focus_industry <- sort(industry_models$industry)[1]
}

focus_model <- industry_models |>
  filter(industry == settings$focus_industry) |>
  pull(model) |>
  pluck(1)

focus_conventional <- broom::tidy(focus_model) |>
  select(term, conventional_se = std.error, conventional_p = p.value)

focus_robust <- robust_tidy(focus_model) |>
  select(term, estimate, robust_se = std.error, robust_p = p.value, conf.low, conf.high)

focus_inference <- focus_robust |>
  left_join(focus_conventional, by = "term") |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Alpha",
      MKT_RF = "Market beta",
      SMB = "SMB beta",
      HML = "HML beta"
    )
  )

focus_inference_table <- focus_inference |>
  mutate(
    across(
      c(estimate, conventional_se, robust_se, conf.low, conf.high),
      ~ if_else(term == "Alpha", 100 * .x, .x)
    )
  ) |>
  rename(
    Term = term,
    Estimate = estimate,
    `Conventional SE` = conventional_se,
    `Conventional p-value` = conventional_p,
    `Robust SE` = robust_se,
    `Robust p-value` = robust_p,
    `Robust CI low` = conf.low,
    `Robust CI high` = conf.high
  ) |>
  gt_compact(
    title = paste(settings$focus_industry, "first-pass regression"),
    subtitle = "Alpha is shown in monthly percentage points; betas are unitless"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

focus_inference_table
```

Q: What is meant by the "robust" inference metrics?

A:

Q: Comment on the model's goodness of fit.

A:

Q: Interpret each coefficient in the focus-industry regression.

A:

## Heteroscedasticity diagnostic

Task: Use the Breusch-Pagan test to examine whether the "focus" regression's residual variance is related to the factor regressors.

```{r}
#| label: breusch-pagan

focus_bp_test <- lmtest::?(?)

focus_bp_table <- broom::tidy(focus_bp_test) |>
  select(statistic, parameter, p.value) |>
  rename(
    `BP statistic` = statistic,
    `Degrees of freedom` = parameter,
    `p-value` = p.value
  ) |>
  gt_compact(title = "Breusch-Pagan heteroscedasticity test") |>
  fmt_number(columns = where(is.numeric), decimals = 4)

focus_bp_table
```

Q: State the Breusch-Pagan null hypothesis. Evaluate. Why might we still prefer robust inference if the test does not reject?

A:

## Individual and joint tests

Task: Conduct a robust joint test of whether all three factor slopes equal zero for the focus industry.

```{r}
#| label: joint-test

focus_intercept_model <- ? data = model.frame(focus_model))

focus_joint_test <- lmtest::?(
  focus_intercept_model,
  focus_model,
  vcov = sandwich::vcovHC(focus_model, type = "HC1"),
  test = "Chisq"
)

focus_joint_table <- broom::tidy(focus_joint_test) |>
  gt_compact(
    title = "Joint test of the three factor slopes",
    subtitle = "H0: beta_MKT = beta_SMB = beta_HML = 0"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 4)

focus_joint_table
```

Q: Interpret the results.

A:

Q: How does the joint test differ from the three individual slope tests?

A:

## Fitted values, residuals, and regression plots

Task: Create a time-series plot of actual and fitted excess returns and a residual-versus-fitted plot for the focus industry.

```{r}
#| label: focus-regression-plots

focus_diagnostics <- industry_models |>
  filter(industry == settings$focus_industry) |>
  transmute(
    diagnostics = map2(
      model,
      data,
      ~ broom::augment(.x, data = .y)
    )
  ) |>
  unnest(diagnostics)

actual_fitted_plot <- focus_diagnostics |>
  select(date, Actual = excess_return, Fitted = .fitted) |>
  pivot_longer(c(Actual, Fitted), names_to = "series", values_to = "return") |>
  ggplot(aes(date, return, color = series)) +
  geom_line(linewidth = 0.45, alpha = 0.85) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_color_manual(values = c(Actual = "black", Fitted = "#2166ac")) +
  labs(
    title = paste(settings$focus_industry, "actual and fitted excess returns"),
    x = NULL, y = "Monthly excess return", color = NULL
  )

residual_fitted_plot <- focus_diagnostics |>
  ggplot(aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.45, color = "#2166ac") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "#b2182b") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = paste(settings$focus_industry, "residuals versus fitted values"),
    x = "Fitted excess return", y = "Residual"
  )

actual_fitted_plot
residual_fitted_plot
```

Q: What do the plots suggest?

A:

## Influential observations

Task: Identify the five focus-industry months with the largest Cook's distances and label them on a diagnostic plot.

```{r}
#| label: influence-diagnostics

influential_months <- focus_diagnostics |>
  arrange(desc(?)) |>
  slice_head(n = 5) |>
  transmute(
    Date = date,
    `Actual excess return` = excess_return,
    `Studentized residual` = .std.resid,
    Leverage = .hat,
    `Cook's distance` = ?
  )

influence_table <- influential_months |>
  mutate(`Actual excess return` = pct(`Actual excess return`)) |>
  gt_compact(title = paste("Most influential", settings$focus_industry, "months")) |>
  fmt_number(columns = c(`Studentized residual`, Leverage, `Cook's distance`), decimals = 3)

influence_plot <- focus_diagnostics |>
  mutate(label = if_else(
    date %in% influential_months$Date,
    format(date, "%b %Y"),
    NA_character_
  )) |>
  ggplot(aes(.hat, .std.resid, size = .cooksd)) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c(2, 1, 2), color = "grey60") +
  geom_point(alpha = 0.45, color = "#2166ac") +
  geom_text_repel(aes(label = label), na.rm = TRUE, size = 3, max.overlaps = Inf) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title = paste(settings$focus_industry, "influence diagnostics"),
    x = "Leverage", y = "Studentized residual", size = "Cook's distance"
  )

influence_table
influence_plot
```

Q: What does the Cook's distance measure?

A:

Q: Should we delete the month with the largest Cook's distance?

A:

## Compare exposures across industries

Task: Plot the estimated factor exposures and robust 95% confidence intervals for all 48 industries.

```{r}
#| label: beta-plot

beta_plot <- first_pass_coefficients |>
  filter(term != "(Intercept)") |>
  mutate(
    factor = recode(term, MKT_RF = "Market", SMB = "Size", HML = "Value"),
    industry = fct_reorder(industry, estimate)
  ) |>
  ggplot(aes(estimate, industry)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    width = 0,
    color = "grey55",
    orientation = "y"
  ) +
  geom_point(color = "#2166ac", size = 1.6) +
  facet_wrap(~ factor, scales = "free_x", ncol = 3) +
  labs(
    title = "First-pass factor exposures by industry",
    subtitle = "HC1 robust 95% confidence intervals",
    x = "Estimated beta", y = NULL
  ) +
  theme(axis.text.y = element_text(size = 6))

plotly::ggplotly(beta_plot)
```

Q: What do you notice about the range of exposures for the market risk factor in contrast to size and value? What's the implication?

A:

# Step 2A: Historical-average cross section

## Estimate average prices of risk

We first run one cross-sectional regression on the averages:

$$
\overline{R_i-R_f}
=
\lambda_0
+\lambda_{MKT}\widehat\beta_{i,MKT}
+\lambda_{SMB}\widehat\beta_{i,SMB}
+\lambda_{HML}\widehat\beta_{i,HML}
+\overline u_i.
$$

Task: Combine each industry's average excess return with its first-pass betas and estimate the historical-average cross-sectional regression.

```{r}
#| label: average-cross-section

betas <- first_pass_coefficients |>
  filter(term != "(Intercept)") |>
  transmute(
    industry,
    term = recode(term, MKT_RF = "beta_mkt", SMB = "beta_smb", HML = "beta_hml"),
    estimate
  ) |>
  pivot_wider(names_from = term, values_from = estimate)

average_cross_section <- panel |>
  group_by(industry) |>
  summarise(average_excess_return = ?, .groups = "drop") |>
  left_join(betas, by = "industry")

average_pricing_model <- lm(
  ?,
  data = ?
)

average_pricing_inference <- robust_tidy(average_pricing_model)
```

Task: Display the estimated historical-average prices of risk with HC1 robust inference.

```{r}
#| label: average-prices-table

average_pricing_table <- average_pricing_inference |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market",
      beta_smb = "Size",
      beta_hml = "Value"
    ),
    across(c(estimate, std.error, conf.low, conf.high), ~ 100 * .x)
  ) |>
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high) |>
  rename(
    Term = term,
    `Estimate (monthly ppts)` = estimate,
    `Robust SE` = std.error,
    `t-statistic` = statistic,
    `p-value` = p.value,
    `CI low` = conf.low,
    `CI high` = conf.high
  ) |>
  gt_compact(
    title = "Historical-average cross-sectional regression",
    subtitle = "48 industry portfolios; HC1 robust inference"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

average_pricing_table
```

Q: Interpret the coefficient on size.

A:

Q: What does a statistically insignificant price of risk mean?

A:

## Correlations and VIFs among estimated betas

Task: Examine correlations and VIFs among the estimated betas used as second-pass regressors.

```{r}
#| label: beta-collinearity

beta_correlations <- average_cross_section |>
  select(beta_mkt, beta_smb, beta_hml) |>
  cor()

beta_correlation_table <- beta_correlations |>
  as.data.frame() |>
  rownames_to_column("Beta") |>
  gt_compact(title = "Correlations among estimated industry betas") |>
  fmt_number(columns = where(is.numeric), decimals = 2)

beta_vif <- calculate_vif(
  average_cross_section,
  regressors = c("beta_mkt", "beta_smb", "beta_hml")
)

beta_vif_table <- beta_vif |>
  rename(
    Beta = variable,
    `Auxiliary R-squared` = auxiliary_r_squared,
    VIF = vif
  ) |>
  gt_compact(title = "Second-pass variance inflation factors") |>
  fmt_number(columns = c(`Auxiliary R-squared`, VIF), decimals = 2)

beta_correlation_table
beta_vif_table
```

Q: Why do we calculate a second set of correlations and VIFs?

A:

Q: Interpret the results.

A:

## Plot the cross-sectional regression

With three beta regressors, the fitted relation is a hyperplane rather than a line. We therefore plot fitted against actual average returns.

Task: Plot each industry's actual average excess return against its fitted value and label the most poorly priced industries.

```{r}
#| label: cross-sectional-plot

average_diagnostics <- broom::augment(
  average_pricing_model,
  data = average_cross_section
) |>
  mutate(
    absolute_pricing_error = ?,
    label = if_else(
      min_rank(desc(absolute_pricing_error)) <= 6,
      industry,
      NA_character_
    )
  )

# coord_equal() below scales the panel so that fitted returns
# (a narrow range) and actual returns (a wider range) share one axis
# unit; giving both scales the same limits keeps the panel from
# collapsing into a tall, narrow strip.
pricing_error_range <- range(
  c(average_diagnostics$.fitted, average_diagnostics$average_excess_return)
)

average_cross_section_plot <- ggplot(
  average_diagnostics,
  aes(.fitted, average_excess_return)
) +
  geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = 2) +
  geom_point(color = "#2166ac", size = 2) +
  geom_text_repel(aes(label = label), na.rm = TRUE, size = 3, max.overlaps = Inf) +
  scale_x_continuous(limits = pricing_error_range, labels = label_percent(accuracy = 0.01)) +
  scale_y_continuous(limits = pricing_error_range, labels = label_percent(accuracy = 0.01)) +
  labs(
    title = "Actual versus model-implied average excess returns",
    subtitle = "Labels identify the six largest absolute pricing errors",
    x = "Fitted average monthly excess return",
    y = "Actual average monthly excess return"
  ) +
  coord_equal()

average_cross_section_plot
```

Q: What does a point above the 45-degree line represent?

A:

# Step 2B: Month-by-month Fama-MacBeth regressions

## Estimate monthly prices of risk

Task: Attach the fixed first-pass betas to every industry-month and estimate one cross-sectional regression in each month.

```{r}
#| label: monthly-cross-sections

monthly_pricing_data <- panel |>
  select(date, industry, excess_return) |>
  left_join(betas, by = "industry")

monthly_pricing_models <- ? |>
  group_nest(date) |>
  mutate(
    model = purrr::map(
      data,
      ~ ?, data = .x)
    ),
    coefficients = purrr::map(model, broom::tidy),
    fit = purrr::map(model, broom::glance)
  )

lambda_time_series <- monthly_pricing_models |>
  select(date, coefficients) |>
  unnest(coefficients) |>
  select(date, term, estimate)
```

Task: For each coefficient, estimate its time-series mean and a Newey-West standard error for that mean.

```{r}
#| label: fama-macbeth-inference

fama_macbeth_summary <- lambda_time_series |>
  group_by(term) |>
  group_modify(~ {
    mean_model <- lm(estimate ~ ?, data = .x)
    nw_variance <- sandwich::?(
      mean_model,
      lag = settings$nw_lag,
      prewhite = FALSE,
      adjust = TRUE
    )
    mean_estimate <- unname(coef(mean_model)[1])
    standard_error <- sqrt(nw_variance[1, 1])
    t_statistic <- mean_estimate / standard_error

    tibble(
      months = nrow(.x),
      mean = mean_estimate,
      nw_se = standard_error,
      t_statistic = t_statistic,
      p_value = 2 * pt(-abs(t_statistic), df = nrow(.x) - 1),
      conf_low = mean_estimate - qt(0.975, df = nrow(.x) - 1) * standard_error,
      conf_high = mean_estimate + qt(0.975, df = nrow(.x) - 1) * standard_error
    )
  }) |>
  ungroup()

fama_macbeth_table <- fama_macbeth_summary |>
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    ),
    across(c(mean, nw_se, conf_low, conf_high), ~ 100 * .x)
  ) |>
  rename(
    Term = term,
    Months = months,
    `Mean monthly price (ppts)` = mean,
    `Newey-West SE` = nw_se,
    `t-statistic` = t_statistic,
    `p-value` = p_value,
    `CI low` = conf_low,
    `CI high` = conf_high
  ) |>
  gt_compact(
    title = "Fama-MacBeth estimates",
    subtitle = paste("Newey-West inference for mean monthly prices; lag =", settings$nw_lag)
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3)

fama_macbeth_table
```

Q: What is being tested in the Fama-MacBeth table? Interpret the results.

A:

## Compare the two second-pass approaches

Task: Compare coefficients from the historical-average cross section with the means of the monthly cross-sectional coefficients.

```{r}
#| label: compare-second-pass

second_pass_comparison <- average_pricing_inference |>
  select(term, historical_average_coefficient = estimate) |>
  left_join(
    fama_macbeth_summary |>
      select(term, mean_monthly_coefficient = mean),
    by = "term"
  ) |>
  mutate(
    difference = ?,
    across(c(historical_average_coefficient, mean_monthly_coefficient, difference), ~ 100 * .x),
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    )
  )

second_pass_comparison_table <- second_pass_comparison |>
  rename(
    Term = term,
    `Average-return regression (ppts)` = historical_average_coefficient,
    `Mean monthly coefficient (ppts)` = mean_monthly_coefficient,
    `Difference (ppts)` = difference
  ) |>
  gt_compact(title = "Two ways to obtain average prices of risk") |>
  fmt_number(columns = where(is.numeric), decimals = 8)

second_pass_comparison_table
```

Q: If the point estimates are effectively identical, why run the month-by-month regressions?

A:

## Plot prices of risk through time

Task: Plot each monthly price of risk and its trailing 12-month average.

```{r}
#| label: lambda-time-series-plot

lambda_plot_data <- lambda_time_series |>
  group_by(term) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    trailing_12_month_mean = zoo::rollmean(
      estimate,
      k = 12,
      fill = NA,
      align = "right"
    ),
    term = recode(
      term,
      `(Intercept)` = "Zero-beta return",
      beta_mkt = "Market price",
      beta_smb = "Size price",
      beta_hml = "Value price"
    )
  ) |>
  ungroup()

lambda_time_series_plot <- ggplot(lambda_plot_data, aes(date, estimate)) +
  geom_hline(yintercept = 0, color = "grey55") +
  geom_line(alpha = 0.22, linewidth = 0.35, color = "grey35") +
  geom_line(aes(y = trailing_12_month_mean), color = "#2166ac", linewidth = 0.7) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Monthly prices of risk",
    subtitle = "Grey: monthly estimate; blue: trailing 12-month mean",
    x = NULL, y = "Monthly price of risk"
  )

lambda_time_series_plot
```

Q: Why can a factor have a positive average price of risk even though its estimated monthly price is often negative?

A:

# Extension: Do prices of risk change in recessions?

We use the monthly NBER recession indicator distributed by FRED as `USREC`. For each estimated price of risk, we fit

$$
\widehat\lambda_{k,t}=\delta_{0,k}+\delta_{1,k}Recession_t+v_{k,t}.
$$

Q: How interpret the $\delta_{0,k}$ and $\delta_{1,k}$?

A:

Q: Why can we not add `recession` directly to each month's cross-sectional regression?

A:

Q: Interpret a positive and statistically significant recession coefficient for the market price of risk.

A:

# Synthesis

Q: Can the first-pass market beta be statistically significant for nearly every industry while the second-pass market price of risk is statistically insignificant?

A:

Q: What are the main limitations of this exercise?

A:

## Investment Implications

Q: How use these results to inform your investment decisions?

A:
