---
title: "Signal Evaluation Demo - Single Hold Out"
subtitle: "Evaluating the efficacy of a Signal over a single hold out period "
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

This file demonstrates how to use the function signal_evaluation_singleholdout, which is custom built for our class.

For this demo we will simulate some trading data, use signal_evaluation_singleholdout to evaluate, and then interpret the performance stats.

# Housekeeping

```{r}
#| echo: false
#| warning: false

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetUp.R"))
```

# Data

## Tickers

::: {.callout-caution collapse="false" title="Task"}
Task: Set a seed for consistent randomization
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
set.seed(123)
```

::: {.callout-caution collapse="false" title="Task"}
Task: Create a vector of 300 ticker labels of the form A001, A002, ..., A300
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
N <- 300
tickers <- paste0("A", sprintf("%03d", 1:N))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
tickers
```
:::

## Signal

::: {.callout-caution collapse="false" title="Task"}
Task: Generate a signal_matrix of length N from a N(0,1) distribution, representing the signal strength
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
signal_matrix_vec <- rnorm(N, mean = 0, sd = 1)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(signal_matrix_vec)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Create a signal_position vector based on the signal_matrix, assigning 1 for values above tau, -1 for values below -tau, and 0 otherwise. Let's set tau = 0.05 for now.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  tau <- 0.05
  signal_position_vec <- ifelse(signal_matrix_vec >  tau,  1,
                                ifelse(signal_matrix_vec < -tau, -1, 0))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(signal_position_vec)
```
:::

## Returns

::: {.callout-caution collapse="false" title="Task"}
Task: Generate test period returns from a linear model loosely related to signal strength. ret = 0.01\*signal + N(0,0.05) noise
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  eps <- rnorm(N, 0, 0.05)
  ret_simple <- 0.01 * signal_matrix_vec + eps
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(ret_simple)
```
:::

## Combine

::: {.callout-caution collapse="false" title="Task"}
Task: Format the signal_position and test_returns as data frames with ticker labels
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  signal_position <- data.frame(ticker = tickers, signal_position = signal_position_vec)
  test_returns <- data.frame(ticker = tickers, test_returns = ret_simple)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(test_returns)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Format the meta data for labels
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  Meta <- list(assetname = "asset", benchmarkname = "market", signalname = "Momentum")
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
Meta
```
:::

# Signal

When you are implementing in practice, you likely won't be simulating the returns and signals as we did above. Instead, you'd be using observed data to generate those objects.

Now that we have the data for the asset returns and signal generated, let's evalaute the signal.

## Call to Signal Evaluator

::: {.callout-caution collapse="false" title="Task"}
Task: Set the threshold for classification to .25%
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
return_threshold <- 0.0025
```

::: {.callout-tip title="Question"}
Q: What's the meaning of this return_threshold within our case?
:::

::: {.callout-note collapse="true" title="Answer"}
A: If test period returns \> return threshold, then we classify that as a "success". This will help form evaluation criterion like confusion matrices and hit rates.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Call the signal_evaluation_static function
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  SignalEval <- signal_evaluation_singleholdout(
    test_returns = test_returns,
    Meta = Meta,
    return_threshold = return_threshold,
    signal_position = signal_position
  )
```

## Signal Evaluation

In this subsection we will call selective components of the SignalEval object and interpret the efficacy of our signal.

::: {.callout-caution collapse="false" title="Task"}
Task: Display the TradeResults and their accompanying percent table.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false 
SignalEval$TradeResults
SignalEval$TradeResults_Percent
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$TradeResults
SignalEval$TradeResults_Percent
```
:::

::: {.callout-tip title="Question"}
Q: Interpret the first row of the Percent table
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- .257: In 25.7% of all observations we went long & there were positive test period returns
- 0.01: In 1% of all observations we didn't trade and there were positive test period returns
- 0.227: In 22.7% of all observations we went short and there were positive test period returns
- 0.493: In 49.3% of all observations where were positive test period returns
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the return comparison table
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false 
SignalEval$return_comparison
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$return_comparison
```
:::

::: {.callout-tip title="Question"}
Q: Interpret the Long vs Short results
:::

::: {.callout-note collapse="true" title="Answer"}
A: The difference in mean returns between long and short returns is 0.0086, indicating that on average, long positions outperformed short positions by 0.86%. The p-value of 0.0696 suggests that this difference is marginally statistically significant at the 10% level, indicating some evidence that the signal has predictive power over test period returns.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the violin plots of test returns by signal, outcome, and all combos
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$plots$signal
SignalEval$plots$facet
SignalEval$plots$combo
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$plots$signal
SignalEval$plots$facet
SignalEval$plots$combo
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display summary table of sample stats
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$summary_table
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$summary_table
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display Info Coefficient
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$summary_ic_tbl
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$summary_ic_tbl
```
:::

::: {.callout-tip title="Question"}
Q: What does the IC all mean? Is it statistically significant?
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- IC captures the correlation between the signal and test period returns
- The IC of 0.0847 has a p value of 0.1432, implying that we fail to reject the null that IC = 0, suggesting that the signal has little predictive power over test period returns
- Note that the results are similar if we focus only on the observations where we took a position (Acted Only)
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the Hit Rate
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$hit_table
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$hit_table
```
:::

::: {.callout-tip title="Question"}
Q: What does the Hit Rate mean? What does the p value and Coverage suggest?
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- The Hit Rate of .52 implies that for 52% of the observations our position correctly predicted the direction of test period returns.
- The p value of .356 suggests that we fail to reject the null that Hit Rate = 0.5, indicating that the signal has little predictive power over test period returns
- The coverage of .966 implies that we were active roughly 96% of the observations.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the confusion matrix
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$confusion_report
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$confusion_report
```
:::

::: {.callout-tip title="Question"}
Q: Interpret each of the metrics for Long positions
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- TP (True Positive): 77 observations we went long and test returns were positive
- FP (False Positive): 59 observations we went long but test returns were negative
- FN (False Negative): 71 observations we didn't go long but test returns were positive
- TN (True Negative): 79 observations we didn't go long and test returns were negative
- n (300): Total number of observations considered for long position classification
- Accuracy (.55): 55% of all observations were classified correctly
- Precision (0.55): 57% of observations where we went long had positive test returns
- Recall (.52): 52% of observations where test returns were positive we went long
- specificity (.58): 58% of observations where test returns were negative we didn't go long
- Type 1 Error (.42): 43% of observations where test returns were negative we went long
- Type 2 Error (.48): 48% of observations test returns were positive we didn't go long
- f1 (.53): The harmonic mean of precision and recall for long positions
- mcc (.10): Matthews correlation coefficient for long positions, measuring the quality of binary classifications
:::