---
title: "Signal Evaluation Demo - Single Hold Out"
subtitle: "Evaluating the efficacy of a Signal over a single hold out period "
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

This file demonstrates how to use the function signal_evaluation_singleholdout, which is custom built for our class.

For this demo we will simulate some trading data, use signal_evaluation_singleholdout to evaluate, and then interpret the performance stats.

# Housekeeping

```{r}
#| echo: false
#| warning: false

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetUp.R"))
```

# Data

## Tickers

::: {.callout-caution collapse="false" title="Task"}
Task: Set a seed for consistent randomization
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
set.seed(123)
```

::: {.callout-caution collapse="false" title="Task"}
Task: Create a vector of 300 ticker labels of the form A001, A002, ..., A300
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
N <- 300
tickers <- paste0("A", sprintf("%03d", 1:N))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
tickers
```
:::

## Signal

::: {.callout-caution collapse="false" title="Task"}
Task: Generate a signal_matrix of length N from a N(0,1) distribution, representing the signal strength
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
signal_matrix_vec <- rnorm(N, mean = 0, sd = 1)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(signal_matrix_vec)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Create a signal_position vector based on the signal_matrix, assigning 1 for values above tau, -1 for values below -tau, and 0 otherwise. Let's set tau = 0.05 for now.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  tau <- 0.05
  signal_position_vec <- ifelse(signal_matrix_vec >  tau,  1,
                                ifelse(signal_matrix_vec < -tau, -1, 0))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(signal_position_vec)
```
:::

## Returns

::: {.callout-caution collapse="false" title="Task"}
Task: Generate test period returns from a linear model loosely related to signal strength. ret = 0.01\*signal + N(0,0.05) noise
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  eps <- rnorm(N, 0, 0.05)
  ret_simple <- 0.01 * signal_matrix_vec + eps
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(ret_simple)
```
:::

## Combine

::: {.callout-caution collapse="false" title="Task"}
Task: Format the signal_position and test_returns as data frames with ticker labels
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  signal_position <- data.frame(ticker = tickers, signal_position = signal_position_vec)
  test_returns <- data.frame(ticker = tickers, test_returns = ret_simple)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(test_returns)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Format the meta data for labels
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  Meta <- list(assetname = "asset", benchmarkname = "market", signalname = "Momentum")
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
Meta
```
:::

# Signal

When you are implementing in practice, you likely won't be simulating the returns and signals as we did above. Instead, you'd be using observed data to generate those objects.

Now that we have the data for the asset returns and signal generated, let's evalaute the signal.

## Call to Signal Evaluator

::: {.callout-caution collapse="false" title="Task"}
Task: Set the threshold for classification to .25%
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
return_threshold <- 0.0025
```

::: {.callout-tip title="Question"}
Q: What's the meaning of this return_threshold within our case?
:::

::: {.callout-note collapse="true" title="Answer"}
A: Here, `return_threshold = 0.0025` creates a 0.25% neutral band around zero. The function classifies each realized test-period return as:

- **Positive:** return \> +0.25%
- **Neutral:** -0.25% $\leq$ return $\leq$ +0.25%
- **Negative:** return \< -0.25%

The threshold defines how large a return must be before we treat it as meaningfully positive or negative. For example, a long signal is a correct positive classification only when the subsequent return exceeds +0.25%; a return inside the neutral band is classified as neutral. The threshold is used in the return buckets and confusion-matrix analysis.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Call the signal_evaluation_static function
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
  SignalEval <- signal_evaluation_singleholdout(
    test_returns = test_returns,
    Meta = Meta,
    return_threshold = return_threshold,
    signal_position = signal_position
  )
```

## Signal Evaluation

In this subsection we will call selective components of the SignalEval object and interpret the efficacy of our signal.

::: {.callout-caution collapse="false" title="Task"}
Task: Display the TradeResults and their accompanying percent table.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false 
SignalEval$TradeResults
SignalEval$TradeResults_Percent
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$TradeResults
SignalEval$TradeResults_Percent
```
:::

::: {.callout-tip title="Question"}
Q: Interpret the first row of the Percent table
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- .257: In 25.7% of all observations we went long & there were positive test period returns
- 0.01: In 1% of all observations we didn't trade and there were positive test period returns
- 0.227: In 22.7% of all observations we went short and there were positive test period returns
- 0.493: In 49.3% of all observations where were positive test period returns
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the return comparison table
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false 
SignalEval$return_comparison
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$return_comparison
```
:::

::: {.callout-tip title="Question"}
Q: Interpret the Long vs Short results
:::

::: {.callout-note collapse="true" title="Answer"}
A: The difference in mean returns between long and short returns is 0.0086, indicating that on average, long positions outperformed short positions by 0.86%. The p-value of 0.0696 suggests that this difference is marginally statistically significant at the 10% level, indicating some evidence that the signal has predictive power over test period returns.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the violin plots of test returns by signal, outcome, and all combos
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$plots$signal
SignalEval$plots$facet
SignalEval$plots$combo
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$plots$signal
SignalEval$plots$facet
SignalEval$plots$combo
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display summary table of sample stats
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$summary_table
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$summary_table
```
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display Info Coefficient
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$summary_ic_tbl
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$summary_ic_tbl
```
:::

::: {.callout-tip title="Question"}
Q: What does the IC all mean? Is it statistically significant?
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- IC captures the correlation between the signal and test period returns
- The IC of 0.0847 has a p value of 0.1432, implying that we fail to reject the null that IC = 0, suggesting that the signal has little predictive power over test period returns
- Note that the results are similar if we focus only on the observations where we took a position (Acted Only)
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the Hit Rate
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$hit_table
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$hit_table
```
:::

::: {.callout-tip title="Question"}
Q: What does the Hit Rate mean? What does the p value and Coverage suggest?
:::

::: {.callout-note collapse="true" title="Answer"}
A:

- The Hit Rate of .52 implies that for 52% of the observations our position correctly predicted the direction of test period returns.
- The p value of .356 suggests that we fail to reject the null that Hit Rate = 0.5, indicating that the signal has little predictive power over test period returns
- The coverage of .966 implies that we were active roughly 96% of the observations.
:::

::: {.callout-caution collapse="false" title="Task"}
Task: Display the confusion matrix
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| eval: false
SignalEval$confusion_report
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
SignalEval$confusion_report
```
:::

::: {.callout-tip title="Question"}
Q: Interpret each of the metrics for Long positions
:::

::: {.callout-note collapse="true" title="Answer"}
A:

For this table, a **predicted positive** means that the signal told us to go long. An **actual positive** means that the test-period return exceeded +0.25%, while an **actual negative** means that it fell below -0.25%. A "not-long" signal is either neutral or short.

- **TP (True Positive = 77):** We went long, and the test-period return exceeded +0.25%.
- **FP (False Positive = 59):** We went long, but the test-period return fell below -0.25%.
- **FN (False Negative = 71):** We did not go long, but the test-period return exceeded +0.25%.
- **TN (True Negative = 79):** We did not go long, and the test-period return fell below -0.25%.
- **n (286):** The number of observations used in the long-versus-not-long classification: $77+59+71+79=286$. The other 14 of the original 300 observations had returns inside the neutral band and are excluded from this particular confusion matrix.
- **Accuracy (0.55):** Approximately 55% of the included observations were classified correctly: $(TP+TN)/n$.
- **Precision (0.57):** Among the observations for which we went long, 57% subsequently earned a return above +0.25%: $TP/(TP+FP)$.
- **Recall (0.52):** Among the observations whose returns exceeded +0.25%, we went long 52% of the time: $TP/(TP+FN)$.
- **Specificity (0.57):** Among the observations whose returns fell below -0.25%, we did not go long 57% of the time: $TN/(TN+FP)$.
- **Type I error (0.43):** Among the observations whose returns fell below -0.25%, we nevertheless went long 43% of the time: $FP/(FP+TN)=1-\text{specificity}$.
- **Type II error (0.48):** Among the observations whose returns exceeded +0.25%, we failed to go long 48% of the time: $FN/(FN+TP)=1-\text{recall}$.
- **F1 score (0.54):** The harmonic mean of precision and recall. It summarizes how well the long signal identifies large positive returns while limiting incorrect long signals; higher is better.
- **MCC (0.09):** The Matthews correlation coefficient summarizes all four confusion-matrix cells and ranges from -1 to +1. A value near zero indicates little classification skill; 0.09 therefore suggests only a weak positive relationship between the long signal and large positive test-period returns.
:::