---
title: "Signal Evaluation Demo - Single Hold Out"
subtitle: "Evaluating the efficacy of a Signal over a single hold out period "
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

This file demonstrates how to use the function signal_evaluation_singleholdout, which is custom built for our class.

For this demo we will simulate some trading data, use signal_evaluation_singleholdout to evaluate, and then interpret the performance stats.

# Housekeeping

```{r}
#| echo: false
#| warning: false

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetUp.R"))
```

# Data

## Tickers


Task: Set a seed for consistent randomization


```{r}

?(123)
```


Task: Create a vector of 300 ticker labels of the form A001, A002, ..., A300


```{r}

N <- ?
tickers <- paste0("A", sprintf("%03d", 1:?))
```

## Signal


Task: Generate a signal_matrix of length N from a N(0,1) distribution, representing the signal strength


```{r}

signal_matrix_vec <- ?(?, mean = ?, sd = ?)
```



Task: Create a signal_position vector based on the signal_matrix, assigning 1 for values above tau, -1 for values below -tau, and 0 otherwise. Let's set tau = 0.05 for now.


```{r}

  ? <- 0.05
  signal_position_vec <- ifelse(signal_matrix_vec >  ?,  ?,
                                ifelse(signal_matrix_vec < ?, ?, 0))
```



## Returns


Task: Generate test period returns from a linear model loosely related to signal strength. ret = 0.01\*signal + N(0,0.05) noise


```{r}

  eps <- ?
  ret_simple <- ?
```


## Combine


Task: Format the signal_position and test_returns as data frames with ticker labels


```{r}

  signal_position <- data.frame(ticker = tickers, signal_position = signal_position_vec)
  test_returns <- data.frame(ticker = tickers, test_returns = ret_simple)
```



Task: Format the meta data for labels


```{r}

  Meta <- list(assetname = "asset", benchmarkname = "market", signalname = "Momentum")
```



# Signal

When you are implementing in practice, you likely won't be simulating the returns and signals as we did above. Instead, you'd be using observed data to generate those objects.

Now that we have the data for the asset returns and signal generated, let's evalaute the signal.

## Call to Signal Evaluator


Task: Set the threshold for classification to .25%


```{r}

return_threshold <- ?
```


Q: What's the meaning of this return_threshold within our case?



A: ?



Task: Call the signal_evaluation_static function


```{r}

  SignalEval <- signal_evaluation_singleholdout(
    ?
  )
```

## Signal Evaluation

In this subsection we will call selective components of the SignalEval object and interpret the efficacy of our signal.


Task: Display the TradeResults and their accompanying percent table.


```{r}
 
SignalEval$?
SignalEval$?
```



Q: Interpret the first row of the Percent table



A:

?



Task: Display the return comparison table


```{r}
 
SignalEval$?
```



Q: Interpret the Long vs Short results



A: ?



Task: Display the violin plots of test returns by signal, outcome, and all combos


```{r}

SignalEval$plots$?
SignalEval$plots$?
SignalEval$plots$?
```



Task: Display summary table of sample stats


```{r}

SignalEval$?
```



Task: Display Info Coefficient


```{r}

SignalEval$?
```



Q: What does the IC all mean? Is it statistically significant?



A:

?



Task: Display the Hit Rate


```{r}

SignalEval$?
```



Q: What does the Hit Rate mean? What does the p value and Coverage suggest?



A:

?



Task: Display the confusion matrix


```{r}

SignalEval$?
```



Q: Interpret each of the metrics for Long positions



A:

?
