---
title: "PCA (Principal Component Analysis)"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
---

# Housekeeping

```{r}
#| warning: FALSE
#| code-fold: true
#| code-summary: "Code"

cat("\014")  # clear console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Background

- You are an investment analyst at a global asset-management firm.
- The firm allocates capital across country equity markets using country equity ETFs.
- You monitor many macroeconomic and equity-market indicators for every country.
- Your supervisor tells you to run a PCA.  

::: {.callout-tip title="Question"}
What are you "running a PCA on"? How could your supervisor use these results? 
:::


::: {.callout-note collapse="true" title="Answer"}
:::

For our case study: 

- We will focus on PCA as our dimension reduction technique.
- The data includes macroeconomic and equity-market indicators for a universe of fictional investable countries.

# Data

## Acquisition

::: {.callout-caution collapse="false" title="Task"}
Load the data dictionary from the Data_Dictionary sheet of country_equity_indicators.xlsx.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
data_dictionary <- read_excel(
  here("data", "country_equity_indicators.xlsx"),
  sheet = "Data_Dictionary"
)
  data_dictionary
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
data_dictionary
```
:::

::: {.callout-tip title="Question"}
Before doing any quantitative analysis, which variables do you suspect to be related to each other?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Load data file country_equity_indicators.xlsx from the Data sheet.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
country_data <- read_excel(
  here("data", "country_equity_indicators.xlsx"),
  sheet = "Data"
)
```

## Cleaning

::: {.callout-caution collapse="false" title="Task"}
Check for missing values
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
sum(is.na(country_data))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
sum(is.na(country_data))
```
:::

## Engineering

::: {.callout-caution collapse="false" title="Task"}
Standardize the data using "scale"
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
country_data_standardized <- scale(country_data %>% select(-country_code)) %>%
    as.data.frame() %>%
    mutate(country_code = country_data$country_code) %>%
    select(country_code, everything())
```

::: {.callout-caution collapse="false" title="Task"}
Convert country_data to a long dataframe. Then compute the mean and standard deviation of each variable.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
country_data%>%
    pivot_longer(-country_code, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(mean = mean(value), sd = sd(value))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
country_data%>%
    pivot_longer(-country_code, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(mean = mean(value), sd = sd(value))
```
:::

::: {.callout-caution collapse="false" title="Task"}
Convert country_data_standardized to a long dataframe. Then compute the mean and standard deviation of each variable.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
country_data_standardized%>%
    pivot_longer(-country_code, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(mean = mean(value), sd = sd(value))

```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
country_data_standardized%>%
    pivot_longer(-country_code, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    summarise(mean = mean(value), sd = sd(value))
```
:::

::: {.callout-tip title="Question"}
What do these summaries suggest?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## EDA

Let's focus on the comovement for this exercise.

### Compute correlation matrix

::: {.callout-caution collapse="false" title="Task"}
Compute the correlation matrix among the indicators
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
correlation_matrix <- cor(country_data_standardized %>% select(-country_code))
correlation_matrix
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
correlation_matrix
```
:::

::: {.callout-tip title="Question"}
Which indicators are candidates for consolidation?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

### Visualize the correlation matrix

::: {.callout-caution collapse="false" title="Task"}
Create a heatmap of the correlation matrix.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
melted_corr <- melt(correlation_matrix)

ggplot(melted_corr, aes(Var1, Var2, fill = value)) +
 geom_tile() +
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                      limit = c(-1, 1), name = "Correlation") +
 scale_y_discrete(limits = rev(levels(melted_corr$Var2))) +
 labs(title = "Correlation Matrix Heatmap", x = NULL, y = NULL) +
 theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
ggplot(melted_corr, aes(Var1, Var2, fill = value)) +
 geom_tile() +
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                      limit = c(-1, 1), name = "Correlation") +
 scale_y_discrete(limits = rev(levels(melted_corr$Var2))) +
 labs(title = "Correlation Matrix Heatmap", x = NULL, y = NULL) +
 theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

### Visualize pairwise plots

::: {.callout-caution collapse="false" title="Task"}
Generate pairwise plots for all indicators
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
ggpairs(country_data_standardized %>% select(-country_code))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
ggpairs(country_data_standardized %>% select(-country_code))
```
:::

::: {.callout-tip title="Question"}
What does the pairs plot suggest?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

# Implementing PCA

::: {.callout-tip title="Question"}
What is our analytical goal for this case study? How might PCA help?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Run PCA on the raw indicators using prcomp
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
pca_raw <- ?, scale. = FALSE)
```

::: {.callout-caution collapse="false" title="Task"}
Run the summary command on the PCA object.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?(?)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?(?)
```
:::

::: {.callout-tip title="Question"}
Can you interpret the output?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-tip title="Question"}
Why are 11 PCs listed?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

# Why is standardizing important?

::: {.callout-caution collapse="false" title="Task"}
Run pca on the standardized data
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
pca_z <- ? scale. = FALSE)
```

::: {.callout-caution collapse="false" title="Task"}
Summarize that pca with the summary command
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
summary(pca_z)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
summary(pca_z)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Re-estimate on country_data (not standardized data), but set scale. = TRUE
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
pca_z2 <- ?
summary(pca_z2)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
pca_z2 <- ?
summary(pca_z2)
```
:::

::: {.callout-tip title="Question"}
Coding implication?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Compare the PC1 loadings for the country indicators using raw vs. standardized data. 
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
country_loadings <- bind_rows(
  tibble(variable = names(pca_raw$rotation[, 1]), loading = pca_raw$rotation[, 1], version = "Raw"),
  tibble(variable = names(pca_z$rotation[, 1]), loading = pca_z$rotation[, 1], version = "Standardized")
)

ggplot(country_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~version) +
  labs(
    title = "Country indicators: PC1 loadings, raw vs. standardized",
    x = NULL, y = "PC1 loading"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
ggplot(country_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~version) +
  labs(
    title = "Country indicators: PC1 loadings, raw vs. standardized",
    x = NULL, y = "PC1 loading"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

::: {.callout-tip title="Question"}
Which indicators dominate the raw PC1? Do they still dominate once standardized?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

# Choosing the number of components

::: {.callout-tip title="Question"}
Our goal is to reduce the dimension of the data. Based on the summary information above, how many principal components do you think we should use?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Construct Scree Plot plotting "plot" pca_z.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
Interpret?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

Note that the PCA method transforms the covariance matrix of the data by rescaling into principal component vectors, which explain the maximum variation in the data. We solve $\Sigma v_{i}=\lambda_{i}v_{i}$ for each PC $(i)$, where $\lambda_{i}$ is the $i^{th}$ eigenvalue and $v_{i}$ is the $i^{th}$ eigenvector aka principal component.

::: {.callout-caution collapse="false" title="Task"}
Impose the Kaiser Criterion; i.e. keep all PCs with eigenvalues \> 1.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
eigenvalues <- ?
kaiser <- data.frame(
  Component = paste0("PC", seq_along(eigenvalues)),
  Eigenvalue = eigenvalues,
  Retain = eigenvalues > ?
)

kaiser
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
kaiser
```
:::

::: {.callout-tip title="Question"}
How many PCs are suggested?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-tip title="Question"}
How many PCs are needed if want to explain 60% of variance? 75%? 90%?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

# Interpreting PCA

## Loadings

::: {.callout-caution collapse="false" title="Task"}
Plot the loadings for PC1, PC2, PC3.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
pc1_order <- names(sort(abs(pca_z$rotation[, 1]), decreasing = TRUE))

pc_loadings <- as_tibble(pca_z$rotation[, 1:3], rownames = "variable") %>%
  pivot_longer(-variable, names_to = "PC", values_to = "loading") %>%
  mutate(
    PC = factor(PC, levels = c("PC1", "PC2", "PC3")),
    # Order indicators by |PC1 loading| so all facets share the same x-axis order
    variable = factor(variable, levels = pc1_order)
  )

ggplot(pc_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~PC, ncol = 1) +
  labs(title = "PCA loadings for PC1, PC2, PC3", x = NULL, y = "Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
ggplot(pc_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~PC, ncol = 1) +
  labs(title = "PCA loadings for PC1, PC2, PC3", x = NULL, y = "Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

::: {.callout-tip title="Question"}
Interpret the loading vector values.
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## Scores

::: {.callout-tip title="Question"}
What is a PCA score? Contrast with a loading.
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Extract the PC1 score for the first country from the pca_z object
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
pca_z$x[1, 1]
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
pca_z$x[1, 1]
```
:::

::: {.callout-caution collapse="false" title="Task"}
Recreate that PC1 score as a weighted sum of the loadings.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
ctry001 <- country_data_standardized %>%
  filter(country_code == "CTRY001") %>%
  select(-country_code) %>%
  unlist()

pc1_loadings <- pca_z$rotation[, 1]

?
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
How do we interpret the score in our context?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Extract the PC1 score for the second, third, and fourth countries
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
What's the interpretation of the sign and magnitude of these PC1 scores relative to CTRY001?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## Reconstruction

The goal of dimension reduction is to represent the original data using fewer dimensions while retaining as much information as possible.

This creates a tradeoff. If we retain all principal components, we lose no information, but we achieve no dimension reduction. If we retain only a few principal components, we obtain a simpler representation but lose some information about the original data.

We can quantify this tradeoff by asking: How accurately can the retained PCs reconstruct the original data?

::: {.callout-tip title="Question"}
How reconstruct the data from the PCA?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Create a function that reconstructs the data via scores times loadings.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
original <- country_data_standardized %>% select(-country_code) %>% as.matrix()

reconstruct <- function(pca, k) {
  scores <- pca$x[, 1:k, drop = FALSE]
  loadings <- pca$rotation[, 1:k, drop = FALSE]
  recon <- scores %*% t(loadings)
  sweep(recon, 2, pca$center, "+")
}
```

::: {.callout-caution collapse="false" title="Task"}
Reconstruct the data using only the first principal component.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
reconstruction_1pc <- reconstruct(?
head(reconstruction_1pc)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(reconstruction_1pc)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the mean squared reconstruction error
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
error_1pc <- ?
error_1pc
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
error_1pc
```
:::

::: {.callout-caution collapse="false" title="Task"}
Reconstruct the data using only the first and second principal components.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
reconstruction_2pc <- reconstruct(?
head(reconstruction_2pc)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(reconstruction_2pc)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the mean squared reconstruction error
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
error_2pc <- ?
error_2pc
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
error_2pc
```
:::

::: {.callout-caution collapse="false" title="Task"}
Reconstruct the data using only the first, second, and third principal components.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
reconstruction_3pc <- reconstruct(?
head(reconstruction_3pc)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(reconstruction_3pc)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the mean squared reconstruction error
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
error_3pc <- ?
error_3pc
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
error_3pc
```
:::

::: {.callout-caution collapse="false" title="Task"}
Create a tibble summarizing the reconstruction error.
:::

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
tibble(
  `# PCs used` = 1:3,
  `Reconstruction error (MSE)` = c(error_1pc, error_2pc, error_3pc)
)
```
:::

::: {.callout-tip title="Question"}
Why not always use all PCs?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## BiPlot

::: {.callout-tip title="Question"}
How read a biplot for PCA analysis?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the Biplot between PC1 and PC2
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?(?, choices = c(1, 2), cex = c(0.6, 0.8))
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
Interpret.
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the Biplot between PC1 and PC3
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
Interpret.
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Compute the Biplot between PC2 and PC3
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
?
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
?
```
:::

::: {.callout-tip title="Question"}
Interpret.
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## Investment interpretation

::: {.callout-tip title="Question"}
What is the interpretation of each PC?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-tip title="Question"}
Which direction of PC2 corresponds to stronger growth and momentum?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Rank the countries by growth and momentum strength.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
growth_momentum <- tibble(
  country_code = country_data_standardized$country_code,
  PC2 = pca_z$x[, 2],
  growth_momentum_score = pca_z$x[, 2]
) %>%
  arrange(desc(growth_momentum_score)) %>%
  mutate(rank = row_number())

growth_momentum
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
growth_momentum
```
:::

::: {.callout-tip title="Question"}
Which countries have the strongest growth and momentum?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Create a table that compares CTRY042 to the average country along all of the growth and momentum indicators.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
growth_validation_raw <- country_data %>%
  select(country_code, c(GDP_Growth_Rate, Industrial_Production_Growth, Earnings_Growth_Rate, Equity_Momentum_12M)) %>%
  pivot_longer(
    cols = -country_code,
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Variable) %>%
  summarise(
    Average_Country = mean(Value, na.rm = TRUE),
    CTRY042 = Value[country_code == "CTRY042"]
  )

growth_validation_raw
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
growth_validation_raw
```
:::

::: {.callout-tip title="Question"}
Does the underlying data support our interpretation of CTRY042's PC2 score?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-tip title="Question"}
Which direction of PC1 corresponds to higher risk?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

::: {.callout-caution collapse="false" title="Task"}
Rank the countries by highest risk.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
risk_ranking <- tibble(
  country_code = country_data_standardized$country_code,
  PC1 = pca_z$x[, 1],
  risk_score = pca_z$x[, 1]
) %>%
  arrange(desc(risk_score)) %>%
  mutate(rank = row_number())
head(risk_ranking, 10)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
head(risk_ranking, 10)
```
:::

::: {.callout-tip title="Question"}
Which countries have the highest risk according to PC1?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

# Robustness Checks

## Sensitivity to outliers

::: {.callout-caution collapse="false" title="Task"}
Remove CTRY022 and CTRY042. Rerun PCA.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
country_data_trimmed <- country_data_standardized %>%
  filter(!country_code %in% c("CTRY022", "CTRY042"))

pca_trimmed <- prcomp(country_data_trimmed %>% select(-country_code), scale. = FALSE)
```

::: {.callout-caution collapse="false" title="Task"}
Compare the summary of the first 3 PCs for the new trimmed set and the original standardized data.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
summary(pca_z)$importance[, 1:3]
summary(pca_trimmed)$importance[, 1:3]
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
summary(pca_z)$importance[, 1:3]
summary(pca_trimmed)$importance[, 1:3]
```
:::

::: {.callout-tip title="Question"}
Does removing these two outlier countries change the PCA results substantially?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## Correlated Variable

Let's examine the impact upon PCA of highly correlated variables.

::: {.callout-caution collapse="false" title="Task"}
Duplicate the GDP growth indicator.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
country_data_extended <- country_data_standardized %>%
  mutate(GDP_Growth_Rate_dup = GDP_Growth_Rate)
```

::: {.callout-caution collapse="false" title="Task"}
Rerun PCA on an extended set that includes this duplicated indicator.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
pca_extended <- prcomp(country_data_extended %>% select(-country_code), scale. = FALSE)
```

::: {.callout-caution collapse="false" title="Task"}
Summarize the PCA results for the original and extended dataset across the first 3 PCs
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
summary(pca_z)$importance[, 1:3]
summary(pca_extended)$importance[, 1:3]
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
summary(pca_z)$importance[, 1:3]
summary(pca_extended)$importance[, 1:3]
```
:::

::: {.callout-caution collapse="false" title="Task"}
Display the loadings for the original and extended dataset of the first 3 PCs.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
round(pca_z$rotation[, 1:3], 3)
round(pca_extended$rotation[, 1:3], 3)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
round(pca_z$rotation[, 1:3], 3)
round(pca_extended$rotation[, 1:3], 3)
```
:::

::: {.callout-caution collapse="false" title="Task"}
Display the scores for the original and extended dataset for the first two PCs.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
tibble(
  country_code = country_data_standardized$country_code,
  PC1_original = pca_z$x[, 1],
  PC1_extended = pca_extended$x[, 1],
  PC2_original = pca_z$x[, 2],
  PC2_extended = pca_extended$x[, 2]
) %>%
  head(10)
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
tibble(
  country_code = country_data_standardized$country_code,
  PC1_original = pca_z$x[, 1],
  PC1_extended = pca_extended$x[, 1],
  PC2_original = pca_z$x[, 2],
  PC2_extended = pca_extended$x[, 2]
) %>%
  head(10)
```
:::

::: {.callout-tip title="Question"}
What's the impact on the variance explained, loading vectors, and scores?
:::

::: {.callout-note collapse="true" title="Answer"}
:::

## Noise

Let's examine how measurement noise in one variable affects PCA. We will add progressively more independent noise to Equity_Volatility and examine how its role in the principal components changes.

::: {.callout-caution collapse="false" title="Task"}
Add increasing noise to Equity_Volatility, re-standardize the data, and record its correlation with the Sovereign Risk Spread and its loading on the risk component. Because PCA signs can flip and components can reorder across repeated runs, identify the risk component and align its sign against the original PCA before recording the loading.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
loadings_orig <- pca_z$rotation[, 1:3]

# The risk block defines one of the first three components. Identify which
# one by the block that loads most strongly on it, rather than assuming a
# fixed index (adding noise can reorder the components).
risk_block <- c(
  "Equity_Volatility", "FX_Volatility",
  "Inflation_Rate", "Sovereign_Risk_Spread_Bps"
)
risk_pc <- which.max(colMeans(abs(loadings_orig[risk_block, ])))
risk_ref <- loadings_orig[, risk_pc]

# Given a noisy rotation matrix, find the component (among the first 3) most
# aligned with the original risk direction, flip its sign to match, and return
# the aligned loading vector.
match_risk_component <- function(rotation, ref, k = 3) {
  dots <- sapply(seq_len(k), function(j) sum(ref * rotation[, j]))
  j <- which.max(abs(dots))
  v <- rotation[, j]
  if (dots[j] < 0) v <- -v
  v
}

noise_effect <- function(sigma2) {

  noisy_data <- country_data_standardized %>%
    mutate(
      Equity_Volatility =
        Equity_Volatility +
        rnorm(n(), mean = 0, sd = sqrt(sigma2))
    )

  # Re-standardize after adding noise
  x <- noisy_data %>%
    select(-country_code) %>%
    scale()

  pca_noisy <- prcomp(x, center = FALSE, scale. = FALSE)

  # Align the risk component with the original PCA, then read off the loading
  risk_loading <- match_risk_component(
    pca_noisy$rotation,
    risk_ref
  )["Equity_Volatility"]

  tibble(
    correlation = cor(
      x[, "Equity_Volatility"],
      x[, "Sovereign_Risk_Spread_Bps"]
    ),
    risk_loading = risk_loading
  )
}
```

::: {.callout-caution collapse="false" title="Task"}
Repeat across increasing noise levels.
:::

```{r}
#| code-fold: true
#| code-summary: "Code"
#| output: false
sigma2_grid <- seq(0, 1, by = 0.1)
n_reps <- 100

set.seed(7194)

noise_results <- map_dfr(sigma2_grid, function(s2) {

  map_dfr(1:n_reps, function(i) {
    noise_effect(s2)
  }) %>%
    summarize(
      correlation = mean(correlation),
      risk_loading = mean(risk_loading)
    ) %>%
    mutate(sigma2 = s2)
})
noise_results
```

::: {.callout-note collapse="true" title="Output"}
```{r}
#| echo: false
noise_results
```
:::

::: {.callout-tip title="Question"}
What happens as we add more noise?
:::

::: {.callout-note collapse="true" title="Answer"}
:::
