---
title: "PCA (Principal Component Analysis)"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
---

# Housekeeping

```{r}
#| warning: FALSE

cat("\014")  # clear console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Background

- You are an investment analyst at a global asset-management firm.
- The firm allocates capital across country equity markets using country equity ETFs.
- You monitor many macroeconomic and equity-market indicators for every country.
- The investment team wants a small number of measures summarizing each country's conditions.
- Can we reduce many correlated indicators into only a few informative dimensions?
- We will focus on PCA as our dimension reduction technique.
- The resulting component scores help us compare countries, screen markets, and monitor exposures. PCA is not itself a trading strategy: a high score is not inherently good or bad, and the components do not mechanically prescribe which ETFs to buy.
- The data includes macroeconomic and equity-market indicators for a universe of fictional investable countries.

# Data

## Acquisition

Task: Load the data dictionary from the Data_Dictionary sheet of country_equity_indicators.xlsx.

```{r}
data_dictionary <- read_excel(
  here("data", "?"),
  sheet = "Data_Dictionary"
)
  data_dictionary
```

Q: Before doing any quantitative analysis, which variables do you suspect to be related to each other?

A:

Task: Load data file country_equity_indicators.xlsx from the Data sheet.

```{r}
country_data <- read_excel(
  here("data", "country_equity_indicators.xlsx"),
  sheet = "?"
)
```

## Cleaning

Task: Check for missing values

```{r}
sum(is.na(country_data))
```

## Engineering

Task: Standardize the data using "scale"

```{r}
country_data_standardized <- ?(country_data %>% select(-country_code)) %>%
    as.data.frame() %>%
    mutate(country_code = country_data$country_code) %>%
    select(country_code, everything())
```

Task: Convert country_data to a long dataframe. Then compute the mean and standard deviation of each variable.

```{r}
country_data%>%
    ?(-country_code, names_to = "variable", values_to = "value") %>%
    ?(variable) %>%
    summarise(mean = mean(value), sd = sd(value))
```

Task: Convert country_data_standardized to a long dataframe. Then compute the mean and standard deviation of each variable.

```{r}
country_data_standardized%>%
    pivot_longer(-country_code, names_to = "variable", values_to = "value") %>%
    group_by(variable) %>%
    ?(mean = mean(value), sd = sd(value))

```

Q: What do these summaries suggest?

A:

## EDA

Let's focus on the comovement for this exercise.

### Compute correlation matrix

Task: Compute the correlation matrix among the indicators

```{r}
correlation_matrix <- ?(country_data_standardized %>% select(-country_code))
correlation_matrix
```

Q: Which indicators are candidates for consolidation?

A:

### Visualize the correlation matrix

Task: Create a heatmap of the correlation matrix.

```{r}
melted_corr <- melt(correlation_matrix)

ggplot(melted_corr, aes(Var1, Var2, fill = value)) +
 geom_tile() +
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, 
                      limit = c(-1, 1), name = "Correlation") +
 scale_y_discrete(limits = rev(levels(melted_corr$Var2))) +
 labs(title = "Correlation Matrix Heatmap", x = NULL, y = NULL) +
 theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Visualize pairwise plots

Task: Generate pairwise plots for all indicators

```{r}
ggpairs(country_data_standardized %>% select(?country_code))
```

Q: What does the pairs plot suggest?

A:

# Implementing PCA

Q: What is our analytical goal for this case study? How might PCA help?

A:

Task: Run PCA on the raw indicators using prcomp

```{r}
pca_raw <- prcomp(country_data %>% select(-country_code), scale. = FALSE)
```

Task: Run the summary command on the PCA object.

```{r}
summary(pca_raw)
```

Q: Can you interpret the output?

A:

Q: Why are 11 PCs listed?

A:

# Why is standardizing important?

Task: Run pca on the standardized data

```{r}
pca_z <- ?(country_data_standardized %>% select(-country_code), scale. = FALSE)
```

Task: Summarize that pca with the summary command

```{r}
summary(?)
```

Task: Re-estimate on country_data (not standardized data), but set scale. = TRUE

```{r}
pca_z2 <- prcomp(country_data%>% select(-country_code), scale. = ?)
summary(pca_z2)
```

Q: Coding implication?

A:

Task: Compare the PC1 loadings for the country indicators using raw vs. standardized data, the same way we did for the toy example.

```{r}
country_loadings <- bind_rows(
  tibble(variable = names(pca_raw$rotation[, 1]), loading = pca_raw$rotation[, 1], version = "Raw"),
  tibble(variable = names(pca_z$rotation[, 1]), loading = pca_z$rotation[, 1], version = "Standardized")
)

ggplot(country_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~version) +
  labs(
    title = "Country indicators: PC1 loadings, raw vs. standardized",
    x = NULL, y = "PC1 loading"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Q: Which indicators dominate the raw PC1? Do they still dominate once standardized?

A:

# Choosing the number of components

Q: Our goal is to reduce the dimension of the data. Based on the summary information above, how many principal components do you think we should use?

A:

Task: Construct Scree Plot plotting "plot" pca_z.

```{r}
plot(pca_z, type = "line")
```

Q: Interpret?

A:

Note that the PCA method transforms the covariance matrix of the data by rescaling into principal component vectors, which explain the maximum variation in the data. We solve $\Sigma v_{i}=\lambda_{i}v_{i}$ for each PC $(i)$, where $\lambda_{i}$ is the $i^{th}$ eigenvalue and $v_{i}$ is the $i^{th}$ eigenvector aka principal component.

Task: Impose the Kaiser Criterion; i.e. keep all PCs with eigenvalues \> 1.

```{r}
eigenvalues <- pca_z$sdev^2
kaiser <- data.frame(
  Component = paste0("PC", seq_along(eigenvalues)),
  Eigenvalue = eigenvalues,
  Retain = eigenvalues > 1
)

kaiser
```

Q: How many PCs are suggested?

A:

Q: How many PCs are needed if want to explain 60% of variance? 75%? 90%?

A:

```{r}
?
```

# Interpreting PCA

## Loadings

Task: Plot the loadings for PC1, PC2, PC3.

```{r}
pc1_order <- names(sort(abs(pca_z$rotation[, 1]), decreasing = TRUE))

pc_loadings <- as_tibble(pca_z$rotation[, 1:3], rownames = "variable") %>%
  pivot_longer(-variable, names_to = "PC", values_to = "loading") %>%
  mutate(
    PC = factor(PC, levels = c("PC1", "PC2", "PC3")),
    # Order indicators by |PC1 loading| so all facets share the same x-axis order
    variable = factor(variable, levels = pc1_order)
  )

ggplot(pc_loadings, aes(variable, loading)) +
  geom_col() +
  facet_wrap(~PC, ncol = 1) +
  labs(title = "PCA loadings for PC1, PC2, PC3", x = NULL, y = "Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Q: Interpret the loading vector values.

A:

## Scores

Q: What is a PCA score? Contrast with a loading.

A:

Task: Extract the PC1 score for the first country from the pca_z object

```{r}
pca_z?
```

Task: Recreate that PC1 score as a weighted sum of the loadings.

```{r}
ctry001 <- country_data_standardized %>%
  ?(country_code == "CTRY001") %>%
  select(-country_code) %>%
  unlist()

pc1_loadings <- pca_z$rotation[, 1]

sum(ctry001 * pc1_loadings)
```

Q: How do we interpret the score in our context?

A:

Task: Extract the PC1 score for the second, third, and fourth countries

```{r}
pca_z$x[2:4, 1]
```

Q: What's the interpretation of the sign and magnitude of these PC1 scores relative to CTRY001?

A:

## Reconstruction

The goal of dimension reduction is to represent the original data using fewer dimensions while retaining as much information as possible.

This creates a tradeoff. If we retain all principal components, we lose no information, but we achieve no dimension reduction. If we retain only a few principal components, we obtain a simpler representation but lose some information about the original data.

We can quantify this tradeoff by asking: How accurately can the retained PCs reconstruct the original data?

Q: How reconstruct the data from the PCA?

A:

Task: Create a function that reconstructs the data via scores times loadings.

```{r}
original <- country_data_standardized %>% select(-country_code) %>% as.matrix()

reconstruct <- function(pca, k) {
  scores <- pca$x[, 1:k, drop = FALSE]
  loadings <- pca$rotation[, 1:k, drop = FALSE]
  recon <- scores %*% t(loadings)
  sweep(recon, 2, pca$center, "+")
}
```

Task: Reconstruct the data using only the first principal component.

```{r}
reconstruction_1pc <- reconstruct(pca_z, 1)
head(reconstruction_1pc)
```

Task: Compute the mean squared reconstruction error

```{r}
error_1pc <- ?
error_1pc
```

Task: Reconstruct the data using only the first and second principal components.

```{r}
reconstruction_2pc <- reconstruct(pca_z, 2)
head(reconstruction_2pc)
```

Task: Compute the mean squared reconstruction error

```{r}
error_2pc <- ?
error_2pc
```

Task: Reconstruct the data using only the first, second, and third principal components.

```{r}
reconstruction_3pc <- reconstruct(pca_z, 3)
head(reconstruction_3pc)
```

Task: Compute the mean squared reconstruction error

```{r}
error_3pc <- ?
error_3pc
```

```{r}
tibble(
  `# PCs used` = 1:3,
  `Reconstruction error (MSE)` = c(error_1pc, error_2pc, error_3pc)
)
```

Q: Why not always use all PCs?

A:

## BiPlot

Q: How read a biplot for PCA analysis?

A:

Task: Compute the Biplot between PC1 and PC2

```{r}
?(pca_z, choices = c(1, 2), cex = c(0.6, 0.8))
```

Q: Interpret.

A:

Task: Compute the Biplot between PC1 and PC3

```{r}
?(pca_z, choices = c(1, 3), cex = c(0.6, 0.8))
```

Q: Interpret.

A:

Task: Compute the Biplot between PC2 and PC3

```{r}
?(pca_z, choices = c(2, 3), cex = c(0.6, 0.8))
```

Q: Interpret.

A:

## Investment interpretation

Q: What is the interpretation of each PC?

A:

Q: Which direction of PC2 corresponds to stronger growth and momentum?

A:

Task: Rank the countries by growth and momentum strength.

```{r}
growth_momentum <- tibble(
  country_code = country_data_standardized$?,
  PC2 = pca_z$x[, 2],
  growth_momentum_score = pca_z$x[, 2]
) %>%
  arrange(desc(growth_momentum_score)) %>%
  mutate(rank = row_number())

growth_momentum
```

Q: Which countries have the strongest growth and momentum?

A:

Task: Create a table that compares CTRY042 to the average country along all of the growth and momentum indicators.

```{r}
growth_validation_raw <- country_data %>%
  select(country_code, c(GDP_Growth_Rate, Industrial_Production_Growth, Earnings_Growth_Rate, Equity_Momentum_12M)) %>%
  pivot_longer(
    cols = -country_code,
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Variable) %>%
  summarise(
    Average_Country = mean(Value, na.rm = TRUE),
    CTRY042 = Value[country_code == "CTRY042"]
  )

growth_validation_raw
```

Q: Does the underlying data support our interpretation of CTRY042's PC2 score?

A:

Q: Which direction of PC1 corresponds to higher risk?

A:

Task: Rank the countries by highest risk.

```{r}
risk_ranking <- tibble(
  country_code = country_data_standardized$country_code,
  PC1 = pca_z$x[, 1],
  risk_score = pca_z$x[, 1]
) %>%
  arrange(desc(risk_score)) %>%
  mutate(rank = row_number())
head(risk_ranking, 10)
```

Q: Which countries have the highest risk according to PC1?

A:

# Robustness Checks

## Sensitivity to outliers

Task: Remove CTRY022 and CTRY042. Rerun PCA.

```{r}
country_data_trimmed <- country_data_standardized %>%
  filter(!country_code %in% c("CTRY022", "CTRY042"))

pca_trimmed <- prcomp(country_data_trimmed %>% select(-country_code), scale. = FALSE)
```

Task: Compare the summary of the first 3 PCs for the new trimmed set and the original standardized data.

```{r}
summary(pca_z)$importance[, 1:3]
summary(pca_trimmed)$importance[, 1:3]
```

Q: Does removing these two outlier countries change the PCA results substantially?

A:

## Correlated Variable

Let's examine the impact upon PCA of highly correlated variables.

Task: Duplicate the GDP growth indicator.

```{r}
country_data_extended <- country_data_standardized %>%
  mutate(GDP_Growth_Rate_dup = GDP_Growth_Rate)
```

Task: Rerun PCA on an extended set that includes this duplicated indicator.

```{r}
pca_extended <- ?(country_data_extended %>% select(-country_code), scale. = FALSE)
```

Task: Summarize the PCA results for the original and extended dataset across the first 3 PCs

```{r}
summary(pca_z)$importance[, 1:3]
summary(pca_extended)$importance[, 1:3]
```

Task: Display the loadings for the original and extended dataset of the first 3 PCs.

```{r}
round(pca_z$rotation[, 1:3], 3)
round(pca_extended$rotation[, 1:3], 3)
```

Task: Display the scores for the original and extended dataset for the first two PCs.

```{r}
tibble(
  country_code = country_data_standardized$country_code,
  PC1_original = pca_z$x[, 1],
  PC1_extended = pca_extended$x[, 1],
  PC2_original = pca_z$x[, 2],
  PC2_extended = pca_extended$x[, 2]
) %>%
  head(10)
```

Q: What's the impact on the variance explained, loading vectors, and scores?

A:

## Noise

Let's examine how measurement noise in one variable affects PCA. We will add progressively more independent noise to Equity_Volatility and examine how its role in the principal components changes.

Task: Add increasing noise to Equity_Volatility, re-standardize the data, and record its correlation with the Sovereign Risk Spread and its loading on the risk component. Because PCA signs can flip and components can reorder across repeated runs, identify the risk component and align its sign against the original PCA before recording the loading.

```{r}
loadings_orig <- pca_z$rotation[, 1:3]

# The risk block defines one of the first three components. Identify which
# one by the block that loads most strongly on it, rather than assuming a
# fixed index (adding noise can reorder the components).
risk_block <- c(
  "Equity_Volatility", "FX_Volatility",
  "Inflation_Rate", "Sovereign_Risk_Spread_Bps"
)
risk_pc <- which.max(colMeans(abs(loadings_orig[risk_block, ])))
risk_ref <- loadings_orig[, risk_pc]

# Given a noisy rotation matrix, find the component (among the first 3) most
# aligned with the original risk direction, flip its sign to match, and return
# the aligned loading vector.
match_risk_component <- function(rotation, ref, k = 3) {
  dots <- sapply(seq_len(k), function(j) sum(ref * rotation[, j]))
  j <- which.max(abs(dots))
  v <- rotation[, j]
  if (dots[j] < 0) v <- -v
  v
}

noise_effect <- function(sigma2) {

  noisy_data <- country_data_standardized %>%
    mutate(
      Equity_Volatility =
        Equity_Volatility +
        rnorm(n(), mean = 0, sd = sqrt(sigma2))
    )

  # Re-standardize after adding noise
  x <- noisy_data %>%
    select(-country_code) %>%
    scale()

  pca_noisy <- prcomp(x, center = FALSE, scale. = FALSE)

  # Align the risk component with the original PCA, then read off the loading
  risk_loading <- match_risk_component(
    pca_noisy$rotation,
    risk_ref
  )["Equity_Volatility"]

  tibble(
    correlation = cor(
      x[, "Equity_Volatility"],
      x[, "Sovereign_Risk_Spread_Bps"]
    ),
    risk_loading = risk_loading
  )
}
```

Task: Repeat across increasing noise levels.

```{r}
sigma2_grid <- seq(0, 1, by = 0.1)
n_reps <- 100

set.seed(7194)

noise_results <- map_dfr(sigma2_grid, function(s2) {

  map_dfr(1:n_reps, function(i) {
    noise_effect(s2)
  }) %>%
    summarize(
      correlation = mean(correlation),
      risk_loading = mean(risk_loading)
    ) %>%
    mutate(sigma2 = s2)
})
noise_results
```

Q: What happens as we add more noise?

A:

-