---
title: "K-Means"
subtitle: "Machine Learning For Finance"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
---

# Learning Objectives

By the end of this lesson, students should be able to:

- Explain how K-means assigns observations to clusters and updates cluster centroids.
- Prepare a cross-sectional feature set for clustering, including checks for missing values, duplicate keys, unstable growth rates, scale differences, and influential observations.
- Explain why feature selection, standardization, random starting values, and the choice of $K$ can change the resulting partition.
- Use the elbow, gap-statistic, and silhouette approaches as complementary evidence when choosing $K$.
- Interpret clusters economically and describe how an investor might use—and validate—the resulting segmentation.

# Housekeeping

```{r}
#| warning: false

cat("\014")  # clear the console when run interactively
rm(list = ls())  # begin from a clean workspace

if (!requireNamespace("here", quietly = TRUE)) {
  stop("Install the 'here' package before rendering this document.")
}

suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Background

- You are an analyst at a systematic investment manager
- The asset manager is focused on fundamentals and operates primarily in ETFs
- Your portfolio manager is asking you to cluster industries by their revenue and cost characteristics.

Q: How might the Portfolio Manager convert these cluster into actionable insights for portfolio allocation and risk management strategies?

A:

- The Portfolio Manager can use the clusters to identify groups of industries with similar financial profiles.
- This allows for tailored portfolio allocation strategies, such as diversifying investments across clusters to manage risk.
- Additionally, the Manager can develop targeted risk management approaches for each cluster, such as setting specific lending policies or stress testing scenarios that reflect the unique characteristics of each group.

Q: What is the unit that we ultimately want to cluster?

A:

- Industries, which could then be used to build trading signals based upon their associated ETFs.
- The features used to compare those industries will be financial levels and growth rates.

# Data

## Acquisition

Task: Load the data dictionary from the `Data_Dictionary` sheet of `industry_fundamentals.xlsx`.

```{r}
data_dictionary <- readxl::read_excel(
  here("data", "industry_fundamentals.xlsx"),
  sheet = "Data_Dictionary"
)

data_dictionary
```

Task: Before analyzing the observations, how do you expect industries to group based on size, growth, and profitability?

Task: Load the `Data` sheet and call the object `industry_fundamentals`.

```{r}
industry_fundamentals <- readxl::read_excel(
  here("data", "industry_fundamentals.xlsx"),
  sheet = "Data"
)

dplyr::glimpse(industry_fundamentals)
```

## Validate the Data Structure

A clustering algorithm will return an answer even when the input has structural problems. We therefore verify the unit of observation and the variables before engineering features.

Task: Display the fields in the dataset.

```{r}
names(industry_fundamentals)
```

Task: Count rows, distinct industries, and distinct years.

```{r}
industry_fundamentals %>%
  summarise(
    rows = n(),
    industries = n_distinct(Industry),
    years = n_distinct(Year)
  )
```

Q: What is the unit of observation in the source data: industry, year, revenue, or something else?

A:

An industry-year. Each row describes one industry in one year.

Task: Verify that each industry-year key is unique.

```{r}
duplicate_keys <- industry_fundamentals %>%
  dplyr::count(Industry, Year, name = "rows_per_key") %>%
  filter(rows_per_key > 1)

duplicate_keys
stopifnot(nrow(duplicate_keys) == 0)
```

Task: Verify that every industry appears in every year represented in the data.

```{r}
analysis_years <- sort(unique(industry_fundamentals$Year))
stopifnot(length(analysis_years) == 2)

year_1 <- analysis_years[1]
year_2 <- analysis_years[2]

incomplete_industries <- industry_fundamentals %>%
  dplyr::count(Industry, name = "observed_years") %>%
  filter(observed_years != length(analysis_years))

incomplete_industries
stopifnot(nrow(incomplete_industries) == 0)
```

## Cleaning and Accounting Checks

Task: Count missing values by field

```{r}
missing_by_field <- industry_fundamentals %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "field",
    values_to = "missing_values"
  )

missing_by_field
stopifnot(sum(missing_by_field$missing_values) == 0)
```

Task: Verify the accounting identity $\text{Profit}=\text{Revenue}-\text{Operating Cost}$.

```{r}
accounting_check <- industry_fundamentals %>%
  summarise(
    maximum_absolute_error = max(
      abs(Profit - (Revenue - Operating_Cost)),
      na.rm = TRUE
    )
  )

accounting_check
```

Q: Why does this identity matter for clustering?

A:

Profit does not add a new independent dimension when it is exactly Revenue minus Operating Cost. Adding it can nevertheless change K-means because it adds another axis to the distance calculation and therefore changes the weighting of the feature space. We will demonstrate that distinction later.

## Engineering Growth Rates

Task: Compute simple growth rates for each industry from the first year to the second year.

```{r}
growth_rates <- industry_fundamentals %>%
  arrange(Industry, Year) %>%
  group_by(Industry) %>%
  summarise(
    Revenue_Year1 = first(Revenue),
    Revenue_Year2 = last(Revenue),
    Operating_Cost_Year1 = first(Operating_Cost),
    Operating_Cost_Year2 = last(Operating_Cost),
    Profit_Year1 = first(Profit),
    Profit_Year2 = last(Profit),
    .groups = "drop"
  ) %>%
  mutate(
    Revenue_Growth = Revenue_Year2 / Revenue_Year1 - 1,
    Cost_Growth = Operating_Cost_Year2 / Operating_Cost_Year1 - 1,
    Profit_Growth = Profit_Year2 / Profit_Year1 - 1
  )

head(growth_rates)
```

Task: Inspect the Year 1 denominators used to calculate growth.

```{r}
growth_denominator_check <- growth_rates %>%
  summarise(
    minimum_revenue_base = min(Revenue_Year1),
    minimum_cost_base = min(Operating_Cost_Year1),
    minimum_absolute_profit_base = min(abs(Profit_Year1)),
    nonpositive_profit_bases = sum(Profit_Year1 <= 0)
  )

growth_denominator_check

stopifnot(
  all(growth_rates$Revenue_Year1 != 0),
  all(growth_rates$Operating_Cost_Year1 != 0),
  all(growth_rates$Profit_Year1 != 0)
)
```

Q: Why should Profit Growth be interpreted more cautiously than Revenue Growth or Cost Growth?

A:

- A conventional growth rate can become unstable or economically awkward when the starting value is close to zero or negative.
- The denominator check should therefore accompany any use of Profit Growth. In a dataset with problematic profit bases, a change in profit margin or a dollar change in profit may be a better feature.\
- Here we have 0 nonpositive profit bases, which rules out sign flips and division by zero. That alone does not make Profit Growth "safe," however: a *small positive* initial profit is still a tiny denominator, so even a modest dollar change in profit can produce an enormous percentage growth rate. Profit Growth is therefore inherently more volatile than Revenue Growth or Cost Growth. For that reason—and because Profit is exactly Revenue minus Operating Cost—Profit Growth is excluded from the main clustering feature set; the main analysis clusters on standardized Revenue Growth and Cost Growth only.

## Construct the Analysis Samples

K-means is sensitive to the sample used to estimate means, standard deviations, and centroids. We will standardize each feature within the cross-sectional sample actually being clustered.

Task: Construct full Year 1 levels sample and full industry growth sample, then create standardized versions of each.

```{r}
z_score <- function(x) {
  as.numeric(scale(x))
}

year1 <- industry_fundamentals %>%
  filter(Year == year_1)

year1_std_full <- year1 %>%
  mutate(
    Revenue_z = z_score(Revenue),
    Operating_Cost_z = z_score(Operating_Cost),
    Profit_z = z_score(Profit)
  )

growth_rates_std_full <- growth_rates %>%
  mutate(
    Revenue_Growth_z = z_score(Revenue_Growth),
    Cost_Growth_z = z_score(Cost_Growth),
    Profit_Growth_z = z_score(Profit_Growth)
  )
```

# Exploratory Data Analysis

We will focus on exploratory analysis that affects the clustering choices made later.

## Levels and Growth Distributions

Task: For Year 1, create side-by-side violin plots of Revenue, Operating Cost, and Profit in their raw dollar units.

```{r}
year1_long <- year1 %>%
  pivot_longer(
    cols = c(Revenue, Operating_Cost, Profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(year1_long, aes(x = metric, y = value)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.05, alpha = 0.45) +
  labs(
    title = "Year 1 Financial Levels",
    x = NULL,
    y = "$ millions"
  )
```

Task: Create side-by-side violin plots of Revenue Growth, Cost Growth, and Profit Growth.

```{r}
growth_long <- growth_rates %>%
  pivot_longer(
    cols = c(Revenue_Growth, Cost_Growth, Profit_Growth),
    names_to = "metric",
    values_to = "value"
  )

ggplot(growth_long, aes(x = metric, y = value)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.05, alpha = 0.45) +
  labs(
    title = "Growth Rates from Year 1 to Year 2",
    x = NULL,
    y = "Growth rate"
  )
```

## Identify Influential Observations

An observation can be unusual in the levels feature space but ordinary in the growth feature space, or vice versa.

Task: Identify observations at least three standard deviations from the mean in either feature space.

```{r}
level_outlier_flags <- year1_std_full %>%
  select(Industry, Revenue_z, Operating_Cost_z, Profit_z) %>%
  pivot_longer(
    cols = -Industry,
    names_to = "metric",
    values_to = "z_score"
  ) %>%
  filter(abs(z_score) >= 3) %>%
  arrange(desc(abs(z_score)))

growth_outlier_flags <- growth_rates_std_full %>%
  select(Industry, Revenue_Growth_z, Cost_Growth_z, Profit_Growth_z) %>%
  pivot_longer(
    cols = -Industry,
    names_to = "metric",
    values_to = "z_score"
  ) %>%
  filter(abs(z_score) >= 3) %>%
  arrange(desc(abs(z_score)))

level_outlier_flags
growth_outlier_flags
```

Task: Inspect the complete records for the two influential industries identified by the plots and z-scores.

```{r}
year1 %>%
  filter(Industry == "Industry32")

growth_rates %>%
  filter(Industry == "Industry07")
```

Q: What do the plots and diagnostic tables suggest? Are we looking at one outlier or more than one?

A:

- There are two different influential observations in two different feature spaces.
- `Industry32` is extremely large in Revenue, Operating Cost, and Profit, but its growth rates are ordinary. It is unusual in size, not necessarily in operating regime.
- `Industry07` is extreme in the two features that drive the main clustering: its Revenue Growth and Cost Growth are each roughly five standard deviations above the mean, with costs outpacing revenue. Its Profit Growth is also extreme, but the exclusion is warranted on the Revenue Growth and Cost Growth values alone. Its dollar levels are not unusually large.
- “Unusual” does not mean “incorrect.” We will retain the full datasets, exclude the influential observation from the corresponding main analysis to reveal the broader structure, and then add each observation back in a formal sensitivity analysis.

Task: Build a levels sample excluding `Industry32` and a growth-rate sample excluding `Industry07`. Re-standardize within each trimmed sample.

```{r}
year1_trim <- year1 %>%
  filter(Industry != "Industry32") %>%
  mutate(
    Revenue_z = z_score(Revenue),
    Operating_Cost_z = z_score(Operating_Cost),
    Profit_z = z_score(Profit)
  )

growth_rates_trim <- growth_rates %>%
  filter(Industry != "Industry07") %>%
  mutate(
    Revenue_Growth_z = z_score(Revenue_Growth),
    Cost_Growth_z = z_score(Cost_Growth),
    Profit_Growth_z = z_score(Profit_Growth)
  )

nrow(year1_trim)
nrow(growth_rates_trim)
```

## Correlation and Outlier Influence

Correlation is unchanged by positive linear rescaling, so it is clearest to compute these correlations in the original growth-rate units.

```{r}
correlation_long <- function(correlation_matrix) {
  as.data.frame(as.table(correlation_matrix)) %>%
    as_tibble() %>%
    rename(
      variable_1 = Var1,
      variable_2 = Var2,
      correlation = Freq
    )
}

plot_correlation <- function(correlation_matrix, plot_title) {
  correlation_long(correlation_matrix) %>%
    ggplot(aes(variable_1, variable_2, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(
      low = "blue",
      high = "red",
      mid = "white",
      midpoint = 0,
      limits = c(-1, 1),
      name = "Correlation"
    ) +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme_minimal()
}
```

Task: Compute and visualize correlations using all industries.

```{r}
growth_cor_full <- growth_rates %>%
  select(Revenue_Growth, Cost_Growth, Profit_Growth) %>%
  cor(use = "complete.obs")

growth_cor_full
plot_correlation(growth_cor_full, "Growth-Rate Correlations: All Industries")
```

Task: Recompute and visualize the correlations after excluding `Industry07`.

```{r}
growth_cor_trim <- growth_rates_trim %>%
  select(Revenue_Growth, Cost_Growth, Profit_Growth) %>%
  cor(use = "complete.obs")

growth_cor_trim
plot_correlation(growth_cor_trim, "Growth-Rate Correlations: Industry07 Excluded")
```

Q: How does removing `Industry07` change the correlation structure?

A:

- Revenue Growth and Cost Growth remain strongly positively correlated: $r=$ `r round(growth_cor_full["Revenue_Growth", "Cost_Growth"], 2)` with the observation included and $r=$ `r round(growth_cor_trim["Revenue_Growth", "Cost_Growth"], 2)` after it is excluded.
- The relationship between Profit Growth and Revenue Growth changes from $r=$ `r round(growth_cor_full["Profit_Growth", "Revenue_Growth"], 2)` to $r=$ `r round(growth_cor_trim["Profit_Growth", "Revenue_Growth"], 2)`.
- The relationship between Profit Growth and Cost Growth changes from $r=$ `r round(growth_cor_full["Profit_Growth", "Cost_Growth"], 2)` to $r=$ `r round(growth_cor_trim["Profit_Growth", "Cost_Growth"], 2)`.
- The broader lesson is that one influential observation can materially alter a statistic that will later shape the geometry and interpretation of a clustering analysis.

# K-Means

## Intuition and Algorithm

Q: How does K-means cluster observations?

A:

- K-means groups observations that are close to one another in the selected feature space.
- Closeness is usually measured with Euclidean distance.
- The analyst chooses the number of clusters, $K$, in advance.
- The algorithm selects $K$ initial cluster centers, called centroids.
- Each observation is assigned to its nearest centroid.
- Each centroid is recomputed as the mean of the observations assigned to that cluster.
- Assignment and centroid updates repeat until no further reassignment improves the objective, or until an implementation-specific stopping rule is reached.
- The final solution minimizes the total squared distance from observations to their assigned centroid, subject to the local solution reached from the starting values.

Q: Does K-means cluster observations or variables?

A:

Observations. The variables define the coordinates used to measure distance between those observations. Here, the observations are industries and the variables are financial growth rates or levels.

## A Basic Fit

For the basic mechanics, use the trimmed Revenue Growth and Cost Growth sample. We set `nstart = 1` here so that the relationship between random initialization and reproducibility remains visible.

Task: Fit K-means with $K=3$.

```{r}
basic_features <- growth_rates_trim %>%
  select(Revenue_Growth, Cost_Growth)

set.seed(1234)
basic_kmeans <- kmeans(
  basic_features,
  centers = 3,
  nstart = 1
)
```

Task: Display each industry with its assigned cluster.

```{r}
basic_assignments <- tibble(
  Industry = growth_rates_trim$Industry,
  cluster = basic_kmeans$cluster
)

basic_assignments
```

Q: How should the cluster vector be interpreted?

A:

Each industry receives one cluster label. The numbers 1, 2, and 3 are arbitrary identifiers, not rankings. “Cluster 3” is not intrinsically larger, riskier, or better than “Cluster 1.” Interpret a cluster only after examining its centroid and members.

Task: Display the cluster centroids.

```{r}
basic_centers <- basic_kmeans$centers %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cluster") %>%
  as_tibble()

basic_centers
```

Q: How should the centroids be interpreted?

A:

Each row contains the mean Revenue Growth and mean Cost Growth of the industries assigned to that cluster. Because the fit used raw growth rates, the centroid values are also in raw growth-rate units.

## The K-Means Objective and R Output

Define:

- $\mathbf{x}_i$: the $p \times 1$ feature vector for observation $i$.
- $\bar{\mathbf{x}}$: the overall sample mean vector.
- $\boldsymbol{\mu}_k$: the centroid of cluster $k$.
- $C_k$: the set of observations assigned to cluster $k$.
- $n_k$: the number of observations in cluster $k$.

The main quantities returned by `kmeans()` are:

### Total Sum of Squares: `totss`

$$
\mathrm{TSS}
=
\sum_{i=1}^{N}
\left\|\mathbf{x}_i-\bar{\mathbf{x}}\right\|^2.
$$

This measures total dispersion around the overall mean before using the cluster assignments.

### Within-Cluster Sum of Squares: `withinss`

For cluster $k$,

$$
\mathrm{WithinSS}_k
=
\sum_{i \in C_k}
\left\|\mathbf{x}_i-\boldsymbol{\mu}_k\right\|^2.
$$

This measures the total squared distance of the observations in cluster $k$ from that cluster’s centroid.

### Total Within-Cluster Sum of Squares: `tot.withinss`

$$
\mathrm{Total\ WithinSS}
=
\sum_{k=1}^{K}\mathrm{WithinSS}_k.
$$

This is the objective that K-means attempts to minimize.

### Between-Cluster Sum of Squares: `betweenss`

$$
\mathrm{BetweenSS}
=
\sum_{k=1}^{K}
 n_k
\left\|\boldsymbol{\mu}_k-\bar{\mathbf{x}}\right\|^2.
$$

This measures how far the cluster centroids are from the overall mean, weighted by cluster size.

The decomposition is

$$
\boxed{
\mathrm{TSS}
=
\mathrm{Total\ WithinSS}
+
\mathrm{BetweenSS}
}.
$$

Task: Display the fit statistics and verify the decomposition.

```{r}
basic_fit_statistics <- tibble(
  totss = basic_kmeans$totss,
  tot_withinss = basic_kmeans$tot.withinss,
  betweenss = basic_kmeans$betweenss,
  decomposition_error = basic_kmeans$totss -
    basic_kmeans$tot.withinss -
    basic_kmeans$betweenss,
  between_share = basic_kmeans$betweenss / basic_kmeans$totss
)

basic_fit_statistics
```

Task: Compare cluster size, `withinss`, and `withinss` per observation.

```{r}
basic_cluster_statistics <- tibble(
  cluster = seq_along(basic_kmeans$size),
  size = basic_kmeans$size,
  withinss = basic_kmeans$withinss,
  withinss_per_observation = basic_kmeans$withinss / basic_kmeans$size
)

basic_cluster_statistics
```

Q: Are these sums of squares best interpreted in relative or absolute terms?

A:

Usually relative.

- `totss` is fixed only when the observations, features, units, and scaling are fixed.
- `tot.withinss` can be compared across alternative fits on the same feature matrix, but it falls mechanically as $K$ increases and therefore cannot choose $K$ by itself.
- `betweenss / totss` describes how much total dispersion is represented between the fitted clusters, but it is not a universal goodness-of-fit threshold.
- A large cluster often has a large `withinss` simply because it contains more observations. `withinss / size` is more informative when comparing average squared dispersion across clusters, although it still depends on the features and their scaling.

## Random Initialization and `set.seed()`

K-means can converge to different local solutions from different random starting centroids. In addition, cluster numbers can be permuted even when the underlying partition is the same.

Task: Run the same one-start specification twice without resetting the seed, then compare both the objective value and the partition. Also compute the "adjustedRandIndex".

Note: The Adjusted Rand Index (ARI) measures how similar two cluster assignments are while ignoring the arbitrary numeric cluster labels, with 1 indicating identical clustering and values near 0 indicating agreement no better than chance.

```{r}
set.seed(9001)
km_random_1 <- kmeans(basic_features, centers = 3, nstart = 1)
km_random_2 <- kmeans(basic_features, centers = 3, nstart = 1)

c(
  run_1_tot_withinss = km_random_1$tot.withinss,
  run_2_tot_withinss = km_random_2$tot.withinss
)

mclust::adjustedRandIndex(
  km_random_1$cluster,
  km_random_2$cluster
)
```

Q: Why might the two results differ even though the model specification did not change?

A:

Each call can begin from different random centroids. A different objective value indicates a substantively different local solution. A high Adjusted Rand Index with different-looking numeric labels may instead reflect simple label switching.

Task: Reset to the same seed before each run and verify exact reproducibility.

```{r}
set.seed(1234)
km_seed_1 <- kmeans(basic_features, centers = 3, nstart = 1)

set.seed(1234)
km_seed_2 <- kmeans(basic_features, centers = 3, nstart = 1)

identical(km_seed_1$cluster, km_seed_2$cluster)
identical(km_seed_1$centers, km_seed_2$centers)
```

Q: What does `set.seed()` accomplish?

A:

`set.seed()` makes the random initialization reproducible. It does not guarantee that a one-start fit is the best available solution. We address that issue with `nstart` in the sensitivity section.

# Why Feature Redundancy Matters

This section is a self-contained pedagogical demonstration built on the **Year 1 levels** (Revenue, Operating Cost, and Profit). Its purpose is to show that adding a redundant, exactly-determined feature can still change a K-means partition by reweighting the distance geometry. The three-variable levels specification used here is **not** the specification used for the main clustering analysis; the main analysis clusters industries on two growth features only (standardized Revenue Growth and Cost Growth), as set up in the "Altering the Number of Clusters" section below.

Q: Suppose we cluster standardized Year 1 Revenue and Operating Cost. What do you expect to happen if standardized Profit is added?

A:

Because Profit is exactly Revenue minus Operating Cost, it adds no new independent accounting information. However, adding Profit adds another term to Euclidean distance and therefore reweights the geometry. We expect high—but not necessarily perfect—agreement.

Task: Verify the accounting identity in the trimmed Year 1 sample.

```{r}
year1_trim %>%
  summarise(
    maximum_absolute_error = max(
      abs(Profit - (Revenue - Operating_Cost))
    )
  )
```

Task: Fit $K=2$ using standardized Revenue and Operating Cost.

```{r}
set.seed(1234)
km_2feat <- kmeans(
  year1_trim %>% select(Revenue_z, Operating_Cost_z),
  centers = 2,
  nstart = 20
)
```

Task: Fit $K=2$ using standardized Revenue, Operating Cost, and Profit.

```{r}
set.seed(1234)
km_all3 <- kmeans(
  year1_trim %>% select(Revenue_z, Operating_Cost_z, Profit_z),
  centers = 2,
  nstart = 20
)
```

Task: Compute cluster profiles for both specifications.

```{r}
year1_feature_comparison <- year1_trim %>%
  mutate(
    cluster_2feat = km_2feat$cluster,
    cluster_all3 = km_all3$cluster
  )

year1_feature_comparison %>%
  group_by(cluster_2feat) %>%
  summarise(
    across(c(Revenue, Operating_Cost, Profit), mean),
    n = n(),
    .groups = "drop"
  )

year1_feature_comparison %>%
  group_by(cluster_all3) %>%
  summarise(
    across(c(Revenue, Operating_Cost, Profit), mean),
    n = n(),
    .groups = "drop"
  )
```

Cluster labels are arbitrary, so they must be aligned before identifying observations that “switch” numeric labels. The helper below is specific to a two-cluster comparison.

```{r}
align_two_cluster_labels <- function(reference, candidate) {
  same_label_agreement <- sum(reference == candidate)
  swapped_label_agreement <- sum(reference == (3 - candidate))

  if (swapped_label_agreement > same_label_agreement) {
    3 - candidate
  } else {
    candidate
  }
}

year1_feature_comparison <- year1_feature_comparison %>%
  mutate(
    cluster_2feat_aligned = align_two_cluster_labels(
      reference = cluster_all3,
      candidate = cluster_2feat
    )
  )

feature_crosstab <- table(
  All3 = year1_feature_comparison$cluster_all3,
  TwoFeat = year1_feature_comparison$cluster_2feat
)

feature_switches <- year1_feature_comparison %>%
  filter(cluster_all3 != cluster_2feat_aligned) %>%
  select(Industry, cluster_all3, cluster_2feat_aligned)

feature_ari <- mclust::adjustedRandIndex(
  year1_feature_comparison$cluster_all3,
  year1_feature_comparison$cluster_2feat
)

feature_switch_text <- if (nrow(feature_switches) == 0) {
  "none"
} else {
  paste(feature_switches$Industry, collapse = ", ")
}

feature_crosstab
feature_switches
feature_ari
```

Q: How much did cluster membership change after Profit was added?

A:

The Adjusted Rand Index is `r round(feature_ari, 2)`, where 1 indicates identical partitions and 0 is the agreement expected by chance under the index’s adjustment. After aligning the arbitrary numeric labels, `r nrow(feature_switches)` industries changed assignment: `r feature_switch_text`.

Q: If Profit adds no new independent information, why can it change the clusters?

A:

Standardizing each field separately produces three axes with unit variance. The all-three-feature objective includes a Profit term,

$$
d^2
=
(\Delta z_{Revenue})^2
+
(\Delta z_{Cost})^2
+
(\Delta z_{Profit})^2,
$$

whereas the two-feature objective omits the final term. The data still lie on a two-dimensional accounting plane, but distances along directions associated with Profit receive additional weight. A borderline observation can therefore move to another nearest centroid.

# Why Standardization Matters

K-means is not invariant to rescaling. Multiplying one feature by 100 multiplies its contribution to squared Euclidean distance by $100^2$.

## Raw Versus Standardized Features in This Dataset

Task: Compare $K=2$ fits for raw and standardized Year 1 levels.

```{r}
set.seed(4821)
km_raw_levels <- kmeans(
  year1_trim %>% select(Revenue, Operating_Cost),
  centers = 2,
  nstart = 20
)

set.seed(4821)
km_std_levels <- kmeans(
  year1_trim %>% select(Revenue_z, Operating_Cost_z),
  centers = 2,
  nstart = 20
)
```

Task: Compare $K=2$ fits for raw and standardized growth rates.

```{r}
set.seed(4821)
km_raw_growth <- kmeans(
  growth_rates_trim %>% select(Revenue_Growth, Cost_Growth),
  centers = 2,
  nstart = 20
)

set.seed(4821)
km_std_growth <- kmeans(
  growth_rates_trim %>% select(Revenue_Growth_z, Cost_Growth_z),
  centers = 2,
  nstart = 20
)
```

Task: Quantify agreement and compare the raw feature standard deviations.

```{r}
scale_agreement <- tibble(
  comparison = c("Year 1 levels", "Growth rates"),
  adjusted_rand_index = c(
    mclust::adjustedRandIndex(
      km_raw_levels$cluster,
      km_std_levels$cluster
    ),
    mclust::adjustedRandIndex(
      km_raw_growth$cluster,
      km_std_growth$cluster
    )
  )
)

feature_scale_summary <- bind_rows(
  year1_trim %>%
    summarise(
      Revenue = sd(Revenue),
      Operating_Cost = sd(Operating_Cost)
    ) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "standard_deviation") %>%
    mutate(sample = "Year 1 levels"),
  growth_rates_trim %>%
    summarise(
      Revenue_Growth = sd(Revenue_Growth),
      Cost_Growth = sd(Cost_Growth)
    ) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "standard_deviation") %>%
    mutate(sample = "Growth rates")
)

scale_agreement
feature_scale_summary
```

Q: Did standardization materially change these particular two-feature partitions?

A:

No. The Adjusted Rand Index is 1 for both comparisons, meaning that standardizing the variables produced exactly the same cluster assignments as using the raw variables. In this dataset, the two variables within each pair have reasonably similar variation, so standardization does not substantially change their relative influence on Euclidean distance. This result does not imply that standardization is generally irrelevant; when variables differ greatly in scale, standardization can materially change the clusters.

## Mixed Units

To make the scaling issue explicit, combine a dollar level with a growth rate. This is an illustrative geometry exercise, not necessarily the preferred economic specification.

Task: Cluster industries on Year 1 Revenue and Revenue Growth before and after standardization.

```{r}
mixed_scale_features <- year1_trim %>%
  select(Industry, Revenue) %>%
  inner_join(
    growth_rates_trim %>% select(Industry, Revenue_Growth),
    by = "Industry"
  ) %>%
  mutate(
    Revenue_z = z_score(Revenue),
    Revenue_Growth_z = z_score(Revenue_Growth)
  )

set.seed(4821)
km_mixed_raw <- kmeans(
  mixed_scale_features %>% select(Revenue, Revenue_Growth),
  centers = 3,
  nstart = 20
)

set.seed(4821)
km_mixed_std <- kmeans(
  mixed_scale_features %>% select(Revenue_z, Revenue_Growth_z),
  centers = 3,
  nstart = 20
)

mixed_scale_ari <- mclust::adjustedRandIndex(
  km_mixed_raw$cluster,
  km_mixed_std$cluster
)

mixed_scale_ari
```

Task: Compare cluster profiles in the original units.

```{r}
mixed_scale_features %>%
  mutate(
    raw_cluster = km_mixed_raw$cluster,
    standardized_cluster = km_mixed_std$cluster
  ) %>%
  group_by(raw_cluster) %>%
  summarise(
    Revenue = mean(Revenue),
    Revenue_Growth = mean(Revenue_Growth),
    n = n(),
    .groups = "drop"
  )

mixed_scale_features %>%
  mutate(
    raw_cluster = km_mixed_raw$cluster,
    standardized_cluster = km_mixed_std$cluster
  ) %>%
  group_by(standardized_cluster) %>%
  summarise(
    Revenue = mean(Revenue),
    Revenue_Growth = mean(Revenue_Growth),
    n = n(),
    .groups = "drop"
  )
```

Q: Why is standardization more consequential in the mixed-unit example?

A:

In raw units, dollar Revenue has much larger numerical variation and dominates Euclidean distance. After standardization, a one-standard-deviation difference in Revenue receives the same initial weight as a one-standard-deviation difference in Revenue Growth. Standardization is therefore also an implicit feature-weighting decision.

# Altering the Number of Clusters

We now return to the trimmed, standardized growth sample. From this point forward, the main clustering analysis uses **exactly two features**: standardized Revenue Growth and standardized Cost Growth. Profit Growth is deliberately excluded—it carries no independent accounting information beyond Revenue and Operating Cost, and its percentage-growth form is unstable when the initial profit base is small. The feature matrix is therefore genuinely two-dimensional, and we name it `X_growth_2feat` to make its composition explicit.

```{r}
growth_features <- growth_rates_trim %>%
  select(Industry, Revenue_Growth_z, Cost_Growth_z)

X_growth_2feat <- growth_features %>%
  select(-Industry) %>%
  as.matrix()

fit_kmeans <- function(k, seed = 4821, nstart = 20) {
  set.seed(seed)
  kmeans(X_growth_2feat, centers = k, nstart = nstart)
}

growth_scatter <- function(fit) {
  plot_data <- growth_features %>%
    mutate(cluster = factor(fit$cluster))

  ggplot(
    plot_data,
    aes(Revenue_Growth_z, Cost_Growth_z, color = cluster)
  ) +
    geom_point(size = 2) +
    labs(
      x = "Revenue Growth (z)",
      y = "Cost Growth (z)",
      color = "Cluster"
    )
}

cluster_profile <- function(fit) {
  growth_rates_trim %>%
    mutate(cluster = fit$cluster) %>%
    group_by(cluster) %>%
    summarise(
      across(c(Revenue_Growth, Cost_Growth), mean),
      n = n(),
      .groups = "drop"
    ) %>%
    arrange(Revenue_Growth)
}
```

## $K=1$

Task: Fit and visualize one cluster.

```{r}
km_k1 <- fit_kmeans(1)
growth_scatter(km_k1)
cluster_profile(km_k1)
```

## $K=2$

Task: Repeat for two clusters.

```{r}
km_k2 <- fit_kmeans(2)
growth_scatter(km_k2)
cluster_profile(km_k2)
```

## $K=3$

Task: Repeat for three clusters.

```{r}
km_k3 <- fit_kmeans(3)
growth_scatter(km_k3)
cluster_profile(km_k3)
```

## $K=4$

Task: Repeat for four clusters.

```{r}
km_k4 <- fit_kmeans(4)
growth_scatter(km_k4)
cluster_profile(km_k4)
```

Numeric cluster labels cannot be tracked meaningfully across different values of $K$. Instead, track both the assignment and the centroid profile attached to that assignment.

```{r}
membership_with_centroid <- function(fit, k_label) {
  centers <- fit$centers %>%
    as.data.frame() %>%
    tibble::rownames_to_column("cluster") %>%
    mutate(cluster = as.integer(cluster)) %>%
    as_tibble()

  tibble(
    Industry = growth_features$Industry,
    cluster = fit$cluster
  ) %>%
    left_join(centers, by = "cluster") %>%
    mutate(K = k_label, .before = Industry)
}

tracked_industries <- bind_rows(
  membership_with_centroid(km_k1, "K = 1"),
  membership_with_centroid(km_k2, "K = 2"),
  membership_with_centroid(km_k3, "K = 3"),
  membership_with_centroid(km_k4, "K = 4")
) %>%
  filter(Industry %in% c("Industry01", "Industry44")) %>%
  arrange(Industry, K)

tracked_industries
```

Q: What happens as $K$ increases?

A:

- At $K=1$, all observations share one centroid.
- Increasing $K$ gives the algorithm more centroids and therefore weakly lowers total within-cluster sum of squares.
- In this simulation, the profile tables indicate that $K=2$ separates a broadly flat-growth group from a high-growth group; $K=3$ subdivides the flat-growth group into a modest-growth group and a mild-contraction group, leaving the high-growth group intact; and $K=4$ further subdivides the high-growth group.
- K-means solutions are not hierarchical. Although the clusters may appear to split neatly in this dataset, moving from $K$ to $K+1$ can reorganize several clusters rather than simply dividing one existing cluster.
- More clusters create more detailed segments, but eventually the extra detail may represent noise, unstable boundaries, or groups too small to support a separate business policy.

# Choosing $K$

No single diagnostic mechanically determines the economically “correct” number of clusters. We combine statistical evidence with interpretability, stability, and the intended decision use.

## Elbow Method

The elbow method examines how quickly total within-cluster sum of squares falls as $K$ increases. We look for a point after which additional clusters provide much smaller improvements.

Task: Construct an elbow plot for $K=1,\ldots,10$.

```{r}
wss <- purrr::map_dbl(1:10, function(k) {
  fit_kmeans(k, seed = 4821, nstart = 20)$tot.withinss
})

elbow_df <- tibble(
  k = 1:10,
  wss = wss
) %>%
  mutate(
    reduction = lag(wss) - wss,
    proportional_reduction = reduction / lag(wss)
  )

elbow_df

ggplot(elbow_df, aes(k, wss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Plot",
    x = "Number of clusters (K)",
    y = "Total within-cluster sum of squares"
  )
```

Q: How many clusters does the elbow plot suggest?

A:

The elbow is fairly pronounced at $K=2$: total within-cluster sum of squares falls by roughly 84% from $K=1$ to $K=2$, and by much less thereafter. $K=2$ is the clearest candidate, with $K=3$ a weaker secondary option. Because the elbow is rarely definitive on its own, we examine additional diagnostics.

## Gap Statistic

The gap statistic compares the observed within-cluster dispersion with dispersion expected from reference datasets that contain no comparable cluster structure.

From the gap plot, we look for the first K where the gap curve "levels off".

For a more rigorous approach, we can follow the Tibs2001SEMax rule: Choose the smallest K for which adding another cluster doesn’t improve the Gap statistic by more than one standard error.

Task: Construct a gap-statistic plot for $K=1,\ldots,10$.

```{r}
set.seed(4821)
gap_stat <- cluster::clusGap(
  X_growth_2feat,
  FUNcluster = function(x, k) {
    kmeans(x, centers = k, nstart = 20)
  },
  K.max = 10,
  B = 100
)

gap_df <- gap_stat$Tab %>%
  as.data.frame() %>%
  tibble::rownames_to_column("k") %>%
  as_tibble() %>%
  mutate(k = as.integer(k))

gap_k <- cluster::maxSE(
  gap_stat$Tab[, "gap"],
  gap_stat$Tab[, "SE.sim"],
  method = "Tibs2001SEmax"
)

gap_df

ggplot(gap_df, aes(k, gap)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(ymin = gap - SE.sim, ymax = gap + SE.sim),
    width = 0.2
  ) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Gap Statistic",
    x = "Number of clusters (K)",
    y = "Gap"
  )

gap_k
```

Q: How many clusters does the gap statistic suggest?

A:

Using the `Tibs2001SEmax` rule, the gap statistic suggests $K=$ `r gap_k`.

## Silhouette Width

For each observation, silhouette width compares cohesion with its assigned cluster against separation from the nearest alternative cluster. Values near 1 indicate strong assignment, values near 0 indicate a boundary observation, and negative values suggest possible misassignment.

Task: Compute average silhouette width for $K=2,\ldots,10$.

```{r}
growth_distance <- dist(X_growth_2feat)

silhouette_scores <- purrr::map_dfr(2:10, function(k) {
  fit <- fit_kmeans(k, seed = 4821, nstart = 20)
  sil <- cluster::silhouette(fit$cluster, growth_distance)

  tibble(
    k = k,
    average_silhouette_width = mean(sil[, "sil_width"])
  )
})

silhouette_k <- silhouette_scores %>%
  slice_max(average_silhouette_width, n = 1, with_ties = FALSE) %>%
  pull(k)

silhouette_scores

ggplot(
  silhouette_scores,
  aes(k, average_silhouette_width)
) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Average Silhouette Width by K",
    x = "Number of clusters (K)",
    y = "Average silhouette width"
  )

silhouette_k
```

Task: Inspect the detailed silhouette plot for the best-scoring $K$.

```{r}
km_silhouette_best <- fit_kmeans(
  silhouette_k,
  seed = 4821,
  nstart = 20
)

silhouette_best <- cluster::silhouette(
  km_silhouette_best$cluster,
  growth_distance
)

plot(
  silhouette_best,
  main = paste("Silhouette Plot, K =", silhouette_k)
)
```

Q: Which cluster appears most internally coherent and well separated from the other clusters?

A:

At $K=2$ the two clusters are almost equally coherent. The 32-industry mature/flat-growth cluster has an average silhouette width of about 0.74, and the 17-industry high-growth cluster is essentially the same at about 0.74. Both clusters are compact and well separated; most observations in each have large positive silhouette widths.

Q: Which cluster appears least clearly separated?

A:

The two clusters are close. The mature/flat-growth cluster contains the single most borderline observation (its lowest individual silhouette width is about 0.52), whereas the high-growth cluster's lowest is about 0.66. Even so, no observation has a silhouette width near zero or negative, so both clusters are cleanly assigned.

Task: Fit K=3, inspect its silhouette plot.

```{r}
km_silhouette_alt <- fit_kmeans(
  3,
  seed = 4821,
  nstart = 20
)

silhouette_alt <- cluster::silhouette(
  km_silhouette_alt$cluster,
  growth_distance
)

plot(
  silhouette_alt,
  main = paste("Silhouette Plot, K =", 3)
)

km_2 <- fit_kmeans(2, seed = 4821, nstart = 20)
km_3 <- fit_kmeans(3, seed = 4821, nstart = 20)
```

Task: Create a cross-tabulation of the K=2 and K=3 assignments to determine how the original clusters were divided.

```{r}
table(
  K2 = km_2$cluster,
  K3 = km_3$cluster
)
```

Q: How read this cross tab? What's the implication?

A:

- The K=2 high-growth cluster (17 industries) stays intact as a single K=3 cluster; none of its members are reassigned.
- The K=2 mature/flat-growth cluster (32 industries) is the one that splits: under K=3 it divides into a 21-industry modest-growth group and an 11-industry mild-contraction group.
- Moving from K=2 to K=3 therefore refines the flat-growth regime rather than reorganizing the whole partition.

Q: Does allowing K-means to form a third cluster improve the quality of the clustering?

A:

No. The overall average silhouette width falls from about 0.74 for K=2 to about 0.64 for K=3. Although K=3 creates an additional group and will necessarily reduce within-cluster sum of squares, the resulting clusters are, on average, less clearly separated from their nearest alternatives.

Q: Why can tot.withinss improve when moving from K=2 to K=3, while average silhouette width gets worse?

A:

Adding another cluster gives K-means more flexibility to place observations closer to a centroid, so tot.withinss must weakly decrease. Silhouette width also considers how well each observation is separated from the nearest competing cluster. Splitting observations into more clusters can therefore improve within-cluster fit while making neighboring clusters less distinct.

Q: How many clusters does the silhouette analysis suggest?

A:

Average silhouette width is maximized at $K=$ `r silhouette_k` over the candidate range.

## Synthesis

Task: Summarize the three diagnostics.

```{r}
selection_summary <- tibble(
  method = c("Elbow", "Gap statistic", "Silhouette"),
  suggested_k = c(
    "2; pronounced first drop",
    as.character(gap_k),
    as.character(silhouette_k)
  )
)

selection_summary
```

Q: How many clusters should we use?

A:

We will use $K=2$ for the remainder of the lesson. All three diagnostics point the same way: the elbow drops sharply through $K=2$, the gap statistic selects $K=2$ under the `Tibs2001SEmax` rule, and average silhouette width is maximized at $K=2$. The two-cluster solution also has a clean economic interpretation—it separates a mature, flat-growth regime from a high-growth regime in which both revenue and operating costs expand together. A three-cluster solution, which would further split the flat-growth group into a modest-growth and a mild-contraction group, remains a reasonable finer-grained alternative if that distinction becomes decision-relevant.

# Interpreting the Final Clusters

Cluster numbers are arbitrary. Economic names are assigned after reviewing the cluster profiles.

```{r}
chosen_k <- 2L

set.seed(4821)
km_final <- kmeans(
  X_growth_2feat,
  centers = chosen_k,
  nstart = 20
)

growth_clustered <- growth_rates_trim %>%
  mutate(cluster = km_final$cluster)

cluster_profiles <- growth_clustered %>%
  group_by(cluster) %>%
  summarise(
    Revenue_Growth = mean(Revenue_Growth),
    Cost_Growth = mean(Cost_Growth),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    average_operating_growth = (Revenue_Growth + Cost_Growth) / 2
  )

flat_cluster <- cluster_profiles %>%
  slice_min(abs(average_operating_growth), n = 1, with_ties = FALSE) %>%
  pull(cluster)

high_growth_cluster <- setdiff(cluster_profiles$cluster, flat_cluster)

cluster_name_map <- tibble(
  cluster = c(
    flat_cluster,
    high_growth_cluster
  ),
  cluster_name = c(
    "Mature / flat growth",
    "High revenue and cost growth"
  )
)

growth_clustered <- growth_clustered %>%
  left_join(cluster_name_map, by = "cluster")

cluster_profiles <- cluster_profiles %>%
  left_join(cluster_name_map, by = "cluster") %>%
  arrange(average_operating_growth)
```

Task: Display cluster sizes and average growth rates.

```{r}
cluster_profiles %>%
  select(
    cluster,
    cluster_name,
    n,
    Revenue_Growth,
    Cost_Growth
  )
```

Task: Visualize the clusters in the Revenue Growth–Cost Growth plane.

```{r}
ggplot(
  growth_clustered,
  aes(
    Revenue_Growth,
    Cost_Growth,
    color = cluster_name
  )
) +
  geom_point(size = 2) +
  labs(
    title = "Growth-Rate Clusters",
    x = "Revenue Growth",
    y = "Cost Growth",
    color = "Economic label"
  )
```

Because the clustering uses exactly these two features, this scatter plot is a direct and complete visualization of the feature space the algorithm actually used—not a projection. Every distance that determined the cluster assignments is visible in this plane.

Task: Compare Year 1 Revenue and Operating Cost while retaining the growth-based cluster colors.

```{r}
level_view <- year1 %>%
  select(Industry, Revenue, Operating_Cost) %>%
  inner_join(
    growth_clustered %>%
      select(Industry, cluster, cluster_name),
    by = "Industry"
  )

ggplot(
  level_view,
  aes(
    Revenue,
    Operating_Cost,
    color = cluster_name
  )
) +
  geom_point(size = 2) +
  labs(
    title = "Year 1 Size, Colored by Growth-Rate Cluster",
    x = "Revenue",
    y = "Operating Cost",
    color = "Economic label"
  )
```

Task: Display cluster membership.

```{r}
growth_clustered %>%
  select(Industry, cluster, cluster_name) %>%
  arrange(cluster_name, Industry)
```

Q: What is the economic meaning of the final clusters?

A:

- The Mature / flat growth cluster contains industries with Revenue Growth and Cost Growth near zero on average.
- The High revenue and cost growth cluster contains industries where both revenue and operating costs expand substantially, with cost growth modestly outpacing revenue growth on average.
- The names describe the Revenue Growth–Cost Growth profiles the clusters actually produced; they are analyst-created summaries, not outputs supplied by `kmeans()`.
- `Industry32` remains in the growth analysis because it is unusual only in dollar size. The raw-level plot shows why a size outlier can still belong to an ordinary growth regime.
- `Industry07` remains excluded from the main growth fit and will be added back next to measure its influence.

# Sensitivity and Stability

## Outlier in Levels: Add `Industry32` Back

Task: Repeat the $K=2$ feature-redundancy comparison using the full, Year 1-standardized sample.

```{r}
set.seed(4821)
km_all3_full <- kmeans(
  year1_std_full %>% select(Revenue_z, Operating_Cost_z, Profit_z),
  centers = 2,
  nstart = 20
)

set.seed(4821)
km_2feat_full <- kmeans(
  year1_std_full %>% select(Revenue_z, Operating_Cost_z),
  centers = 2,
  nstart = 20
)

full_level_comparison <- year1_std_full %>%
  transmute(
    Industry,
    cluster_all3 = km_all3_full$cluster,
    cluster_2feat = km_2feat_full$cluster
  )

full_level_ari <- mclust::adjustedRandIndex(
  full_level_comparison$cluster_all3,
  full_level_comparison$cluster_2feat
)

table(
  All3 = full_level_comparison$cluster_all3,
  TwoFeat = full_level_comparison$cluster_2feat
)

full_level_comparison %>% 
    dplyr::count(cluster_all3)
full_level_ari
```

Q: What changes once `Industry32` is added back?

A:

- The one observation in cluster 1 under the three-variable model is also in cluster 1 under the two-variable model.
- The other 49 observations are all in cluster 2 under both models.
- No observations switch clusters.
- The ARI=1, which means the two cluster assignments are exactly the same, ignoring arbitrary cluster labels.\
- Adding Profit as a third feature does not change the K=2 partition at all. Both specifications isolate the same single industry from the other 49 industries.

## Outlier in Growth Rates: Add `Industry07` Back

Task: Repeat the $K=2$ growth clustering with all industries and compare the common observations with the trimmed result.

```{r}
X_growth_2feat_full <- growth_rates_std_full %>%
  select(Revenue_Growth_z, Cost_Growth_z) %>%
  as.matrix()

set.seed(4821)
km_full_k2 <- kmeans(
  X_growth_2feat_full,
  centers = chosen_k,
  nstart = 20
)

full_growth_assignments <- tibble(
  Industry = growth_rates_std_full$Industry,
  cluster_full = km_full_k2$cluster
)

full_growth_profiles <- growth_rates %>%
  left_join(full_growth_assignments, by = "Industry") %>%
  group_by(cluster_full) %>%
  summarise(
    across(c(Revenue_Growth, Cost_Growth), mean),
    n = n(),
    .groups = "drop"
  )

compare_growth <- growth_clustered %>%
  select(Industry, cluster_trimmed = cluster) %>%
  left_join(full_growth_assignments, by = "Industry")

growth_outlier_ari <- mclust::adjustedRandIndex(
  compare_growth$cluster_trimmed,
  compare_growth$cluster_full
)

full_growth_profiles

table(
  trimmed = compare_growth$cluster_trimmed,
  full = compare_growth$cluster_full
)

growth_outlier_ari
```

Q: What changes once `Industry07` is added back?

A:

- Industry07's growth features are extreme—about 95% revenue growth and 120% cost growth, each roughly five standard deviations from the mean. K-means devotes an entire centroid to this single observation.
- With only two clusters available, that leaves just one centroid for the other 49 industries, so the mature/flat-growth and high-growth groups collapse into a single cluster.
  - Cluster sizes without outlier: 32, 17
  - Cluster sizes with outlier: 49, 1
- The ARI over the 49 common industries is 0—agreement no better than chance—because every common industry lands in the same single cluster once the outlier is present. One influential observation completely erases the two-regime structure at $K=2$.

Q: How did adding the outlier affect the K=2 solution?

A:

The outlier received a cluster of its own. Because the total number of clusters was fixed at two, only one cluster remained for the other 49 industries, and the mature/flat versus high-growth distinction disappeared entirely. The ARI of 0 shows the two partitions share no structure beyond chance. This is a sharper version of the levels-outlier lesson: at small $K$, a single extreme point can consume an entire centroid and dominate the partition.

## Starting Values and `nstart`

`nstart` repeats K-means from multiple random initial centroid sets and retains the fit with the smallest total within-cluster sum of squares.

Task: Compare `nstart = 1` and `nstart = 20` across 50 random seeds.

```{r}
initialization_results <- purrr::map_dfr(
  c(1L, 20L),
  function(number_of_starts) {
    purrr::map_dfr(1:50, function(seed_value) {
      set.seed(seed_value)

      fit <- kmeans(
        X_growth_2feat,
        centers = chosen_k,
        nstart = number_of_starts
      )

      tibble(
        nstart = number_of_starts,
        seed = seed_value,
        tot_withinss = fit$tot.withinss
      )
    })
  }
)

initialization_summary <- initialization_results %>%
  group_by(nstart) %>%
  summarise(
    minimum_wss = min(tot_withinss),
    median_wss = median(tot_withinss),
    maximum_wss = max(tot_withinss),
    distinct_objective_values = n_distinct(round(tot_withinss, 6)),
    .groups = "drop"
  )

initialization_summary

ggplot(
  initialization_results,
  aes(factor(nstart), tot_withinss)
) +
  geom_boxplot() +
  geom_jitter(width = 0.08, alpha = 0.4) +
  labs(
    title = "Sensitivity to Random Starting Values",
    x = "nstart",
    y = "Total within-cluster sum of squares"
  )
```

```{r}
nstart_1_range <- initialization_summary %>%
  filter(nstart == 1) %>%
  transmute(range = maximum_wss - minimum_wss) %>%
  pull(range)

nstart_20_range <- initialization_summary %>%
  filter(nstart == 20) %>%
  transmute(range = maximum_wss - minimum_wss) %>%
  pull(range)
```

Q: What is the impact of `nstart` on the final result?

A:

- Across these seeds, the range of total within-cluster sum of squares is `r round(nstart_1_range, 3)` for `nstart = 1` and `r round(nstart_20_range, 3)` for `nstart = 20`.
- Variation under `nstart = 1` indicates sensitivity to random initial centroids and convergence to different local solutions.
- A larger `nstart` substantially reduces the risk of retaining a poor local solution because the best objective value is selected from many attempts.
- `nstart` does not mathematically guarantee the global optimum, but it is inexpensive insurance for a dataset of this size. Setting a seed makes the selected result reproducible.

# Implications for Investment Analysis

Q: How could an investor use these clusters when investing in industry ETFs?

A:

- **Portfolio diversification:** Use the clusters to identify industry ETFs with similar underlying operating characteristics. Allocating across clusters may provide more meaningful diversification than simply holding several ETFs whose industries share the same revenue and cost dynamics.

- **Peer-relative analysis:** Compare an industry ETF with ETFs whose underlying industries belong to the same cluster. This may provide a more appropriate peer group than comparing all industry ETFs with one another.

- **Portfolio risk management:** Measure how much of the portfolio is exposed to each cluster and establish limits when several holdings represent similar economic profiles. Multiple ETF positions do not necessarily provide diversification if their underlying industries belong to the same cluster.

- **Stress testing:** Apply common revenue, cost, or margin scenarios to industries with similar growth dynamics. For example, industries in the high revenue and cost growth cluster may be especially sensitive to a scenario in which cost growth remains high while revenue growth slows.

- **Exception handling:** Treat influential observations such as `Industry07` or `Industry32` as candidates for individual analysis rather than allowing one unusual industry to determine the broader grouping of ETFs.

- **Portfolio monitoring:** Track changes in an industry's cluster membership or distance from its cluster centroid. A material change can serve as a prompt for further research, but it should not automatically trigger a trade.

- **Validation:** Test whether the clusters are stable over time and whether they are associated with relevant subsequent ETF outcomes, such as returns, volatility, drawdowns, or correlations. Until such relationships are established out of sample, the clusters should be interpreted as descriptive portfolio-construction and risk-management tools rather than validated trading signals.