---
title: "K-Means"
subtitle: "Machine Learning For Finance"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
---

# Learning Objectives

By the end of this lesson, students should be able to:

- Explain how K-means assigns observations to clusters and updates cluster centroids.
- Prepare a cross-sectional feature set for clustering, including checks for missing values, duplicate keys, unstable growth rates, scale differences, and influential observations.
- Explain why feature selection, standardization, random starting values, and the choice of $K$ can change the resulting partition.
- Use the elbow, gap-statistic, and silhouette approaches as complementary evidence when choosing $K$.
- Interpret clusters economically and describe how an investor might use—and validate—the resulting segmentation.

# Housekeeping

```{r}
#| warning: false

cat("\014")  # clear the console when run interactively
rm(list = ls())  # begin from a clean workspace

if (!requireNamespace("here", quietly = TRUE)) {
  stop("Install the 'here' package before rendering this document.")
}

suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Background

- You are an analyst at a systematic investment manager
- The asset manager is focused on fundamentals and operates primarily in ETFs
- Your portfolio manager is asking you to cluster industries by their revenue and cost characteristics.

Q: How might the Portfolio Manager convert these cluster into actionable insights for portfolio allocation and risk management strategies?

A:

Q: What is the unit that we ultimately want to cluster?

A:

# Data

## Acquisition

Task: Load the data dictionary from the `Data_Dictionary` sheet of `industry_fundamentals.xlsx`.

```{r}
data_dictionary <- readxl::read_excel(
  here("data", "?"),
  sheet = "?"
)

data_dictionary
```

Task: Before analyzing the observations, how do you expect industries to group based on size, growth, and profitability?

Task: Load the `Data` sheet and call the object `industry_fundamentals`.

```{r}
industry_fundamentals <- readxl::read_excel(
  here("data", "industry_fundamentals.xlsx"),
  sheet = "?"
)

dplyr::glimpse(industry_fundamentals)
```

## Validate the Data Structure

A clustering algorithm will return an answer even when the input has structural problems. We therefore verify the unit of observation and the variables before engineering features.

Task: Display the fields in the dataset.

```{r}
names(industry_fundamentals)
```

Task: Count rows, distinct industries, and distinct years.

```{r}
industry_fundamentals %>%
  summarise(
    rows = ?,
    industries = n_distinct(Industry),
    years = n_distinct(Year)
  )
```

Q: What is the unit of observation in the source data: industry, year, revenue, or something else?

A:

Task: Verify that each industry-year key is unique.

```{r}
duplicate_keys <- industry_fundamentals %>%
  dplyr::count(Industry, Year, name = "rows_per_key") %>%
  filter(rows_per_key > 1)

duplicate_keys
stopifnot(nrow(duplicate_keys) == 0)
```

Task: Verify that every industry appears in every year represented in the data.

```{r}
analysis_years <- sort(unique(industry_fundamentals$Year))
stopifnot(length(analysis_years) == 2)

year_1 <- analysis_years[1]
year_2 <- analysis_years[2]

incomplete_industries <- industry_fundamentals %>%
  dplyr::count(Industry, name = "observed_years") %>%
  filter(observed_years != length(analysis_years))

incomplete_industries
stopifnot(nrow(incomplete_industries) == 0)
```

## Cleaning and Accounting Checks

Task: Count missing values by field

```{r}
missing_by_field <- industry_fundamentals %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  ?(
    cols = everything(),
    names_to = "field",
    values_to = "missing_values"
  )

missing_by_field
stopifnot(sum(missing_by_field$missing_values) == 0)
```

Task: Verify the accounting identity $\text{Profit}=\text{Revenue}-\text{Operating Cost}$.

```{r}
accounting_check <- industry_fundamentals %>%
  summarise(
    maximum_absolute_error = max(
      abs(Profit - (Revenue - Operating_Cost)),
      na.rm = TRUE
    )
  )

accounting_check
```

Q: Why does this identity matter for clustering?

A:

## Engineering Growth Rates

Task: Compute simple growth rates for each industry from the first year to the second year.

```{r}
growth_rates <- industry_fundamentals %>%
  arrange(Industry, Year) %>%
  group_by(Industry) %>%
  summarise(
    Revenue_Year1 = first(Revenue),
    Revenue_Year2 = last(Revenue),
    Operating_Cost_Year1 = first(Operating_Cost),
    Operating_Cost_Year2 = last(Operating_Cost),
    Profit_Year1 = first(Profit),
    Profit_Year2 = last(Profit),
    .groups = "drop"
  ) %>%
  mutate(
    Revenue_Growth = ?,
    Cost_Growth = ?,
    Profit_Growth = ?
  )

head(growth_rates)
```

Task: Inspect the Year 1 denominators used to calculate growth.

```{r}
growth_denominator_check <- growth_rates %>%
  summarise(
    minimum_revenue_base = min(Revenue_Year1),
    minimum_cost_base = min(Operating_Cost_Year1),
    minimum_absolute_profit_base = min(abs(Profit_Year1)),
    nonpositive_profit_bases = sum(Profit_Year1 <= 0)
  )

growth_denominator_check

stopifnot(
  all(growth_rates$Revenue_Year1 != 0),
  all(growth_rates$Operating_Cost_Year1 != 0),
  all(growth_rates$Profit_Year1 != 0)
)
```

Q: Why should Profit Growth be interpreted more cautiously than Revenue Growth or Cost Growth?

A:

## Construct the Analysis Samples

K-means is sensitive to the sample used to estimate means, standard deviations, and centroids. We will standardize each feature within the cross-sectional sample actually being clustered.

Task: Construct full Year 1 levels sample and full industry growth sample, then create standardized versions of each.

```{r}
z_score <- function(x) {
  as.numeric(scale(x))
}

year1 <- industry_fundamentals %>%
  filter(Year == year_1)

year1_std_full <- year1 %>%
  mutate(
    Revenue_z = z_score(?),
    Operating_Cost_z = z_score(?),
    Profit_z = z_score(?)
  )

growth_rates_std_full <- growth_rates %>%
  mutate(
    Revenue_Growth_z = z_score(?),
    Cost_Growth_z = z_score(?),
    Profit_Growth_z = z_score(?)
  )
```

# Exploratory Data Analysis

We will focus on exploratory analysis that affects the clustering choices made later.

## Levels and Growth Distributions

Task: For Year 1, create side-by-side violin plots of Revenue, Operating Cost, and Profit in their raw dollar units.

```{r}
year1_long <- year1 %>%
  pivot_longer(
    cols = c(Revenue, Operating_Cost, Profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(year1_long, aes(x = metric, y = value)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.05, alpha = 0.45) +
  labs(
    title = "Year 1 Financial Levels",
    x = NULL,
    y = "$ millions"
  )
```

Task: Create side-by-side violin plots of Revenue Growth, Cost Growth, and Profit Growth.

```{r}
growth_long <- growth_rates %>%
  pivot_longer(
    cols = c(Revenue_Growth, Cost_Growth, Profit_Growth),
    names_to = "metric",
    values_to = "value"
  )

ggplot(growth_long, aes(x = metric, y = value)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.05, alpha = 0.45) +
  labs(
    title = "Growth Rates from Year 1 to Year 2",
    x = NULL,
    y = "Growth rate"
  )
```

## Identify Influential Observations

An observation can be unusual in the levels feature space but ordinary in the growth feature space, or vice versa.

Task: Identify observations at least three standard deviations from the mean in either feature space.

```{r}
level_outlier_flags <- year1_std_full %>%
  select(Industry, Revenue_z, Operating_Cost_z, Profit_z) %>%
  pivot_longer(
    cols = -Industry,
    names_to = "metric",
    values_to = "z_score"
  ) %>%
  filter(?) %>%
  arrange(desc(abs(z_score)))

growth_outlier_flags <- growth_rates_std_full %>%
  select(Industry, Revenue_Growth_z, Cost_Growth_z, Profit_Growth_z) %>%
  pivot_longer(
    cols = -Industry,
    names_to = "metric",
    values_to = "z_score"
  ) %>%
  filter(?) %>%
  arrange(desc(abs(z_score)))

level_outlier_flags
growth_outlier_flags
```

Task: Inspect the complete records for the two influential industries identified by the plots and z-scores.

```{r}
year1 %>%
  ?(Industry == "Industry32")

growth_rates %>%
  ?(Industry == "Industry07")
```

Q: What do the plots and diagnostic tables suggest? Are we looking at one outlier or more than one?

A:

Task: Build a levels sample excluding `Industry32` and a growth-rate sample excluding `Industry07`. Re-standardize within each trimmed sample.

```{r}
year1_trim <- year1 %>%
  filter(Industry ?) %>%
  mutate(
    Revenue_z = z_score(Revenue),
    Operating_Cost_z = z_score(Operating_Cost),
    Profit_z = z_score(Profit)
  )

growth_rates_trim <- growth_rates %>%
  filter(Industry ?) %>%
  mutate(
    Revenue_Growth_z = z_score(Revenue_Growth),
    Cost_Growth_z = z_score(Cost_Growth),
    Profit_Growth_z = z_score(Profit_Growth)
  )

nrow(year1_trim)
nrow(growth_rates_trim)
```

## Correlation and Outlier Influence

Correlation is unchanged by positive linear rescaling, so it is clearest to compute these correlations in the original growth-rate units.

```{r}
correlation_long <- function(correlation_matrix) {
  as.data.frame(as.table(correlation_matrix)) %>%
    as_tibble() %>%
    rename(
      variable_1 = Var1,
      variable_2 = Var2,
      correlation = Freq
    )
}

plot_correlation <- function(correlation_matrix, plot_title) {
  correlation_long(correlation_matrix) %>%
    ggplot(aes(variable_1, variable_2, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(
      low = "blue",
      high = "red",
      mid = "white",
      midpoint = 0,
      limits = c(-1, 1),
      name = "Correlation"
    ) +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme_minimal()
}
```

Task: Compute and visualize correlations using all industries.

```{r}
growth_cor_full <- growth_rates %>%
  select(Revenue_Growth, Cost_Growth, Profit_Growth) %>%
  ?(use = "complete.obs")

growth_cor_full
?(growth_cor_full, "Growth-Rate Correlations: All Industries")
```

Task: Recompute and visualize the correlations after excluding `Industry07`.

```{r}
growth_cor_trim <- growth_rates_trim %>%
  ?(Revenue_Growth, Cost_Growth, Profit_Growth) %>%
  cor(use = "complete.obs")

growth_cor_trim
plot_correlation(growth_cor_trim, "Growth-Rate Correlations: Industry07 Excluded")
```

Q: How does removing `Industry07` change the correlation structure?

A:

# K-Means

## Intuition and Algorithm

Q: How does K-means cluster observations?

A:

Q: Does K-means cluster observations or variables?

A:

## A Basic Fit

For the basic mechanics, use the trimmed Revenue Growth and Cost Growth sample. We set `nstart = 1` here so that the relationship between random initialization and reproducibility remains visible.

Task: Fit K-means with $K=3$.

```{r}
basic_features <- growth_rates_trim %>%
  select(Revenue_Growth, Cost_Growth)

set.seed(1234)
basic_kmeans <- ?(
  basic_features,
  centers = ?,
  nstart = 1
)
```

Task: Display each industry with its assigned cluster.

```{r}
basic_assignments <- tibble(
  Industry = growth_rates_trim$?,
  cluster = basic_kmeans$?
)

basic_assignments
```

Q: How should the cluster vector be interpreted?

A:

Task: Display the cluster centroids.

```{r}
basic_centers <- basic_kmeans$? %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cluster") %>%
  as_tibble()

basic_centers
```

Q: How should the centroids be interpreted?

A:

## The K-Means Objective and R Output

Define:

- $\mathbf{x}_i$: the $p \times 1$ feature vector for observation $i$.
- $\bar{\mathbf{x}}$: the overall sample mean vector.
- $\boldsymbol{\mu}_k$: the centroid of cluster $k$.
- $C_k$: the set of observations assigned to cluster $k$.
- $n_k$: the number of observations in cluster $k$.

The main quantities returned by `kmeans()` are:

### Total Sum of Squares: `totss`

$$
\mathrm{TSS}
=
\sum_{i=1}^{N}
\left\|\mathbf{x}_i-\bar{\mathbf{x}}\right\|^2.
$$

This measures total dispersion around the overall mean before using the cluster assignments.

### Within-Cluster Sum of Squares: `withinss`

For cluster $k$,

$$
\mathrm{WithinSS}_k
=
\sum_{i \in C_k}
\left\|\mathbf{x}_i-\boldsymbol{\mu}_k\right\|^2.
$$

This measures the total squared distance of the observations in cluster $k$ from that cluster’s centroid.

### Total Within-Cluster Sum of Squares: `tot.withinss`

$$
\mathrm{Total\ WithinSS}
=
\sum_{k=1}^{K}\mathrm{WithinSS}_k.
$$

This is the objective that K-means attempts to minimize.

### Between-Cluster Sum of Squares: `betweenss`

$$
\mathrm{BetweenSS}
=
\sum_{k=1}^{K}
 n_k
\left\|\boldsymbol{\mu}_k-\bar{\mathbf{x}}\right\|^2.
$$

This measures how far the cluster centroids are from the overall mean, weighted by cluster size.

The decomposition is

$$
\boxed{
\mathrm{TSS}
=
\mathrm{Total\ WithinSS}
+
\mathrm{BetweenSS}
}.
$$

Task: Display the fit statistics and verify the decomposition.

```{r}
basic_fit_statistics <- tibble(
  totss = basic_kmeans$?,
  tot_withinss = basic_kmeans$tot.withinss,
  betweenss = basic_kmeans$betweenss,
  decomposition_error = basic_kmeans$totss -
    basic_kmeans$tot.withinss -
    basic_kmeans$betweenss,
  between_share = basic_kmeans$betweenss / basic_kmeans$totss
)

basic_fit_statistics
```

Task: Compare cluster size, `withinss`, and `withinss` per observation.

```{r}
basic_cluster_statistics <- tibble(
  cluster = seq_along(basic_kmeans$size),
  size = basic_kmeans$size,
  withinss = basic_kmeans$withinss,
  withinss_per_observation = basic_kmeans$withinss / basic_kmeans$size
)

basic_cluster_statistics
```

Q: Are these sums of squares best interpreted in relative or absolute terms?

A:

## Random Initialization and `set.seed()`

K-means can converge to different local solutions from different random starting centroids. In addition, cluster numbers can be permuted even when the underlying partition is the same.

Task: Run the same one-start specification twice without resetting the seed, then compare both the objective value and the partition. Also compute the "adjustedRandIndex".

Note: The Adjusted Rand Index (ARI) measures how similar two cluster assignments are while ignoring the arbitrary numeric cluster labels, with 1 indicating identical clustering and values near 0 indicating agreement no better than chance.

```{r}
set.seed(9001)
km_random_1 <- kmeans(basic_features, centers = 3, nstart = 1)
km_random_2 <- kmeans(basic_features, centers = 3, nstart = 1)

c(
  run_1_tot_withinss = km_random_1$tot.withinss,
  run_2_tot_withinss = km_random_2$tot.withinss
)

mclust::adjustedRandIndex(
  km_random_1$cluster,
  km_random_2$cluster
)
```

Q: Why might the two results differ even though the model specification did not change?

A:

Task: Reset to the same seed before each run and verify exact reproducibility.

```{r}
set.seed(1234)
km_seed_1 <- kmeans(basic_features, centers = 3, nstart = 1)

set.seed(1234)
km_seed_2 <- kmeans(basic_features, centers = 3, nstart = 1)

identical(km_seed_1$cluster, km_seed_2$cluster)
identical(km_seed_1$centers, km_seed_2$centers)
```

Q: What does `set.seed()` accomplish?

A:

# Why Feature Redundancy Matters

This section is a self-contained pedagogical demonstration built on the **Year 1 levels** (Revenue, Operating Cost, and Profit). Its purpose is to show that adding a redundant, exactly-determined feature can still change a K-means partition by reweighting the distance geometry. The three-variable levels specification used here is **not** the specification used for the main clustering analysis; the main analysis clusters industries on two growth features only (standardized Revenue Growth and Cost Growth), as set up in the "Altering the Number of Clusters" section below.

Q: Suppose we cluster standardized Year 1 Revenue and Operating Cost. What do you expect to happen if standardized Profit is added?

A:

Task: Verify the accounting identity in the trimmed Year 1 sample.

```{r}
year1_trim %>%
  summarise(
    maximum_absolute_error = max(
      abs(Profit - ?)
    )
  )
```

Task: Fit $K=2$ using standardized Revenue and Operating Cost.

```{r}
set.seed(1234)
km_2feat <- kmeans(
  year1_trim %>% select(Revenue_z, Operating_Cost_z),
  ? = 2,
  nstart = 20
)
```

Task: Fit $K=2$ using standardized Revenue, Operating Cost, and Profit.

```{r}
set.seed(1234)
km_all3 <- kmeans(
  year1_trim %>% select(?),
  centers = 2,
  nstart = 20
)
```

Task: Compute cluster profiles for both specifications.

```{r}
year1_feature_comparison <- year1_trim %>%
  mutate(
    cluster_2feat = km_2feat$cluster,
    cluster_all3 = km_all3$cluster
  )

year1_feature_comparison %>%
  group_by(cluster_2feat) %>%
  summarise(
    across(c(Revenue, Operating_Cost, Profit), mean),
    n = n(),
    .groups = "drop"
  )

year1_feature_comparison %>%
  group_by(cluster_all3) %>%
  summarise(
    across(c(Revenue, Operating_Cost, Profit), mean),
    n = n(),
    .groups = "drop"
  )
```

Cluster labels are arbitrary, so they must be aligned before identifying observations that “switch” numeric labels. The helper below is specific to a two-cluster comparison.

```{r}
align_two_cluster_labels <- function(reference, candidate) {
  same_label_agreement <- sum(reference == candidate)
  swapped_label_agreement <- sum(reference == (3 - candidate))

  if (swapped_label_agreement > same_label_agreement) {
    3 - candidate
  } else {
    candidate
  }
}

year1_feature_comparison <- year1_feature_comparison %>%
  mutate(
    cluster_2feat_aligned = align_two_cluster_labels(
      reference = cluster_all3,
      candidate = cluster_2feat
    )
  )

feature_crosstab <- table(
  All3 = year1_feature_comparison$cluster_all3,
  TwoFeat = year1_feature_comparison$cluster_2feat
)

feature_switches <- year1_feature_comparison %>%
  filter(cluster_all3 != cluster_2feat_aligned) %>%
  select(Industry, cluster_all3, cluster_2feat_aligned)

feature_ari <- mclust::adjustedRandIndex(
  year1_feature_comparison$cluster_all3,
  year1_feature_comparison$cluster_2feat
)

feature_switch_text <- if (nrow(feature_switches) == 0) {
  "none"
} else {
  paste(feature_switches$Industry, collapse = ", ")
}

feature_crosstab
feature_switches
feature_ari
```

Q: How much did cluster membership change after Profit was added?

A:

Q: If Profit adds no new independent information, why can it change the clusters?

A:

$$
d^2
=
(\Delta z_{Revenue})^2
+
(\Delta z_{Cost})^2
+
(\Delta z_{Profit})^2,
$$

whereas the two-feature objective omits the final term. The data still lie on a two-dimensional accounting plane, but distances along directions associated with Profit receive additional weight. A borderline observation can therefore move to another nearest centroid.

# Why Standardization Matters

K-means is not invariant to rescaling. Multiplying one feature by 100 multiplies its contribution to squared Euclidean distance by $100^2$.

## Raw Versus Standardized Features in This Dataset

Task: Compare $K=2$ fits for raw and standardized Year 1 levels.

```{r}
set.seed(4821)
km_raw_levels <- ?(
  year1_trim %>% ?(Revenue, Operating_Cost),
  centers = 2,
  nstart = 20
)

set.seed(4821)
km_std_levels <- ?(
  year1_trim %>% ?(Revenue_z, Operating_Cost_z),
  centers = 2,
  nstart = 20
)
```

Task: Compare $K=2$ fits for raw and standardized growth rates.

```{r}
set.seed(4821)
km_raw_growth <- kmeans(
  growth_rates_trim %>% select(Revenue_Growth, Cost_Growth),
  centers = ?,
  nstart = 20
)

set.seed(4821)
km_std_growth <- kmeans(
  growth_rates_trim %>% select(Revenue_Growth_z, Cost_Growth_z),
  centers = ?,
  nstart = 20
)
```

Task: Quantify agreement and compare the raw feature standard deviations.

```{r}
scale_agreement <- tibble(
  comparison = c("Year 1 levels", "Growth rates"),
  adjusted_rand_index = c(
    mclust::adjustedRandIndex(
      km_raw_levels$cluster,
      km_std_levels$cluster
    ),
    mclust::adjustedRandIndex(
      km_raw_growth$cluster,
      km_std_growth$cluster
    )
  )
)

feature_scale_summary <- bind_rows(
  year1_trim %>%
    summarise(
      Revenue = sd(Revenue),
      Operating_Cost = sd(Operating_Cost)
    ) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "standard_deviation") %>%
    mutate(sample = "Year 1 levels"),
  growth_rates_trim %>%
    summarise(
      Revenue_Growth = sd(Revenue_Growth),
      Cost_Growth = sd(Cost_Growth)
    ) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "standard_deviation") %>%
    mutate(sample = "Growth rates")
)

scale_agreement
feature_scale_summary
```

Q: Did standardization materially change these particular two-feature partitions?

A:

## Mixed Units

To make the scaling issue explicit, combine a dollar level with a growth rate. This is an illustrative geometry exercise, not necessarily the preferred economic specification.

Task: Cluster industries on Year 1 Revenue and Revenue Growth before and after standardization.

```{r}
mixed_scale_features <- year1_trim %>%
  select(Industry, Revenue) %>%
  inner_join(
    growth_rates_trim %>% select(Industry, Revenue_Growth),
    by = "Industry"
  ) %>%
  mutate(
    Revenue_z = z_score(Revenue),
    Revenue_Growth_z = z_score(Revenue_Growth)
  )

set.seed(4821)
km_mixed_raw <- kmeans(
  mixed_scale_features %>% select(Revenue, Revenue_Growth),
  centers = 3,
  nstart = 20
)

set.seed(4821)
km_mixed_std <- kmeans(
  mixed_scale_features %>% select(Revenue_z, Revenue_Growth_z),
  centers = 3,
  nstart = 20
)

mixed_scale_ari <- mclust::adjustedRandIndex(
  km_mixed_raw$cluster,
  km_mixed_std$cluster
)

mixed_scale_ari
```

Task: Compare cluster profiles in the original units.

```{r}
mixed_scale_features %>%
  mutate(
    raw_cluster = km_mixed_raw$cluster,
    standardized_cluster = km_mixed_std$cluster
  ) %>%
  group_by(raw_cluster) %>%
  summarise(
    Revenue = mean(Revenue),
    Revenue_Growth = mean(Revenue_Growth),
    n = n(),
    .groups = "drop"
  )

mixed_scale_features %>%
  mutate(
    raw_cluster = km_mixed_raw$cluster,
    standardized_cluster = km_mixed_std$cluster
  ) %>%
  group_by(standardized_cluster) %>%
  summarise(
    Revenue = mean(Revenue),
    Revenue_Growth = mean(Revenue_Growth),
    n = n(),
    .groups = "drop"
  )
```

Q: Why is standardization more consequential in the mixed-unit example?

A:

# Altering the Number of Clusters

We now return to the trimmed, standardized growth sample. From this point forward, the main clustering analysis uses **exactly two features**: standardized Revenue Growth and standardized Cost Growth. Profit Growth is deliberately excluded—it carries no independent accounting information beyond Revenue and Operating Cost, and its percentage-growth form is unstable when the initial profit base is small. The feature matrix is therefore genuinely two-dimensional, and we name it `X_growth_2feat` to make its composition explicit.

```{r}
growth_features <- growth_rates_trim %>%
  select(Industry, Revenue_Growth_z, Cost_Growth_z)

X_growth_2feat <- growth_features %>%
  select(-Industry) %>%
  as.matrix()

fit_kmeans <- function(k, seed = 4821, nstart = 20) {
  set.seed(seed)
  kmeans(X_growth_2feat, centers = k, nstart = nstart)
}

growth_scatter <- function(fit) {
  plot_data <- growth_features %>%
    mutate(cluster = factor(fit$cluster))

  ggplot(
    plot_data,
    aes(Revenue_Growth_z, Cost_Growth_z, color = cluster)
  ) +
    geom_point(size = 2) +
    labs(
      x = "Revenue Growth (z)",
      y = "Cost Growth (z)",
      color = "Cluster"
    )
}

cluster_profile <- function(fit) {
  growth_rates_trim %>%
    mutate(cluster = fit$cluster) %>%
    group_by(cluster) %>%
    summarise(
      across(c(Revenue_Growth, Cost_Growth), mean),
      n = n(),
      .groups = "drop"
    ) %>%
    arrange(Revenue_Growth)
}
```

## $K=1$

Task: Fit and visualize one cluster.

```{r}
km_k1 <- fit_kmeans(?)
growth_scatter(km_k1)
cluster_profile(km_k1)
```

## $K=2$

Task: Repeat for two clusters.

```{r}
km_k2 <- fit_kmeans(?)
growth_scatter(km_k2)
cluster_profile(km_k2)
```

## $K=3$

Task: Repeat for three clusters.

```{r}
km_k3 <- fit_kmeans(?)
growth_scatter(km_k3)
cluster_profile(km_k3)
```

## $K=4$

Task: Repeat for four clusters.

```{r}
km_k4 <- fit_kmeans(?)
growth_scatter(km_k4)
cluster_profile(km_k4)
```

Numeric cluster labels cannot be tracked meaningfully across different values of $K$. Instead, track both the assignment and the centroid profile attached to that assignment.

```{r}
membership_with_centroid <- function(fit, k_label) {
  centers <- fit$centers %>%
    as.data.frame() %>%
    tibble::rownames_to_column("cluster") %>%
    mutate(cluster = as.integer(cluster)) %>%
    as_tibble()

  tibble(
    Industry = growth_features$Industry,
    cluster = fit$cluster
  ) %>%
    left_join(centers, by = "cluster") %>%
    mutate(K = k_label, .before = Industry)
}

tracked_industries <- bind_rows(
  membership_with_centroid(km_k1, "K = 1"),
  membership_with_centroid(km_k2, "K = 2"),
  membership_with_centroid(km_k3, "K = 3"),
  membership_with_centroid(km_k4, "K = 4")
) %>%
  filter(Industry %in% c("Industry01", "Industry44")) %>%
  arrange(Industry, K)

tracked_industries
```

Q: What happens as $K$ increases?

A:

# Choosing $K$

No single diagnostic mechanically determines the economically “correct” number of clusters. We combine statistical evidence with interpretability, stability, and the intended decision use.

## Elbow Method

The elbow method examines how quickly total within-cluster sum of squares falls as $K$ increases. We look for a point after which additional clusters provide much smaller improvements.

Task: Construct an elbow plot for $K=1,\ldots,10$.

```{r}
wss <- purrr::map_dbl(1:10, function(k) {
  fit_kmeans(k, seed = 4821, nstart = 20)$tot.withinss
})

elbow_df <- tibble(
  k = 1:10,
  wss = wss
) %>%
  mutate(
    reduction = lag(wss) - wss,
    proportional_reduction = reduction / lag(wss)
  )

elbow_df

ggplot(elbow_df, aes(k, wss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Plot",
    x = "Number of clusters (K)",
    y = "Total within-cluster sum of squares"
  )
```

Q: How many clusters does the elbow plot suggest?

A:

## Gap Statistic

The gap statistic compares the observed within-cluster dispersion with dispersion expected from reference datasets that contain no comparable cluster structure.

From the gap plot, we look for the first K where the gap curve "levels off".

For a more rigorous approach, we can follow the Tibs2001SEMax rule: Choose the smallest K for which adding another cluster doesn’t improve the Gap statistic by more than one standard error.

Task: Construct a gap-statistic plot for $K=1,\ldots,10$.

```{r}
set.seed(4821)
gap_stat <- cluster::clusGap(
  X_growth_2feat,
  FUNcluster = function(x, k) {
    kmeans(x, centers = k, nstart = 20)
  },
  K.max = 10,
  B = 100
)

gap_df <- gap_stat$Tab %>%
  as.data.frame() %>%
  tibble::rownames_to_column("k") %>%
  as_tibble() %>%
  mutate(k = as.integer(k))

gap_k <- cluster::maxSE(
  gap_stat$Tab[, "gap"],
  gap_stat$Tab[, "SE.sim"],
  method = "Tibs2001SEmax"
)

gap_df

ggplot(gap_df, aes(?)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(ymin = gap - SE.sim, ymax = gap + SE.sim),
    width = 0.2
  ) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Gap Statistic",
    x = "Number of clusters (K)",
    y = "Gap"
  )

gap_k
```

Q: How many clusters does the gap statistic suggest?

A:

## Silhouette Width

For each observation, silhouette width compares cohesion with its assigned cluster against separation from the nearest alternative cluster. Values near 1 indicate strong assignment, values near 0 indicate a boundary observation, and negative values suggest possible misassignment.

Task: Compute average silhouette width for $K=2,\ldots,10$.

```{r}
growth_distance <- dist(X_growth_2feat)

silhouette_scores <- purrr::map_dfr(2:10, function(k) {
  fit <- fit_kmeans(k, seed = 4821, nstart = 20)
  sil <- cluster::silhouette(fit$cluster, growth_distance)

  tibble(
    k = k,
    average_silhouette_width = mean(sil[, "sil_width"])
  )
})

silhouette_k <- silhouette_scores %>%
  slice_max(average_silhouette_width, n = 1, with_ties = FALSE) %>%
  pull(k)

silhouette_scores

ggplot(
  silhouette_scores,
  aes(k, average_silhouette_width)
) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Average Silhouette Width by K",
    x = "Number of clusters (K)",
    y = "Average silhouette width"
  )

silhouette_k
```

Task: Inspect the detailed silhouette plot for the best-scoring $K$.

```{r}
km_silhouette_best <- fit_kmeans(
  silhouette_k,
  seed = 4821,
  nstart = 20
)

silhouette_best <- cluster::silhouette(
  km_silhouette_best$cluster,
  growth_distance
)

plot(
  silhouette_best,
  main = paste("Silhouette Plot, K =", silhouette_k)
)
```

Q: Which cluster appears most internally coherent and well separated from the other clusters?

A:

Q: Which cluster appears least clearly separated?

A:

Task: Fit K=3, inspect its silhouette plot.

```{r}
km_silhouette_alt <- fit_kmeans(
  3,
  seed = 4821,
  nstart = 20
)

silhouette_alt <- cluster::silhouette(
  km_silhouette_alt$cluster,
  growth_distance
)

plot(
  silhouette_alt,
  main = paste("Silhouette Plot, K =", 3)
)

km_2 <- ?(?, seed = 4821, nstart = 20)
km_3 <- ?(?, seed = 4821, nstart = 20)
```

Task: Create a cross-tabulation of the K=2 and K=3 assignments to determine how the original clusters were divided.

```{r}
table(
  K2 = km_2$cluster,
  K3 = km_3$cluster
)
```

Q: How read this cross tab? What's the implication?

A:

Q: Does allowing K-means to form a third cluster improve the quality of the clustering?

A:

Q: Why can tot.withinss improve when moving from K=2 to K=3, while average silhouette width gets worse?

A:

Q: How many clusters does the silhouette analysis suggest?

A:

## Synthesis

Task: Summarize the three diagnostics.

```{r}
selection_summary <- tibble(
  method = c("Elbow", "Gap statistic", "Silhouette"),
  suggested_k = c(
    "2; pronounced first drop",
    as.character(gap_k),
    as.character(silhouette_k)
  )
)

selection_summary
```

Q: How many clusters should we use?

A:

# Interpreting the Final Clusters

Cluster numbers are arbitrary. Economic names are assigned after reviewing the cluster profiles.

```{r}
chosen_k <- 2L

set.seed(4821)
km_final <- kmeans(
  X_growth_2feat,
  centers = chosen_k,
  nstart = 20
)

growth_clustered <- growth_rates_trim %>%
  mutate(cluster = km_final$cluster)

cluster_profiles <- growth_clustered %>%
  group_by(cluster) %>%
  summarise(
    Revenue_Growth = mean(Revenue_Growth),
    Cost_Growth = mean(Cost_Growth),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    average_operating_growth = (Revenue_Growth + Cost_Growth) / 2
  )

flat_cluster <- cluster_profiles %>%
  slice_min(abs(average_operating_growth), n = 1, with_ties = FALSE) %>%
  pull(cluster)

high_growth_cluster <- setdiff(cluster_profiles$cluster, flat_cluster)

cluster_name_map <- tibble(
  cluster = c(
    flat_cluster,
    high_growth_cluster
  ),
  cluster_name = c(
    "Mature / flat growth",
    "High revenue and cost growth"
  )
)

growth_clustered <- growth_clustered %>%
  left_join(cluster_name_map, by = "cluster")

cluster_profiles <- cluster_profiles %>%
  left_join(cluster_name_map, by = "cluster") %>%
  arrange(average_operating_growth)
```

Task: Display cluster sizes and average growth rates.

```{r}
cluster_profiles %>%
  select(
    cluster,
    cluster_name,
    n,
    Revenue_Growth,
    Cost_Growth
  )
```

Task: Visualize the clusters in the Revenue Growth–Cost Growth plane.

```{r}
ggplot(
  growth_clustered,
  aes(
    Revenue_Growth,
    Cost_Growth,
    color = cluster_name
  )
) +
  geom_point(size = 2) +
  labs(
    title = "Growth-Rate Clusters",
    x = "Revenue Growth",
    y = "Cost Growth",
    color = "Economic label"
  )
```

Because the clustering uses exactly these two features, this scatter plot is a direct and complete visualization of the feature space the algorithm actually used—not a projection. Every distance that determined the cluster assignments is visible in this plane.

Task: Compare Year 1 Revenue and Operating Cost while retaining the growth-based cluster colors.

```{r}
level_view <- year1 %>%
  select(Industry, Revenue, Operating_Cost) %>%
  inner_join(
    growth_clustered %>%
      select(Industry, cluster, cluster_name),
    by = "Industry"
  )

ggplot(
  level_view,
  aes(
    Revenue,
    Operating_Cost,
    color = cluster_name
  )
) +
  geom_point(size = 2) +
  labs(
    title = "Year 1 Size, Colored by Growth-Rate Cluster",
    x = "Revenue",
    y = "Operating Cost",
    color = "Economic label"
  )
```

Task: Display cluster membership.

```{r}
growth_clustered %>%
  select(Industry, cluster, cluster_name) %>%
  arrange(cluster_name, Industry)
```

Q: What is the economic meaning of the final clusters?

A:

# Sensitivity and Stability

## Outlier in Levels: Add `Industry32` Back

Task: Repeat the $K=2$ feature-redundancy comparison using the full, Year 1-standardized sample.

```{r}
set.seed(4821)
km_all3_full <- kmeans(
  ? %>% select(Revenue_z, Operating_Cost_z, Profit_z),
  centers = 2,
  nstart = 20
)

set.seed(4821)
km_2feat_full <- kmeans(
  year1_std_full %>% select(Revenue_z, Operating_Cost_z),
  centers = 2,
  nstart = 20
)

full_level_comparison <- year1_std_full %>%
  transmute(
    Industry,
    cluster_all3 = km_all3_full$cluster,
    cluster_2feat = km_2feat_full$cluster
  )

full_level_ari <- mclust::adjustedRandIndex(
  full_level_comparison$cluster_all3,
  full_level_comparison$cluster_2feat
)

table(
  All3 = full_level_comparison$cluster_all3,
  TwoFeat = full_level_comparison$cluster_2feat
)

full_level_comparison %>% 
    dplyr::count(cluster_all3)
full_level_ari
```

Q: What changes once `Industry32` is added back?

A:

## Outlier in Growth Rates: Add `Industry07` Back

Task: Repeat the $K=2$ growth clustering with all industries and compare the common observations with the trimmed result.

```{r}
X_growth_2feat_full <- ? %>%
  select(Revenue_Growth_z, Cost_Growth_z) %>%
  as.matrix()

set.seed(4821)
km_full_k2 <- kmeans(
  X_growth_2feat_full,
  centers = chosen_k,
  nstart = 20
)

full_growth_assignments <- tibble(
  Industry = growth_rates_std_full$Industry,
  cluster_full = km_full_k2$cluster
)

full_growth_profiles <- growth_rates %>%
  left_join(full_growth_assignments, by = "Industry") %>%
  group_by(cluster_full) %>%
  summarise(
    across(c(Revenue_Growth, Cost_Growth), mean),
    n = n(),
    .groups = "drop"
  )

compare_growth <- growth_clustered %>%
  select(Industry, cluster_trimmed = cluster) %>%
  left_join(full_growth_assignments, by = "Industry")

growth_outlier_ari <- mclust::adjustedRandIndex(
  compare_growth$cluster_trimmed,
  compare_growth$cluster_full
)

full_growth_profiles

table(
  trimmed = compare_growth$cluster_trimmed,
  full = compare_growth$cluster_full
)

growth_outlier_ari
```

Q: What changes once `Industry07` is added back?

A:

Q: How did adding the outlier affect the K=2 solution?

A:

## Starting Values and `nstart`

`nstart` repeats K-means from multiple random initial centroid sets and retains the fit with the smallest total within-cluster sum of squares.

Task: Compare `nstart = 1` and `nstart = 20` across 50 random seeds.

```{r}
initialization_results <- purrr::map_dfr(
  c(1L, 20L),
  function(number_of_starts) {
    purrr::map_dfr(?, function(seed_value) {
      set.seed(seed_value)

      fit <- kmeans(
        X_growth_2feat,
        centers = chosen_k,
        nstart = number_of_starts
      )

      tibble(
        nstart = number_of_starts,
        seed = seed_value,
        tot_withinss = fit$tot.withinss
      )
    })
  }
)

initialization_summary <- initialization_results %>%
  group_by(nstart) %>%
  summarise(
    minimum_wss = min(tot_withinss),
    median_wss = median(tot_withinss),
    maximum_wss = max(tot_withinss),
    distinct_objective_values = n_distinct(round(tot_withinss, 6)),
    .groups = "drop"
  )

initialization_summary

ggplot(
  initialization_results,
  aes(factor(nstart), tot_withinss)
) +
  geom_boxplot() +
  geom_jitter(width = 0.08, alpha = 0.4) +
  labs(
    title = "Sensitivity to Random Starting Values",
    x = "nstart",
    y = "Total within-cluster sum of squares"
  )
```

```{r}
nstart_1_range <- initialization_summary %>%
  filter(nstart == 1) %>%
  transmute(range = maximum_wss - minimum_wss) %>%
  pull(range)

nstart_20_range <- initialization_summary %>%
  filter(nstart == 20) %>%
  transmute(range = maximum_wss - minimum_wss) %>%
  pull(range)
```

Q: What is the impact of `nstart` on the final result?

A:

# Implications for Investment Analysis

Q: How could an investor use these clusters when investing in industry ETFs?

A: