---
title: "The U.S. Listed Equity Universe Versus the S&P 500"
subtitle: "A reproducible exploratory data analysis of scale, size, liquidity, valuation, and sector concentration"
author: "Author Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    code-fold: true
    code-summary: "Show R code"
    code-tools: true
    code-overflow: wrap
    df-print: paged
    theme: cosmo
    fig-responsive: true
    embed-resources: true
execute:
  echo: true
  warning: true
  message: false
  error: false
  freeze: false
knitr:
  opts_chunk:
    fig.align: center
    fig.width: 11
    fig.height: 6.5
    dpi: 160
---

## Purpose and analytical scope

This report compares the current U.S. listed-security universe on NASDAQ, NYSE, and AMEX with the current S&P 500 benchmark. It uses two related samples:

1. **Full listed-security universe:** all usable exchange-directory records, including common stocks, ADRs, funds, preferred securities, warrants, rights, units, and other listings. This sample is used in Module 1.
2. **Operating-company equity universe:** securities heuristically classified as common equity or ADRs, reduced to one record per Yahoo-compatible symbol, with valid Yahoo market data, positive market capitalization and price, and populated sector and industry fields. This sample is used in Modules 2 through 4.

The security-type classification is heuristic. Missing fundamental fields are never replaced with fabricated values. The report preserves missing values, reports coverage explicitly, and continues when individual remote requests fail.

Data sources are the NASDAQ exchange screener accessed through `tidyquant::tq_exchange()`, State Street holdings accessed through `tidyquant::tq_index("SP500")`, and Yahoo Finance market data accessed through `quantmod` and `tidyquant`.

## Configuration and setup

```{r}
#| label: setup
#| include: true

required_packages <- c(
  "tidyverse",
  "tidyquant",
  "quantmod",
  "dplyr",
  "ggplot2",
  "patchwork",
  "scales",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    paste0(
      "Install the following required packages before rendering this document: ",
      paste(missing_packages, collapse = ", ")
    ),
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(tidyquant)
  library(quantmod)
  library(patchwork)
  library(scales)
})

options(
  scipen = 999,
  dplyr.summarise.inform = FALSE,
  width = 120
)

# ---- User-adjustable configuration -----------------------------------------
analysis_date <- Sys.Date()
lookback_calendar_days <- 60L
liquidity_trading_days <- 30L
minimum_valid_trading_days <- 20L
cache_directory <- "data_cache"
refresh_data <- FALSE # FALSE = use cache when available; TRUE = force new downloads
# On the first run, data are downloaded automatically because no cache exists.
# On subsequent runs, FALSE reuses the existing cache.
quote_batch_size <- 150L
price_batch_size <- 50L
max_retry_attempts <- 3L
max_recursive_split_depth <- 2L
retry_base_delay_seconds <- 1
pause_between_batches_seconds <- 0.35

# quantmod's current Yahoo implementation accepts fewer than 100 symbols per
# direct request. The configured quote_batch_size remains the user-facing
# maximum, while this provider cap prevents oversized direct calls.
provider_quote_limit <- 90L

dir.create(cache_directory, recursive = TRUE, showWarnings = FALSE)

price_from_date <- analysis_date - lookback_calendar_days
price_to_date <- analysis_date + 1L

benchmark_colors <- c(
  "S&P 500" = "#123B66",
  "Broader Market" = "#667480"
)

sector_measure_colors <- c(
  "S&P 500 constituent weight" = "#123B66",
  "Broad-universe ticker-count share" = "#667480",
  "Broad-universe market-cap share" = "#A7AFB5"
)

benchmark_levels <- c("Broader Market", "S&P 500")

report_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10.5, margin = margin(b = 8)),
    plot.caption = element_text(size = 8.5, color = "grey35", hjust = 0),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(8, 12, 8, 8)
  )

theme_set(report_theme)

# ---- General helpers --------------------------------------------------------
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0L) y else x
}

snake_names <- function(x) {
  x |>
    stringr::str_replace_all("([a-z0-9])([A-Z])", "\\1_\\2") |>
    stringr::str_replace_all("[^A-Za-z0-9]+", "_") |>
    stringr::str_replace_all("^_+|_+$", "") |>
    stringr::str_to_lower()
}

clean_text_na <- function(x) {
  y <- as.character(x)
  y <- stringr::str_squish(y)
  missing_tokens <- c("", "NA", "N/A", "NONE", "NULL", "NAN", "--", "-")
  y[stringr::str_to_upper(y) %in% missing_tokens] <- NA_character_
  y
}

safe_numeric <- function(x) {
  if (is.numeric(x)) {
    return(as.numeric(x))
  }

  y <- clean_text_na(x)
  y <- stringr::str_replace_all(y, ",", "")
  y <- stringr::str_replace_all(y, "\\$", "")
  y <- stringr::str_replace_all(y, "%", "")
  suppressWarnings(as.numeric(y))
}

positive_or_na <- function(x) {
  y <- safe_numeric(x)
  y[!is.finite(y) | y <= 0] <- NA_real_
  y
}

nonnegative_or_na <- function(x) {
  y <- safe_numeric(x)
  y[!is.finite(y) | y < 0] <- NA_real_
  y
}

ensure_columns <- function(data, columns) {
  missing_columns <- setdiff(columns, names(data))
  for (column in missing_columns) {
    data[[column]] <- rep(NA, nrow(data))
  }
  data
}

first_existing_column <- function(data, candidates, required = TRUE) {
  hit <- intersect(candidates, names(data))
  if (length(hit) > 0L) {
    return(hit[[1]])
  }

  if (required) {
    stop(
      paste0(
        "None of the required columns were found: ",
        paste(candidates, collapse = ", ")
      ),
      call. = FALSE
    )
  }

  NULL
}

normalize_yahoo_symbol <- function(x) {
  y <- clean_text_na(x)
  y <- stringr::str_to_upper(y)
  y <- stringr::str_replace_all(y, "\\s+", "")

  # Common preferred-share notations used by U.S. listing directories.
  y <- stringr::str_replace(y, "^([A-Z0-9]+)\\^([A-Z0-9]+)$", "\\1-P\\2")
  y <- stringr::str_replace(y, "^([A-Z0-9]+)[.]PR([A-Z0-9]+)$", "\\1-P\\2")
  y <- stringr::str_replace(y, "^([A-Z0-9]+)-PR([A-Z0-9]+)$", "\\1-P\\2")

  # Yahoo generally uses dashes for class-share separators.
  y <- stringr::str_replace_all(y, "[./]", "-")
  y <- stringr::str_replace_all(y, "\\$", "-P")
  y <- stringr::str_replace_all(y, "[^A-Z0-9=^-]", "")
  y[y == ""] <- NA_character_
  y
}

standardize_sector <- function(x) {
  original <- clean_text_na(x)
  y <- stringr::str_to_lower(original)

  dplyr::case_when(
    is.na(y) ~ NA_character_,
    stringr::str_detect(y, "information technology|technology") ~ "Information Technology",
    stringr::str_detect(y, "health care|healthcare|health") ~ "Health Care",
    stringr::str_detect(y, "financial services|financials|finance") ~ "Financials",
    stringr::str_detect(y, "consumer discretionary|consumer cyclical|consumer services") ~ "Consumer Discretionary",
    stringr::str_detect(y, "consumer staples|consumer defensive|consumer non-durables") ~ "Consumer Staples",
    stringr::str_detect(y, "communication services|telecommunication") ~ "Communication Services",
    stringr::str_detect(y, "basic materials|materials") ~ "Materials",
    stringr::str_detect(y, "industrial") ~ "Industrials",
    stringr::str_detect(y, "real estate") ~ "Real Estate",
    stringr::str_detect(y, "utilit") ~ "Utilities",
    stringr::str_detect(y, "energy") ~ "Energy",
    TRUE ~ stringr::str_to_title(original)
  )
}

classify_security_type <- function(company, symbol) {
  company_upper <- stringr::str_to_upper(dplyr::coalesce(clean_text_na(company), ""))
  symbol_upper <- stringr::str_to_upper(dplyr::coalesce(clean_text_na(symbol), ""))

  is_warrant_or_right <-
    stringr::str_detect(company_upper, "\\bWARRANTS?\\b|\\bRIGHTS?\\b") |
    stringr::str_detect(symbol_upper, "[-.](WS|WT|RT)$")

  is_adr <-
    stringr::str_detect(
      company_upper,
      "AMERICAN DEPOSITARY|DEPOSITARY RECEIPT|\\bADR\\b|\\bADS\\b"
    ) &
    !stringr::str_detect(company_upper, "PREFERRED")

  is_preferred <-
    stringr::str_detect(
      company_upper,
      "PREFERRED|PREFERENCE SHARE|DEPOSITARY SHARES.*PREFERRED|\\bPFD\\b"
    ) |
    stringr::str_detect(symbol_upper, "\\^|[-.]P[A-Z0-9]*$")

  is_etf_or_etn <- stringr::str_detect(
    company_upper,
    paste0(
      "\\bETF\\b|\\bETN\\b|EXCHANGE[- ]TRADED|",
      "ISHARES|SPDR|PROSHARES|DIREXION|GLOBAL X|WISDOMTREE|",
      "VANGUARD.*INDEX|INVESCO.*ETF|FIRST TRUST.*ETF"
    )
  )

  is_unit_or_spac <-
    stringr::str_detect(
      company_upper,
      "\\bUNITS?\\b|ACQUISITION CORP|BLANK CHECK|SPECIAL PURPOSE ACQUISITION|\\bSPAC\\b"
    ) |
    (
      stringr::str_detect(symbol_upper, "U$") &
      stringr::str_detect(company_upper, "ACQUISITION|HOLDINGS")
    )

  is_fund <- stringr::str_detect(
    company_upper,
    paste0(
      "CLOSED[- ]END|\\bFUND\\b|PORTFOLIO|MUTUAL FUND|",
      "INCOME TRUST|INVESTMENT TRUST|TERM TRUST"
    )
  )

  is_other_security <- stringr::str_detect(
    company_upper,
    "\\bBONDS?\\b|\\bDEBENTURES?\\b|CERTIFICATES?|STRUCTURED NOTES?"
  )

  dplyr::case_when(
    is_warrant_or_right ~ "Warrant or Right",
    is_adr ~ "ADR",
    is_preferred ~ "Preferred Stock",
    is_etf_or_etn ~ "ETF or ETN",
    is_unit_or_spac ~ "Unit or SPAC",
    is_fund ~ "Closed-End Fund or Other Fund",
    is_other_security ~ "Other or Unclassified",
    TRUE ~ "Common Equity"
  )
}

safe_median <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0L) NA_real_ else stats::median(x)
}

safe_quantile <- function(x, probability) {
  x <- x[is.finite(x)]
  if (length(x) == 0L) {
    return(NA_real_)
  }
  unname(stats::quantile(x, probs = probability, na.rm = TRUE, names = FALSE))
}

safe_rate <- function(numerator, denominator) {
  if (is.na(denominator) || denominator == 0) NA_real_ else numerator / denominator
}

format_count <- function(x) {
  dplyr::if_else(
    is.na(x),
    "NA",
    scales::number(x, accuracy = 1, big.mark = ",")
  )
}

format_dollar_short <- function(x) {
  dplyr::if_else(
    is.na(x),
    "NA",
    scales::dollar(x, accuracy = 0.1, scale_cut = scales::cut_short_scale())
  )
}

format_number_short <- function(x) {
  dplyr::if_else(
    is.na(x),
    "NA",
    scales::number(x, accuracy = 0.1, scale_cut = scales::cut_short_scale())
  )
}

format_percent_1 <- function(x) {
  dplyr::if_else(
    is.na(x),
    "NA",
    scales::percent(x, accuracy = 0.1)
  )
}

show_table <- function(data, caption = NULL, digits = 2) {
  if (!is.data.frame(data) || nrow(data) == 0L) {
    cat("*No observations are available for this table.*\n\n")
    return(invisible(NULL))
  }

  print(
    knitr::kable(
      data,
      format = "html",
      caption = caption,
      digits = digits,
      escape = TRUE,
      row.names = FALSE
    )
  )
  cat("\n\n")
  invisible(data)
}

empty_plot <- function(title, message, caption = NULL) {
  ggplot() +
    annotate("text", x = 0, y = 0, label = message, size = 4.2) +
    xlim(-1, 1) +
    ylim(-1, 1) +
    labs(title = title, caption = caption) +
    theme_void(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.caption = element_text(size = 8.5, color = "grey35", hjust = 0),
      plot.margin = margin(8, 12, 8, 8)
    )
}

split_vector <- function(x, size) {
  if (length(x) == 0L) {
    return(list())
  }
  size <- max(1L, as.integer(size))
  split(x, ceiling(seq_along(x) / size))
}

symbol_signature <- function(symbols) {
  symbols <- sort(unique(stats::na.omit(symbols)))
  temporary_file <- tempfile(fileext = ".txt")
  on.exit(unlink(temporary_file), add = TRUE)
  writeLines(symbols, temporary_file, useBytes = TRUE)
  substr(unname(tools::md5sum(temporary_file)), 1L, 12L)
}

# ---- Condition capture, retries, and caching -------------------------------
capture_call <- function(function_to_run) {
  captured_warnings <- character()

  result <- tryCatch(
    withCallingHandlers(
      function_to_run(),
      warning = function(warning_condition) {
        captured_warnings <<- c(
          captured_warnings,
          conditionMessage(warning_condition)
        )
        invokeRestart("muffleWarning")
      }
    ),
    error = function(error_condition) {
      structure(
        list(message = conditionMessage(error_condition)),
        class = "captured_error"
      )
    }
  )

  if (inherits(result, "captured_error")) {
    return(list(
      ok = FALSE,
      value = NULL,
      warnings = unique(captured_warnings),
      error = result$message
    ))
  }

  list(
    ok = TRUE,
    value = result,
    warnings = unique(captured_warnings),
    error = NA_character_
  )
}

retry_captured <- function(
  function_to_run,
  attempts = max_retry_attempts,
  base_delay_seconds = retry_base_delay_seconds
) {
  attempts <- max(1L, as.integer(attempts))
  all_warnings <- character()
  all_errors <- character()

  for (attempt_number in seq_len(attempts)) {
    attempt <- capture_call(function_to_run)
    all_warnings <- c(all_warnings, attempt$warnings)

    if (isTRUE(attempt$ok)) {
      return(list(
        ok = TRUE,
        value = attempt$value,
        warnings = unique(all_warnings),
        errors = unique(all_errors),
        attempts = attempt_number
      ))
    }

    all_errors <- c(all_errors, attempt$error)

    if (attempt_number < attempts) {
      Sys.sleep(base_delay_seconds * 2^(attempt_number - 1L))
    }
  }

  list(
    ok = FALSE,
    value = NULL,
    warnings = unique(all_warnings),
    errors = unique(all_errors),
    attempts = attempts
  )
}

read_cache_record <- function(path) {
  cached <- readRDS(path)

  if (is.list(cached) && all(c("value", "metadata") %in% names(cached))) {
    return(cached)
  }

  list(
    value = cached,
    metadata = list(
      cached_at = as.POSIXct(file.info(path)$mtime),
      analysis_date = NA_character_,
      cache_stem = tools::file_path_sans_ext(basename(path))
    )
  )
}

cache_get <- function(
  cache_stem,
  fetch_function,
  validator = function(x) !is.null(x),
  fallback_prefix = NULL,
  fetch_attempts = max_retry_attempts,
  return_invalid_fetch = FALSE
) {
  exact_path <- file.path(cache_directory, paste0(cache_stem, ".rds"))

  make_metadata_row <- function(
    source,
    path,
    cached_at,
    cache_analysis_date,
    attempts,
    warnings,
    errors
  ) {
    tibble::tibble(
      dataset = cache_stem,
      source = source,
      cache_file = ifelse(is.na(path), NA_character_, basename(path)),
      cache_analysis_date = as.character(cache_analysis_date),
      cached_at = as.character(cached_at),
      attempts = attempts,
      warning_count = length(unique(stats::na.omit(warnings))),
      error_count = length(unique(stats::na.omit(errors)))
    )
  }

  if (!refresh_data && file.exists(exact_path)) {
    cached_attempt <- capture_call(function() read_cache_record(exact_path))

    if (isTRUE(cached_attempt$ok) && validator(cached_attempt$value$value)) {
      record <- cached_attempt$value
      return(list(
        value = record$value,
        meta = make_metadata_row(
          source = "Exact cache",
          path = exact_path,
          cached_at = record$metadata$cached_at %||% file.info(exact_path)$mtime,
          cache_analysis_date = record$metadata$analysis_date %||% NA_character_,
          attempts = 0L,
          warnings = cached_attempt$warnings,
          errors = character()
        )
      ))
    }
  }

  fetched <- retry_captured(fetch_function, attempts = fetch_attempts)

  if (isTRUE(fetched$ok) && validator(fetched$value)) {
    record <- list(
      value = fetched$value,
      metadata = list(
        cached_at = Sys.time(),
        analysis_date = as.character(analysis_date),
        cache_stem = cache_stem
      )
    )

    save_attempt <- capture_call(function() saveRDS(record, exact_path))

    cache_write_succeeded <- isTRUE(save_attempt$ok)

    return(list(
      value = fetched$value,
      meta = make_metadata_row(
        source = if (cache_write_succeeded) {
          "Fresh download"
        } else {
          "Fresh download (cache write failed)"
        },
        path = if (cache_write_succeeded) exact_path else NA_character_,
        cached_at = record$metadata$cached_at,
        cache_analysis_date = record$metadata$analysis_date,
        attempts = fetched$attempts,
        warnings = c(fetched$warnings, save_attempt$warnings),
        errors = c(
          fetched$errors,
          if (cache_write_succeeded) character() else save_attempt$error
        )
      )
    ))
  }

  if (!is.null(fallback_prefix)) {
    fallback_files <- list.files(
      cache_directory,
      full.names = TRUE
    )
    fallback_basenames <- basename(fallback_files)
    fallback_files <- fallback_files[
      startsWith(fallback_basenames, fallback_prefix) &
        stringr::str_detect(fallback_basenames, "[.]rds$")
    ]
    fallback_files <- setdiff(fallback_files, exact_path)

    if (length(fallback_files) > 0L) {
      fallback_files <- fallback_files[
        order(file.info(fallback_files)$mtime, decreasing = TRUE)
      ]

      for (fallback_path in fallback_files) {
        fallback_attempt <- capture_call(function() read_cache_record(fallback_path))

        if (
          isTRUE(fallback_attempt$ok) &&
          validator(fallback_attempt$value$value)
        ) {
          record <- fallback_attempt$value
          return(list(
            value = record$value,
            meta = make_metadata_row(
              source = "Prior cache fallback",
              path = fallback_path,
              cached_at = record$metadata$cached_at %||% file.info(fallback_path)$mtime,
              cache_analysis_date = record$metadata$analysis_date %||% NA_character_,
              attempts = fetched$attempts,
              warnings = c(fetched$warnings, fallback_attempt$warnings),
              errors = fetched$errors
            )
          ))
        }
      }
    }
  }

  if (
    isTRUE(return_invalid_fetch) &&
    isTRUE(fetched$ok) &&
    !is.null(fetched$value)
  ) {
    return(list(
      value = fetched$value,
      meta = make_metadata_row(
        source = "Fresh download (not cached)",
        path = NA_character_,
        cached_at = Sys.time(),
        cache_analysis_date = as.character(analysis_date),
        attempts = fetched$attempts,
        warnings = fetched$warnings,
        errors = fetched$errors
      )
    ))
  }

  list(
    value = NULL,
    meta = make_metadata_row(
      source = "Unavailable",
      path = NA_character_,
      cached_at = NA_character_,
      cache_analysis_date = NA_character_,
      attempts = fetched$attempts,
      warnings = fetched$warnings,
      errors = fetched$errors
    )
  )
}

# ---- Reusable reporting helpers --------------------------------------------
metric_coverage_table <- function(data, metrics, labels = metrics) {
  if (!is.data.frame(data) || nrow(data) == 0L) {
    return(tibble::tibble())
  }

  purrr::map2_dfr(metrics, labels, function(metric, label) {
    values <- data[[metric]]
    valid <- is.finite(values)

    tibble::tibble(
      metric = label,
      initial_observations = nrow(data),
      valid_observations = sum(valid, na.rm = TRUE),
      missing_observations = sum(!valid, na.rm = TRUE),
      missingness_rate = mean(!valid),
      sp500_valid = sum(valid & data$is_sp500, na.rm = TRUE),
      broader_market_valid = sum(valid & !data$is_sp500, na.rm = TRUE)
    )
  })
}

make_density_plot <- function(
  data,
  variable,
  title,
  subtitle,
  x_label,
  caption,
  log10_axis = FALSE,
  lower_bound = NULL,
  upper_bound = NULL
) {
  if (!is.data.frame(data) || nrow(data) == 0L || !(variable %in% names(data))) {
    return(empty_plot(title, "No usable observations are available.", caption))
  }

  plot_data <- data |>
    dplyr::filter(is.finite(.data[[variable]]))

  if (!is.null(lower_bound)) {
    plot_data <- plot_data |>
      dplyr::filter(.data[[variable]] >= lower_bound)
  }

  if (!is.null(upper_bound)) {
    plot_data <- plot_data |>
      dplyr::filter(.data[[variable]] <= upper_bound)
  }

  if (isTRUE(log10_axis)) {
    plot_data <- plot_data |>
      dplyr::filter(.data[[variable]] > 0)
  }

  plot_data <- plot_data |>
    dplyr::group_by(benchmark_group) |>
    dplyr::filter(
      dplyr::n() >= 2L,
      dplyr::n_distinct(.data[[variable]]) >= 2L
    ) |>
    dplyr::ungroup()

  if (nrow(plot_data) < 2L || dplyr::n_distinct(plot_data[[variable]]) < 2L) {
    return(empty_plot(title, "Too few distinct observations are available for a density estimate.", caption))
  }

  plot <- ggplot(
    plot_data,
    aes(
      x = .data[[variable]],
      color = benchmark_group,
      fill = benchmark_group
    )
  ) +
    geom_density(alpha = 0.28, linewidth = 0.9, adjust = 1.05, na.rm = TRUE) +
    scale_color_manual(values = benchmark_colors, drop = FALSE) +
    scale_fill_manual(values = benchmark_colors, drop = FALSE) +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_label,
      y = "Density",
      caption = caption
    )

  if (isTRUE(log10_axis)) {
    plot <- plot +
      scale_x_log10(
        labels = scales::label_dollar(
          accuracy = 1,
          scale_cut = scales::cut_short_scale()
        )
      )
  }

  plot
}

package_versions <- tibble::tibble(
  package = required_packages,
  version = vapply(
    required_packages,
    function(package_name) as.character(utils::packageVersion(package_name)),
    character(1)
  )
)
```

```{r}
#| label: configuration-summary
#| results: asis

configuration_table <- tibble::tibble(
  setting = c(
    "Analysis date",
    "Historical lookback (calendar days)",
    "Liquidity window (trading days)",
    "Minimum valid trading days",
    "Quote batch size (configured maximum)",
    "Quote batch size (provider-effective maximum)",
    "Price batch size",
    "Maximum retry attempts",
    "Maximum recursive split depth",
    "Pause between batches (seconds)",
    "Refresh remote data",
    "Cache directory"
  ),
  value = c(
    as.character(analysis_date),
    as.character(lookback_calendar_days),
    as.character(liquidity_trading_days),
    as.character(minimum_valid_trading_days),
    as.character(quote_batch_size),
    as.character(min(quote_batch_size, provider_quote_limit)),
    as.character(price_batch_size),
    as.character(max_retry_attempts),
    as.character(max_recursive_split_depth),
    as.character(pause_between_batches_seconds),
    as.character(refresh_data),
    cache_directory
  )
)

show_table(configuration_table, "Analysis configuration")
show_table(package_versions, "Package versions used for this render")
```

## Data acquisition and construction of the master universe

### Exchange directories and S&P 500 membership

```{r}
#| label: acquire-directories-and-index

standardize_exchange_download <- function(data, exchange_name) {
  if (!is.data.frame(data) || nrow(data) == 0L) {
    stop(paste0("No records were returned for ", exchange_name, "."), call. = FALSE)
  }

  names(data) <- snake_names(names(data))
  data <- ensure_columns(
    data,
    c(
      "symbol",
      "company",
      "last_sale_price",
      "market_cap",
      "country",
      "ipo_year",
      "sector",
      "industry"
    )
  )

  data |>
    dplyr::transmute(
      symbol_original = clean_text_na(symbol),
      company = clean_text_na(company),
      exchange = stringr::str_to_upper(exchange_name),
      last_sale_price_exchange = positive_or_na(last_sale_price),
      market_cap_exchange = positive_or_na(market_cap),
      country = clean_text_na(country),
      ipo_year = suppressWarnings(as.integer(ipo_year)),
      sector_raw = clean_text_na(sector),
      industry_raw = clean_text_na(industry)
    )
}

fetch_exchange_one <- function(exchange_name) {
  downloaded <- tidyquant::tq_exchange(exchange_name)

  if (!is.data.frame(downloaded) || nrow(downloaded) == 0L) {
    stop(
      paste0("tidyquant::tq_exchange() returned no usable records for ", exchange_name, "."),
      call. = FALSE
    )
  }

  standardize_exchange_download(downloaded, exchange_name)
}

standardize_sp500_download <- function(data, retrieval_mode) {
  if (!is.data.frame(data) || nrow(data) == 0L) {
    stop("No S&P 500 constituent records were returned.", call. = FALSE)
  }

  names(data) <- snake_names(names(data))

  symbol_column <- first_existing_column(data, c("symbol", "ticker"))
  company_column <- first_existing_column(data, c("company", "name"), required = FALSE)
  weight_column <- first_existing_column(
    data,
    c("weight", "weight_percent", "portfolio_weight"),
    required = FALSE
  )
  sector_column <- first_existing_column(
    data,
    c("sector", "gics_sector"),
    required = FALSE
  )

  output <- tibble::tibble(
    sp500_symbol_original = clean_text_na(data[[symbol_column]]),
    sp500_company = if (is.null(company_column)) {
      rep(NA_character_, nrow(data))
    } else {
      clean_text_na(data[[company_column]])
    },
    sp500_weight_raw = if (is.null(weight_column)) {
      rep(NA_real_, nrow(data))
    } else {
      safe_numeric(data[[weight_column]])
    },
    sp500_sector_raw = if (is.null(sector_column)) {
      rep(NA_character_, nrow(data))
    } else {
      clean_text_na(data[[sector_column]])
    },
    sp500_retrieval_mode = retrieval_mode
  ) |>
    dplyr::mutate(
      symbol_yahoo = normalize_yahoo_symbol(sp500_symbol_original),
      sp500_sector_clean = standardize_sector(sp500_sector_raw)
    ) |>
    dplyr::filter(!is.na(symbol_yahoo)) |>
    dplyr::distinct(symbol_yahoo, .keep_all = TRUE)

  positive_weight_total <- sum(
    output$sp500_weight_raw[
      is.finite(output$sp500_weight_raw) & output$sp500_weight_raw > 0
    ],
    na.rm = TRUE
  )

  output |>
    dplyr::mutate(
      sp500_weight = dplyr::if_else(
        is.finite(sp500_weight_raw) &
          sp500_weight_raw > 0 &
          positive_weight_total > 0,
        sp500_weight_raw / positive_weight_total,
        NA_real_
      ),
      sp500_member = TRUE
    )
}

fetch_sp500 <- function() {
  tq_index_formals <- names(formals(tidyquant::tq_index))
  supports_explicit_fallback <- "use_fallback" %in% tq_index_formals

  live_attempt <- capture_call(function() {
    if (supports_explicit_fallback) {
      tidyquant::tq_index("SP500", use_fallback = FALSE)
    } else {
      tidyquant::tq_index("SP500")
    }
  })

  if (
    isTRUE(live_attempt$ok) &&
    is.data.frame(live_attempt$value) &&
    nrow(live_attempt$value) > 0L
  ) {
    retrieval_mode <- if (supports_explicit_fallback) {
      "Live SSGA holdings"
    } else {
      "tidyquant default retrieval"
    }

    return(standardize_sp500_download(live_attempt$value, retrieval_mode))
  }

  if (!supports_explicit_fallback) {
    stop(
      "The S&P 500 constituent table could not be retrieved by tidyquant::tq_index().",
      call. = FALSE
    )
  }

  packaged_fallback <- tidyquant::tq_index("SP500", use_fallback = TRUE)

  if (!is.data.frame(packaged_fallback) || nrow(packaged_fallback) == 0L) {
    stop(
      "Both the live S&P 500 retrieval and tidyquant's packaged fallback were unavailable.",
      call. = FALSE
    )
  }

  standardize_sp500_download(
    packaged_fallback,
    "tidyquant packaged fallback"
  )
}

exchange_names <- c("NASDAQ", "NYSE", "AMEX")

exchange_results <- purrr::set_names(exchange_names) |>
  purrr::map(function(exchange_name) {
    cache_get(
      cache_stem = paste0(
        "exchange_",
        stringr::str_to_lower(exchange_name),
        "_",
        analysis_date
      ),
      fetch_function = function() fetch_exchange_one(exchange_name),
      validator = function(x) {
        required_exchange_columns <- c(
          "symbol_original",
          "company",
          "exchange",
          "last_sale_price_exchange",
          "market_cap_exchange",
          "sector_raw",
          "industry_raw"
        )

        is.data.frame(x) &&
          nrow(x) > 0L &&
          all(required_exchange_columns %in% names(x))
      },
      fallback_prefix = paste0(
        "exchange_",
        stringr::str_to_lower(exchange_name),
        "_"
      )
    )
  })

available_exchange_data <- exchange_results |>
  purrr::map("value") |>
  purrr::keep(~ is.data.frame(.x) && nrow(.x) > 0L)

if (length(available_exchange_data) == 0L) {
  stop(
    "No exchange directory could be retrieved and no usable exchange cache was found.",
    call. = FALSE
  )
}

sp500_result <- cache_get(
  cache_stem = paste0("sp500_constituents_", analysis_date),
  fetch_function = fetch_sp500,
  validator = function(x) {
    required_sp500_columns <- c(
      "sp500_symbol_original",
      "symbol_yahoo",
      "sp500_company",
      "sp500_weight",
      "sp500_sector_clean",
      "sp500_retrieval_mode"
    )

    is.data.frame(x) &&
      nrow(x) > 0L &&
      all(required_sp500_columns %in% names(x))
  },
  fallback_prefix = "sp500_constituents_"
)

if (!is.data.frame(sp500_result$value) || nrow(sp500_result$value) == 0L) {
  stop(
    "The S&P 500 constituent table is unavailable and no usable cache was found.",
    call. = FALSE
  )
}

exchange_cache_manifest <- exchange_results |>
  purrr::map_dfr("meta")

cache_manifest <- dplyr::bind_rows(
  exchange_cache_manifest,
  sp500_result$meta
)

exchange_download_status <- tibble::tibble(
  exchange = exchange_names,
  records = purrr::map_int(
    exchange_results,
    function(result) {
      if (is.data.frame(result$value)) nrow(result$value) else 0L
    }
  ),
  source = purrr::map_chr(exchange_results, ~ .x$meta$source[[1]])
)

sp500_df <- sp500_result$value
raw_universe_df <- dplyr::bind_rows(available_exchange_data)
```

```{r}
#| label: acquisition-tables
#| results: asis

show_table(exchange_download_status, "Exchange-directory retrieval status")
show_table(
  cache_manifest,
  "Cache and retrieval manifest for essential source tables"
)

unavailable_exchanges <- exchange_download_status |>
  dplyr::filter(records == 0L) |>
  dplyr::pull(exchange)

if (length(unavailable_exchanges) > 0L) {
  cat(
    paste0(
      "::: {.callout-warning}
",
      "## Partial exchange universe

",
      "No live or cached directory was available for: ",
      paste(unavailable_exchanges, collapse = ", "),
      ". The report continues with the available exchanges, and all universe-level results should be interpreted as incomplete.
",
      ":::

"
    )
  )
}

sp500_retrieval_table <- sp500_df |>
  dplyr::count(sp500_retrieval_mode, name = "constituent_records")
show_table(sp500_retrieval_table, "S&P 500 source mode")
```

### Cleaning, symbol normalization, and membership matching

```{r}
#| label: clean-master-universe

raw_universe_count <- nrow(raw_universe_df)

invalid_entry_log <- raw_universe_df |>
  dplyr::mutate(
    invalid_symbol = is.na(symbol_original),
    invalid_company = is.na(company)
  ) |>
  dplyr::summarise(
    starting_records = dplyr::n(),
    missing_symbol = sum(invalid_symbol),
    missing_company = sum(invalid_company),
    removed_invalid_records = sum(invalid_symbol | invalid_company)
  )

universe_clean_stage_1 <- raw_universe_df |>
  dplyr::filter(!is.na(symbol_original), !is.na(company)) |>
  dplyr::mutate(
    symbol_original = stringr::str_to_upper(symbol_original),
    symbol_yahoo = normalize_yahoo_symbol(symbol_original),
    company = clean_text_na(company),
    sector_raw = clean_text_na(sector_raw),
    industry_raw = clean_text_na(industry_raw),
    country = clean_text_na(country),
    market_cap_exchange = positive_or_na(market_cap_exchange),
    last_sale_price_exchange = positive_or_na(last_sale_price_exchange)
  )

exact_duplicate_count <- nrow(universe_clean_stage_1) -
  nrow(dplyr::distinct(universe_clean_stage_1))

universe_clean_stage_2 <- universe_clean_stage_1 |>
  dplyr::distinct()

same_listing_duplicate_table <- universe_clean_stage_2 |>
  dplyr::count(exchange, symbol_original, name = "record_count") |>
  dplyr::filter(record_count > 1L) |>
  dplyr::arrange(dplyr::desc(record_count), exchange, symbol_original)

same_listing_duplicate_count <- sum(
  pmax(same_listing_duplicate_table$record_count - 1L, 0L)
)

universe_clean_stage_3 <- universe_clean_stage_2 |>
  dplyr::mutate(
    completeness_score =
      as.integer(!is.na(company)) +
      as.integer(!is.na(market_cap_exchange)) +
      as.integer(!is.na(last_sale_price_exchange)) +
      as.integer(!is.na(sector_raw)) +
      as.integer(!is.na(industry_raw)) +
      as.integer(!is.na(country))
  ) |>
  dplyr::group_by(exchange, symbol_original) |>
  dplyr::arrange(dplyr::desc(completeness_score), .by_group = TRUE) |>
  dplyr::slice_head(n = 1L) |>
  dplyr::ungroup() |>
  dplyr::select(-completeness_score)

universe_df <- universe_clean_stage_3 |>
  dplyr::mutate(
    security_type = classify_security_type(company, symbol_original),
    sector_exchange_clean = standardize_sector(sector_raw),
    industry_clean = clean_text_na(industry_raw)
  ) |>
  dplyr::left_join(
    sp500_df |>
      dplyr::select(
        symbol_yahoo,
        sp500_symbol_original,
        sp500_company,
        sp500_weight,
        sp500_sector_raw,
        sp500_sector_clean,
        sp500_member
      ),
    by = "symbol_yahoo"
  ) |>
  dplyr::mutate(
    is_sp500 = tidyr::replace_na(sp500_member, FALSE),
    benchmark_group = factor(
      dplyr::if_else(is_sp500, "S&P 500", "Broader Market"),
      levels = benchmark_levels
    ),
    sector_clean = dplyr::coalesce(
      sector_exchange_clean,
      sp500_sector_clean
    ),
    sector_display = dplyr::coalesce(sector_clean, "Unclassified"),
    industry_display = dplyr::coalesce(industry_clean, "Unclassified"),
    symbol_was_normalized =
      !is.na(symbol_yahoo) & symbol_original != symbol_yahoo
  ) |>
  dplyr::select(-sp500_member)

yahoo_symbol_duplicate_table <- universe_df |>
  dplyr::filter(!is.na(symbol_yahoo)) |>
  dplyr::count(symbol_yahoo, name = "listing_records") |>
  dplyr::filter(listing_records > 1L) |>
  dplyr::arrange(dplyr::desc(listing_records), symbol_yahoo)

sp500_universe_match <- sp500_df |>
  dplyr::anti_join(
    universe_df |>
      dplyr::distinct(symbol_yahoo),
    by = "symbol_yahoo"
  )

symbol_normalization_summary <- tibble::tibble(
  measure = c(
    "Raw exchange records",
    "Master-universe records after cleaning",
    "Original symbols",
    "Symbols changed for Yahoo compatibility",
    "Unique Yahoo-compatible symbols",
    "Yahoo-compatible symbols represented by multiple listing records",
    "S&P 500 constituent symbols",
    "S&P 500 constituent symbols matched to the exchange universe",
    "S&P 500 constituent symbols not matched to the exchange universe"
  ),
  value = c(
    raw_universe_count,
    nrow(universe_df),
    dplyr::n_distinct(universe_df$symbol_original, na.rm = TRUE),
    sum(universe_df$symbol_was_normalized, na.rm = TRUE),
    dplyr::n_distinct(universe_df$symbol_yahoo, na.rm = TRUE),
    nrow(yahoo_symbol_duplicate_table),
    dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE),
    dplyr::n_distinct(
      universe_df$symbol_yahoo[universe_df$is_sp500],
      na.rm = TRUE
    ),
    nrow(sp500_universe_match)
  )
)

master_data_quality <- tibble::tibble(
  metric = c(
    "Total ticker records",
    "Exact duplicate records removed",
    "Additional same-exchange/same-symbol records resolved",
    "Missing market capitalization rate",
    "Missing sector rate",
    "Missing industry rate",
    "Missing exchange last-price rate",
    "Other or unclassified security-type rate",
    "S&P 500 exchange-universe match rate"
  ),
  value = c(
    nrow(universe_df),
    exact_duplicate_count,
    same_listing_duplicate_count,
    mean(is.na(universe_df$market_cap_exchange)),
    mean(is.na(universe_df$sector_clean)),
    mean(is.na(universe_df$industry_clean)),
    mean(is.na(universe_df$last_sale_price_exchange)),
    mean(universe_df$security_type == "Other or Unclassified"),
    safe_rate(
      dplyr::n_distinct(
        universe_df$symbol_yahoo[universe_df$is_sp500],
        na.rm = TRUE
      ),
      dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE)
    )
  ),
  display_value = c(
    format_count(nrow(universe_df)),
    format_count(exact_duplicate_count),
    format_count(same_listing_duplicate_count),
    format_percent_1(mean(is.na(universe_df$market_cap_exchange))),
    format_percent_1(mean(is.na(universe_df$sector_clean))),
    format_percent_1(mean(is.na(universe_df$industry_clean))),
    format_percent_1(mean(is.na(universe_df$last_sale_price_exchange))),
    format_percent_1(mean(universe_df$security_type == "Other or Unclassified")),
    format_percent_1(
      safe_rate(
        dplyr::n_distinct(
          universe_df$symbol_yahoo[universe_df$is_sp500],
          na.rm = TRUE
        ),
        dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE)
      )
    )
  )
)
```

```{r}
#| label: master-quality-tables
#| results: asis

show_table(invalid_entry_log, "Invalid exchange-directory records")
show_table(symbol_normalization_summary, "Symbol normalization and S&P 500 matching")
show_table(
  master_data_quality |>
    dplyr::select(metric, display_value),
  "Master-universe data quality"
)

if (nrow(same_listing_duplicate_table) > 0L) {
  show_table(
    utils::head(same_listing_duplicate_table, 20L),
    "Same-exchange/same-symbol records resolved using metadata completeness (first 20)"
  )
}

if (nrow(sp500_universe_match) > 0L) {
  show_table(
    sp500_universe_match |>
      dplyr::select(
        symbol_yahoo,
        sp500_company,
        sp500_sector_clean,
        sp500_weight
      ),
    "S&P 500 constituents not matched to the exchange directory"
  )
}
```

The symbol-normalization rules preserve `symbol_original` and create a separate `symbol_yahoo`. Periods and slashes used for share classes are converted to dashes, and common preferred-share notations are translated to Yahoo-compatible forms. The master table retains exchange records; unmatched S&P 500 constituents are added only to the downstream download-candidate table so benchmark download coverage can be diagnosed without altering Module 1 listing counts.

## Operating-company candidates and market-data retrieval

### Candidate selection

```{r}
#| label: build-download-candidates

exchange_equity_candidates_all <- universe_df |>
  dplyr::filter(
    security_type %in% c("Common Equity", "ADR"),
    !is.na(symbol_yahoo)
  ) |>
  dplyr::mutate(
    source_record = "Exchange directory",
    exchange_priority = dplyr::case_when(
      exchange == "NYSE" ~ 1L,
      exchange == "NASDAQ" ~ 2L,
      exchange == "AMEX" ~ 3L,
      TRUE ~ 9L
    )
  )

candidate_yahoo_duplicate_table <- exchange_equity_candidates_all |>
  dplyr::count(symbol_yahoo, name = "candidate_records") |>
  dplyr::filter(candidate_records > 1L) |>
  dplyr::arrange(dplyr::desc(candidate_records), symbol_yahoo)

exchange_equity_candidates <- exchange_equity_candidates_all |>
  dplyr::arrange(
    symbol_yahoo,
    dplyr::desc(is_sp500),
    dplyr::desc(!is.na(market_cap_exchange)),
    dplyr::desc(!is.na(sector_clean)),
    dplyr::desc(!is.na(industry_clean)),
    exchange_priority
  ) |>
  dplyr::group_by(symbol_yahoo) |>
  dplyr::slice_head(n = 1L) |>
  dplyr::ungroup() |>
  dplyr::select(-exchange_priority)

sp500_candidate_supplement <- sp500_df |>
  dplyr::anti_join(
    exchange_equity_candidates |>
      dplyr::distinct(symbol_yahoo),
    by = "symbol_yahoo"
  ) |>
  dplyr::transmute(
    symbol_original = sp500_symbol_original,
    company = dplyr::coalesce(sp500_company, sp500_symbol_original),
    exchange = NA_character_,
    last_sale_price_exchange = NA_real_,
    market_cap_exchange = NA_real_,
    country = NA_character_,
    ipo_year = NA_integer_,
    sector_raw = sp500_sector_raw,
    industry_raw = NA_character_,
    symbol_yahoo = symbol_yahoo,
    security_type = "Common Equity",
    sector_exchange_clean = NA_character_,
    industry_clean = NA_character_,
    sp500_symbol_original = sp500_symbol_original,
    sp500_company = sp500_company,
    sp500_weight = sp500_weight,
    sp500_sector_raw = sp500_sector_raw,
    sp500_sector_clean = sp500_sector_clean,
    is_sp500 = TRUE,
    benchmark_group = factor("S&P 500", levels = benchmark_levels),
    sector_clean = sp500_sector_clean,
    sector_display = dplyr::coalesce(sp500_sector_clean, "Unclassified"),
    industry_display = "Unclassified",
    symbol_was_normalized =
      !is.na(symbol_yahoo) & sp500_symbol_original != symbol_yahoo,
    source_record = "S&P 500 supplement"
  )

equity_candidates_df <- dplyr::bind_rows(
  exchange_equity_candidates,
  sp500_candidate_supplement
) |>
  dplyr::arrange(symbol_yahoo, dplyr::desc(is_sp500)) |>
  dplyr::distinct(symbol_yahoo, .keep_all = TRUE) |>
  dplyr::mutate(
    benchmark_group = factor(benchmark_group, levels = benchmark_levels)
  )

download_symbols <- equity_candidates_df |>
  dplyr::filter(!is.na(symbol_yahoo)) |>
  dplyr::pull(symbol_yahoo) |>
  unique() |>
  sort()

download_symbol_hash <- symbol_signature(download_symbols)

candidate_summary <- tibble::tibble(
  measure = c(
    "Common-equity and ADR exchange records",
    "Duplicate Yahoo symbols resolved for download",
    "Unique exchange-derived download candidates",
    "Unmatched S&P 500 symbols supplemented for download diagnosis",
    "Total unique Yahoo symbols requested"
  ),
  value = c(
    nrow(exchange_equity_candidates_all),
    sum(pmax(candidate_yahoo_duplicate_table$candidate_records - 1L, 0L)),
    nrow(exchange_equity_candidates),
    nrow(sp500_candidate_supplement),
    length(download_symbols)
  )
)
```

```{r}
#| label: candidate-summary-table
#| results: asis

show_table(candidate_summary, "Operating-company download candidates")

if (nrow(candidate_yahoo_duplicate_table) > 0L) {
  show_table(
    utils::head(candidate_yahoo_duplicate_table, 20L),
    "Yahoo-compatible candidate symbols represented by multiple exchange records (first 20)"
  )
}
```

### Snapshot quote data

```{r}
#| label: quote-retrieval-functions

quote_fields <- c(
  "regularMarketPrice",
  "marketCap",
  "trailingPE",
  "priceToBook",
  "averageDailyVolume3Month",
  "averageDailyVolume10Day"
)

empty_condition_log <- function() {
  tibble::tibble(
    stage = character(),
    batch_id = character(),
    symbol_yahoo = character(),
    status = character(),
    message = character()
  )
}

empty_quote_rows <- function(symbols) {
  tibble::tibble(
    symbol_yahoo = symbols,
    quote_trade_time = NA_character_,
    quote_price = NA_real_,
    market_cap_quote = NA_real_,
    trailing_pe = NA_real_,
    price_to_book = NA_real_,
    average_daily_volume_3m_quote = NA_real_,
    average_daily_volume_10d_quote = NA_real_,
    quote_success = FALSE
  )
}

fetch_quote_batch_once <- function(symbols) {
  raw_quotes <- quantmod::getQuote(
    Symbols = symbols,
    src = "yahoo",
    what = quote_fields
  )

  if (!is.data.frame(raw_quotes) || nrow(raw_quotes) == 0L) {
    stop("Yahoo returned no quote rows for this batch.", call. = FALSE)
  }

  quote_data <- raw_quotes |>
    as.data.frame(stringsAsFactors = FALSE, check.names = FALSE) |>
    tibble::rownames_to_column("symbol_yahoo")

  names(quote_data) <- snake_names(names(quote_data))
  quote_data <- ensure_columns(
    quote_data,
    c(
      "symbol_yahoo",
      "regular_market_time",
      "regular_market_price",
      "market_cap",
      "trailing_pe",
      "price_to_book",
      "average_daily_volume3_month",
      "average_daily_volume10_day"
    )
  )

  quote_data <- quote_data |>
    dplyr::transmute(
      symbol_yahoo = normalize_yahoo_symbol(symbol_yahoo),
      quote_trade_time = as.character(regular_market_time),
      quote_price = positive_or_na(regular_market_price),
      market_cap_quote = positive_or_na(market_cap),
      trailing_pe = safe_numeric(trailing_pe),
      price_to_book = safe_numeric(price_to_book),
      average_daily_volume_3m_quote = positive_or_na(average_daily_volume3_month),
      average_daily_volume_10d_quote = positive_or_na(average_daily_volume10_day)
    ) |>
    dplyr::distinct(symbol_yahoo, .keep_all = TRUE) |>
    dplyr::mutate(
      quote_success = dplyr::if_any(
        c(
          quote_price,
          market_cap_quote,
          trailing_pe,
          price_to_book,
          average_daily_volume_3m_quote,
          average_daily_volume_10d_quote
        ),
        ~ is.finite(.x)
      )
    )

  tibble::tibble(symbol_yahoo = symbols) |>
    dplyr::left_join(quote_data, by = "symbol_yahoo") |>
    dplyr::mutate(quote_success = tidyr::replace_na(quote_success, FALSE))
}

condition_log_rows <- function(attempt, stage, batch_id, symbols) {
  warning_rows <- if (length(attempt$warnings) > 0L) {
    tibble::tibble(
      stage = stage,
      batch_id = batch_id,
      symbol_yahoo = NA_character_,
      status = "warning",
      message = attempt$warnings
    )
  } else {
    tibble::tibble()
  }

  error_rows <- if (length(attempt$errors) > 0L) {
    tibble::tibble(
      stage = stage,
      batch_id = batch_id,
      symbol_yahoo = if (length(symbols) == 1L) symbols else NA_character_,
      status = "error",
      message = attempt$errors
    )
  } else {
    tibble::tibble()
  }

  dplyr::bind_rows(warning_rows, error_rows)
}

is_global_download_error <- function(messages) {
  messages <- unique(stats::na.omit(messages))
  if (length(messages) == 0L) {
    return(FALSE)
  }

  message_text <- stringr::str_to_lower(paste(messages, collapse = " | "))

  stringr::str_detect(
    message_text,
    paste0(
      "crumb|consent|could not resolve|failed to connect|connection refused|",
      "cannot open.*url|cannot open the connection|download failed|",
      "timed out|timeout|ssl|rate limit|too many requests|",
      "http[^0-9]*(401|403|429|500|502|503|504)|service unavailable"
    )
  )
}

fetch_quote_resilient <- function(symbols, batch_id, depth = 0L) {
  attempt <- retry_captured(
    function() fetch_quote_batch_once(symbols),
    attempts = max_retry_attempts
  )

  attempt_log <- condition_log_rows(
    attempt,
    stage = "snapshot_quote",
    batch_id = batch_id,
    symbols = symbols
  )

  if (isTRUE(attempt$ok)) {
    quote_data <- attempt$value
    global_failure <-
      !any(quote_data$quote_success, na.rm = TRUE) &&
      is_global_download_error(attempt$warnings)

    no_data_log <- quote_data |>
      dplyr::filter(!quote_success) |>
      dplyr::transmute(
        stage = "snapshot_quote",
        batch_id = batch_id,
        symbol_yahoo = symbol_yahoo,
        status = dplyr::if_else(
          global_failure,
          "global_failure",
          "no_data"
        ),
        message = dplyr::if_else(
          global_failure,
          "The provider returned no usable quote data and emitted a global connection or authorization warning.",
          "No requested quote field was returned."
        )
      )

    return(list(
      data = quote_data,
      log = dplyr::bind_rows(attempt_log, no_data_log),
      global_failure = global_failure
    ))
  }

  global_failure <- is_global_download_error(c(attempt$errors, attempt$warnings))
  may_split <-
    !global_failure &&
    length(symbols) > 1L &&
    depth < max_recursive_split_depth

  if (may_split) {
    split_point <- ceiling(length(symbols) / 2L)
    left_symbols <- symbols[seq_len(split_point)]
    right_symbols <- symbols[-seq_len(split_point)]

    left_result <- fetch_quote_resilient(
      left_symbols,
      paste0(batch_id, "L"),
      depth = depth + 1L
    )
    right_result <- fetch_quote_resilient(
      right_symbols,
      paste0(batch_id, "R"),
      depth = depth + 1L
    )

    return(list(
      data = dplyr::bind_rows(left_result$data, right_result$data),
      log = dplyr::bind_rows(
        attempt_log,
        left_result$log,
        right_result$log
      ),
      global_failure = isTRUE(left_result$global_failure) &&
        isTRUE(right_result$global_failure)
    ))
  }

  failure_message <- paste(
    unique(c(attempt$errors, attempt$warnings)),
    collapse = " | "
  )
  if (identical(failure_message, "")) {
    failure_message <- "The quote batch failed without an error message."
  }

  list(
    data = empty_quote_rows(symbols),
    log = dplyr::bind_rows(
      attempt_log,
      tibble::tibble(
        stage = "snapshot_quote",
        batch_id = batch_id,
        symbol_yahoo = symbols,
        status = if (global_failure) "global_failure" else "failed_batch",
        message = failure_message
      )
    ),
    global_failure = global_failure
  )
}

fetch_all_quotes <- function(symbols) {
  effective_batch_size <- min(
    as.integer(quote_batch_size),
    as.integer(provider_quote_limit)
  )
  batches <- split_vector(symbols, effective_batch_size)
  results <- vector("list", length(batches))

  if (length(batches) == 0L) {
    return(list(
      data = empty_quote_rows(character()),
      log = empty_condition_log(),
      requested_symbols = symbols,
      effective_batch_size = effective_batch_size,
      fetched_at = Sys.time()
    ))
  }

  for (batch_number in seq_along(batches)) {
    message(
      "Retrieving quote batch ",
      batch_number,
      " of ",
      length(batches),
      "."
    )

    results[[batch_number]] <- fetch_quote_resilient(
      batches[[batch_number]],
      sprintf("Q%04d", batch_number)
    )

    if (isTRUE(results[[batch_number]]$global_failure)) {
      remaining_batches <- if (batch_number < length(batches)) {
        seq.int(batch_number + 1L, length(batches))
      } else {
        integer()
      }

      for (remaining_number in remaining_batches) {
        skipped_symbols <- batches[[remaining_number]]
        results[[remaining_number]] <- list(
          data = empty_quote_rows(skipped_symbols),
          log = tibble::tibble(
            stage = "snapshot_quote",
            batch_id = sprintf("Q%04d", remaining_number),
            symbol_yahoo = skipped_symbols,
            status = "skipped_after_global_failure",
            message = "Skipped after a provider-wide quote failure was detected."
          ),
          global_failure = TRUE
        )
      }
      break
    }

    if (batch_number < length(batches)) {
      Sys.sleep(pause_between_batches_seconds)
    }
  }

  quote_data <- purrr::map_dfr(results, "data") |>
    dplyr::arrange(symbol_yahoo) |>
    dplyr::distinct(symbol_yahoo, .keep_all = TRUE)

  quote_data <- tibble::tibble(symbol_yahoo = symbols) |>
    dplyr::left_join(quote_data, by = "symbol_yahoo") |>
    dplyr::mutate(quote_success = tidyr::replace_na(quote_success, FALSE))

  list(
    data = quote_data,
    log = purrr::map_dfr(results, "log"),
    requested_symbols = symbols,
    effective_batch_size = effective_batch_size,
    fetched_at = Sys.time()
  )
}

quote_cache <- cache_get(
  cache_stem = paste0(
    "quote_snapshot_",
    analysis_date,
    "_",
    download_symbol_hash
  ),
  fetch_function = function() fetch_all_quotes(download_symbols),
  validator = function(x) {
    required_quote_columns <- c(
      "symbol_yahoo",
      "quote_trade_time",
      "quote_price",
      "market_cap_quote",
      "trailing_pe",
      "price_to_book",
      "average_daily_volume_3m_quote",
      "average_daily_volume_10d_quote",
      "quote_success"
    )

    is.list(x) &&
      all(
        c("data", "log", "requested_symbols", "effective_batch_size") %in%
          names(x)
      ) &&
      is.data.frame(x$data) &&
      is.data.frame(x$log) &&
      all(required_quote_columns %in% names(x$data)) &&
      (
        length(x$requested_symbols) == 0L ||
          any(x$data$quote_success, na.rm = TRUE)
      )
  },
  fallback_prefix = NULL,
  fetch_attempts = 1L,
  return_invalid_fetch = TRUE
)

if (is.null(quote_cache$value)) {
  quote_fetch_result <- list(
    data = empty_quote_rows(download_symbols),
    log = tibble::tibble(
      stage = "snapshot_quote",
      batch_id = NA_character_,
      symbol_yahoo = download_symbols,
      status = "failed_symbol",
      message = "The quote retrieval workflow returned no result."
    ),
    requested_symbols = download_symbols,
    effective_batch_size = min(quote_batch_size, provider_quote_limit),
    fetched_at = Sys.time()
  )
} else {
  quote_fetch_result <- quote_cache$value
}

quote_df <- quote_fetch_result$data
quote_log <- dplyr::bind_rows(empty_condition_log(), quote_fetch_result$log)

quote_retrieval_summary <- tibble::tibble(
  measure = c(
    "Requested symbols",
    "Symbols with at least one requested quote field",
    "Symbols with no requested quote field",
    "Quote success rate",
    "Quote log rows",
    "Effective quote batch size"
  ),
  value = c(
    length(download_symbols),
    sum(quote_df$quote_success, na.rm = TRUE),
    sum(!quote_df$quote_success, na.rm = TRUE),
    safe_rate(sum(quote_df$quote_success, na.rm = TRUE), length(download_symbols)),
    nrow(quote_log),
    quote_fetch_result$effective_batch_size
  ),
  display_value = c(
    format_count(length(download_symbols)),
    format_count(sum(quote_df$quote_success, na.rm = TRUE)),
    format_count(sum(!quote_df$quote_success, na.rm = TRUE)),
    format_percent_1(
      safe_rate(sum(quote_df$quote_success, na.rm = TRUE), length(download_symbols))
    ),
    format_count(nrow(quote_log)),
    format_count(quote_fetch_result$effective_batch_size)
  )
)
```

### Historical prices and 30-day liquidity

```{r}
#| label: price-retrieval-functions

empty_price_data <- function() {
  tibble::tibble(
    symbol_yahoo = character(),
    date = as.Date(character()),
    open = numeric(),
    high = numeric(),
    low = numeric(),
    close = numeric(),
    volume = numeric(),
    adjusted = numeric()
  )
}

fetch_price_batch_once <- function(symbols) {
  raw_prices <- tidyquant::tq_get(
    x = symbols,
    get = "stock.prices",
    from = price_from_date,
    to = price_to_date,
    complete_cases = FALSE
  )

  if (!is.data.frame(raw_prices)) {
    stop("Yahoo returned an invalid historical-price object.", call. = FALSE)
  }

  names(raw_prices) <- snake_names(names(raw_prices))

  if (!("symbol" %in% names(raw_prices))) {
    if (length(symbols) == 1L && nrow(raw_prices) > 0L) {
      raw_prices$symbol <- symbols[[1]]
    } else if (nrow(raw_prices) > 0L) {
      stop(
        "Historical prices for a multi-symbol batch did not include a symbol column.",
        call. = FALSE
      )
    }
  }

  raw_prices <- ensure_columns(
    raw_prices,
    c("symbol", "date", "open", "high", "low", "close", "volume", "adjusted")
  )

  raw_prices |>
    dplyr::transmute(
      symbol_yahoo = normalize_yahoo_symbol(symbol),
      date = as.Date(date),
      open = safe_numeric(open),
      high = safe_numeric(high),
      low = safe_numeric(low),
      close = safe_numeric(close),
      volume = safe_numeric(volume),
      adjusted = safe_numeric(adjusted)
    ) |>
    dplyr::filter(!is.na(symbol_yahoo), !is.na(date)) |>
    dplyr::distinct(symbol_yahoo, date, .keep_all = TRUE)
}

fetch_price_resilient <- function(symbols, batch_id, depth = 0L) {
  attempt <- retry_captured(
    function() fetch_price_batch_once(symbols),
    attempts = max_retry_attempts
  )

  attempt_log <- condition_log_rows(
    attempt,
    stage = "historical_price",
    batch_id = batch_id,
    symbols = symbols
  )

  if (isTRUE(attempt$ok)) {
    price_data <- attempt$value
    successful_symbols <- price_data |>
      dplyr::filter(
        !is.na(date),
        is.finite(close),
        close > 0,
        is.finite(volume),
        volume >= 0
      ) |>
      dplyr::distinct(symbol_yahoo) |>
      dplyr::pull(symbol_yahoo)

    missing_symbols <- setdiff(symbols, successful_symbols)
    global_failure <-
      length(successful_symbols) == 0L &&
      is_global_download_error(attempt$warnings)

    no_data_log <- if (length(missing_symbols) > 0L) {
      tibble::tibble(
        stage = "historical_price",
        batch_id = batch_id,
        symbol_yahoo = missing_symbols,
        status = if (global_failure) "global_failure" else "no_data",
        message = if (global_failure) {
          "The provider returned no usable price data and emitted a global connection or authorization warning."
        } else {
          "No valid close-and-volume observation was returned."
        }
      )
    } else {
      tibble::tibble()
    }

    return(list(
      data = price_data,
      log = dplyr::bind_rows(attempt_log, no_data_log),
      global_failure = global_failure
    ))
  }

  global_failure <- is_global_download_error(c(attempt$errors, attempt$warnings))
  may_split <-
    !global_failure &&
    length(symbols) > 1L &&
    depth < max_recursive_split_depth

  if (may_split) {
    split_point <- ceiling(length(symbols) / 2L)
    left_symbols <- symbols[seq_len(split_point)]
    right_symbols <- symbols[-seq_len(split_point)]

    left_result <- fetch_price_resilient(
      left_symbols,
      paste0(batch_id, "L"),
      depth = depth + 1L
    )
    right_result <- fetch_price_resilient(
      right_symbols,
      paste0(batch_id, "R"),
      depth = depth + 1L
    )

    return(list(
      data = dplyr::bind_rows(left_result$data, right_result$data),
      log = dplyr::bind_rows(
        attempt_log,
        left_result$log,
        right_result$log
      ),
      global_failure = isTRUE(left_result$global_failure) &&
        isTRUE(right_result$global_failure)
    ))
  }

  failure_message <- paste(
    unique(c(attempt$errors, attempt$warnings)),
    collapse = " | "
  )
  if (identical(failure_message, "")) {
    failure_message <- "The historical-price batch failed without an error message."
  }

  list(
    data = empty_price_data(),
    log = dplyr::bind_rows(
      attempt_log,
      tibble::tibble(
        stage = "historical_price",
        batch_id = batch_id,
        symbol_yahoo = symbols,
        status = if (global_failure) "global_failure" else "failed_batch",
        message = failure_message
      )
    ),
    global_failure = global_failure
  )
}

fetch_all_prices <- function(symbols) {
  batches <- split_vector(symbols, price_batch_size)
  results <- vector("list", length(batches))

  if (length(batches) == 0L) {
    return(list(
      data = empty_price_data(),
      log = empty_condition_log(),
      requested_symbols = symbols,
      effective_batch_size = price_batch_size,
      from = price_from_date,
      to = price_to_date,
      fetched_at = Sys.time()
    ))
  }

  for (batch_number in seq_along(batches)) {
    message(
      "Retrieving historical-price batch ",
      batch_number,
      " of ",
      length(batches),
      "."
    )

    results[[batch_number]] <- fetch_price_resilient(
      batches[[batch_number]],
      sprintf("P%04d", batch_number)
    )

    if (isTRUE(results[[batch_number]]$global_failure)) {
      remaining_batches <- if (batch_number < length(batches)) {
        seq.int(batch_number + 1L, length(batches))
      } else {
        integer()
      }

      for (remaining_number in remaining_batches) {
        skipped_symbols <- batches[[remaining_number]]
        results[[remaining_number]] <- list(
          data = empty_price_data(),
          log = tibble::tibble(
            stage = "historical_price",
            batch_id = sprintf("P%04d", remaining_number),
            symbol_yahoo = skipped_symbols,
            status = "skipped_after_global_failure",
            message = "Skipped after a provider-wide historical-price failure was detected."
          ),
          global_failure = TRUE
        )
      }
      break
    }

    if (batch_number < length(batches)) {
      Sys.sleep(pause_between_batches_seconds)
    }
  }

  price_data <- purrr::map_dfr(results, "data") |>
    dplyr::distinct(symbol_yahoo, date, .keep_all = TRUE) |>
    dplyr::arrange(symbol_yahoo, date)

  list(
    data = price_data,
    log = purrr::map_dfr(results, "log"),
    requested_symbols = symbols,
    effective_batch_size = price_batch_size,
    from = price_from_date,
    to = price_to_date,
    fetched_at = Sys.time()
  )
}

price_cache <- cache_get(
  cache_stem = paste0(
    "price_history_",
    price_from_date,
    "_",
    analysis_date,
    "_",
    download_symbol_hash
  ),
  fetch_function = function() fetch_all_prices(download_symbols),
  validator = function(x) {
    required_price_columns <- c(
      "symbol_yahoo",
      "date",
      "open",
      "high",
      "low",
      "close",
      "volume",
      "adjusted"
    )

    is.list(x) &&
      all(c("data", "log", "requested_symbols") %in% names(x)) &&
      is.data.frame(x$data) &&
      is.data.frame(x$log) &&
      all(required_price_columns %in% names(x$data)) &&
      (
        length(x$requested_symbols) == 0L ||
          any(
            is.finite(x$data$close) &
              x$data$close > 0 &
              is.finite(x$data$volume) &
              x$data$volume >= 0,
            na.rm = TRUE
          )
      )
  },
  fallback_prefix = NULL,
  fetch_attempts = 1L,
  return_invalid_fetch = TRUE
)

if (is.null(price_cache$value)) {
  price_fetch_result <- list(
    data = empty_price_data(),
    log = tibble::tibble(
      stage = "historical_price",
      batch_id = NA_character_,
      symbol_yahoo = download_symbols,
      status = "failed_symbol",
      message = "The historical-price retrieval workflow returned no result."
    ),
    requested_symbols = download_symbols,
    effective_batch_size = price_batch_size,
    from = price_from_date,
    to = price_to_date,
    fetched_at = Sys.time()
  )
} else {
  price_fetch_result <- price_cache$value
}

price_history_df <- price_fetch_result$data
price_log <- dplyr::bind_rows(empty_condition_log(), price_fetch_result$log)

valid_price_history_df <- price_history_df |>
  dplyr::filter(
    date <= analysis_date,
    is.finite(close),
    close > 0,
    is.finite(volume),
    volume >= 0
  ) |>
  dplyr::distinct(symbol_yahoo, date, .keep_all = TRUE)

liquidity_df <- valid_price_history_df |>
  dplyr::group_by(symbol_yahoo) |>
  dplyr::arrange(dplyr::desc(date), .by_group = TRUE) |>
  dplyr::slice_head(n = liquidity_trading_days) |>
  dplyr::summarise(
    valid_trading_days = dplyr::n(),
    latest_history_date = dplyr::first(date),
    earliest_history_date_used = dplyr::last(date),
    latest_close = dplyr::first(close),
    avg_daily_share_volume_raw = mean(volume, na.rm = TRUE),
    avg_daily_dollar_volume_raw = mean(close * volume, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    historical_data_available = valid_trading_days > 0L,
    liquidity_history_adequate =
      valid_trading_days >= minimum_valid_trading_days,
    avg_daily_share_volume_30d = dplyr::if_else(
      liquidity_history_adequate,
      avg_daily_share_volume_raw,
      NA_real_
    ),
    avg_daily_dollar_volume_30d = dplyr::if_else(
      liquidity_history_adequate,
      avg_daily_dollar_volume_raw,
      NA_real_
    )
  )

price_symbol_coverage <- tibble::tibble(symbol_yahoo = download_symbols) |>
  dplyr::left_join(
    liquidity_df |>
      dplyr::select(
        symbol_yahoo,
        valid_trading_days,
        historical_data_available,
        liquidity_history_adequate
      ),
    by = "symbol_yahoo"
  ) |>
  dplyr::mutate(
    valid_trading_days = tidyr::replace_na(valid_trading_days, 0L),
    historical_data_available =
      tidyr::replace_na(historical_data_available, FALSE),
    liquidity_history_adequate =
      tidyr::replace_na(liquidity_history_adequate, FALSE)
  )

price_retrieval_summary <- tibble::tibble(
  measure = c(
    "Requested symbols",
    "Symbols with at least one valid close-and-volume observation",
    "Symbols with adequate liquidity history",
    "Symbols with insufficient or unavailable liquidity history",
    "Historical-price success rate",
    "Median valid trading observations among retrieved symbols",
    "Historical-price log rows"
  ),
  value = c(
    length(download_symbols),
    sum(price_symbol_coverage$historical_data_available),
    sum(price_symbol_coverage$liquidity_history_adequate),
    sum(!price_symbol_coverage$liquidity_history_adequate),
    safe_rate(
      sum(price_symbol_coverage$historical_data_available),
      length(download_symbols)
    ),
    safe_median(
      price_symbol_coverage$valid_trading_days[
        price_symbol_coverage$historical_data_available
      ]
    ),
    nrow(price_log)
  ),
  display_value = c(
    format_count(length(download_symbols)),
    format_count(sum(price_symbol_coverage$historical_data_available)),
    format_count(sum(price_symbol_coverage$liquidity_history_adequate)),
    format_count(sum(!price_symbol_coverage$liquidity_history_adequate)),
    format_percent_1(
      safe_rate(
        sum(price_symbol_coverage$historical_data_available),
        length(download_symbols)
      )
    ),
    format_count(
      safe_median(
        price_symbol_coverage$valid_trading_days[
          price_symbol_coverage$historical_data_available
        ]
      )
    ),
    format_count(nrow(price_log))
  )
)
```

```{r}
#| label: retrieval-summary-tables
#| results: asis

market_data_cache_manifest <- dplyr::bind_rows(
  quote_cache$meta,
  price_cache$meta
)

show_table(
  market_data_cache_manifest,
  "Cache and retrieval manifest for quote and historical-price data"
)
show_table(
  quote_retrieval_summary |>
    dplyr::select(measure, display_value),
  "Snapshot-quote retrieval summary"
)
show_table(
  price_retrieval_summary |>
    dplyr::select(measure, display_value),
  "Historical-price and liquidity retrieval summary"
)

quote_log_summary <- quote_log |>
  dplyr::count(status, name = "log_rows") |>
  dplyr::arrange(dplyr::desc(log_rows))
price_log_summary <- price_log |>
  dplyr::count(status, name = "log_rows") |>
  dplyr::arrange(dplyr::desc(log_rows))

show_table(quote_log_summary, "Snapshot-quote log by status")
show_table(price_log_summary, "Historical-price log by status")
```

### Final operating-company equity universe

```{r}
#| label: build-final-equity-universe

analysis_enriched_df <- equity_candidates_df |>
  dplyr::left_join(quote_df, by = "symbol_yahoo") |>
  dplyr::left_join(liquidity_df, by = "symbol_yahoo") |>
  dplyr::mutate(
    quote_success = tidyr::replace_na(quote_success, FALSE),
    historical_data_available =
      tidyr::replace_na(historical_data_available, FALSE),
    liquidity_history_adequate =
      tidyr::replace_na(liquidity_history_adequate, FALSE),
    valid_yahoo_data = quote_success | historical_data_available,
    market_cap = dplyr::case_when(
      is.finite(market_cap_exchange) & market_cap_exchange > 0 ~
        market_cap_exchange,
      is.finite(market_cap_quote) & market_cap_quote > 0 ~
        market_cap_quote,
      TRUE ~ NA_real_
    ),
    market_cap_source = dplyr::case_when(
      is.finite(market_cap_exchange) & market_cap_exchange > 0 ~
        "Exchange directory",
      is.finite(market_cap_quote) & market_cap_quote > 0 ~
        "Yahoo quote fallback",
      TRUE ~ "Unavailable"
    ),
    reference_price = dplyr::case_when(
      is.finite(quote_price) & quote_price > 0 ~ quote_price,
      is.finite(latest_close) & latest_close > 0 ~ latest_close,
      is.finite(last_sale_price_exchange) & last_sale_price_exchange > 0 ~
        last_sale_price_exchange,
      TRUE ~ NA_real_
    ),
    price_source = dplyr::case_when(
      is.finite(quote_price) & quote_price > 0 ~ "Yahoo snapshot quote",
      is.finite(latest_close) & latest_close > 0 ~ "Yahoo historical close",
      is.finite(last_sale_price_exchange) & last_sale_price_exchange > 0 ~
        "Exchange directory",
      TRUE ~ "Unavailable"
    ),
    trailing_pe = dplyr::if_else(
      is.finite(trailing_pe),
      trailing_pe,
      NA_real_
    ),
    price_to_book = dplyr::if_else(
      is.finite(price_to_book),
      price_to_book,
      NA_real_
    ),
    operating_margin = NA_real_,
    benchmark_group = factor(
      dplyr::if_else(is_sp500, "S&P 500", "Broader Market"),
      levels = benchmark_levels
    )
  )

apply_equity_filters <- function(data) {
  current <- data
  log_rows <- list(
    tibble::tibble(
      step = "Unique Common Equity/ADR candidate symbols",
      starting_observations = nrow(current),
      removed_at_step = 0L,
      remaining_observations = nrow(current),
      removal_rate_from_previous = 0
    )
  )

  filter_steps <- list(
    list(
      label = "Require valid Yahoo quote or historical-price data",
      keep = function(x) x$valid_yahoo_data
    ),
    list(
      label = "Require positive market capitalization",
      keep = function(x) is.finite(x$market_cap) & x$market_cap > 0
    ),
    list(
      label = "Require positive reference price",
      keep = function(x) is.finite(x$reference_price) & x$reference_price > 0
    ),
    list(
      label = "Require populated standardized sector",
      keep = function(x) !is.na(x$sector_clean)
    ),
    list(
      label = "Require populated industry classification",
      keep = function(x) !is.na(x$industry_clean)
    )
  )

  for (step_index in seq_along(filter_steps)) {
    step <- filter_steps[[step_index]]
    starting_count <- nrow(current)
    keep_vector <- step$keep(current)
    keep_vector[is.na(keep_vector)] <- FALSE
    next_data <- current[keep_vector, , drop = FALSE]
    remaining_count <- nrow(next_data)
    removed_count <- starting_count - remaining_count

    log_rows[[length(log_rows) + 1L]] <- tibble::tibble(
      step = step$label,
      starting_observations = starting_count,
      removed_at_step = removed_count,
      remaining_observations = remaining_count,
      removal_rate_from_previous = safe_rate(
        removed_count,
        starting_count
      )
    )

    current <- next_data
  }

  list(
    data = current,
    log = dplyr::bind_rows(log_rows)
  )
}

post_retrieval_data_quality <- tibble::tibble(
  metric = c(
    "Unique download-candidate symbols",
    "Duplicate Yahoo-compatible symbol rate",
    "Valid Yahoo-data rate",
    "Missing positive market-capitalization rate",
    "Missing positive reference-price rate",
    "Missing standardized-sector rate",
    "Missing industry-classification rate",
    "S&P 500 candidate valid Yahoo-data rate"
  ),
  value = c(
    nrow(analysis_enriched_df),
    safe_rate(
      nrow(analysis_enriched_df) -
        dplyr::n_distinct(analysis_enriched_df$symbol_yahoo, na.rm = TRUE),
      nrow(analysis_enriched_df)
    ),
    safe_rate(
      sum(analysis_enriched_df$valid_yahoo_data, na.rm = TRUE),
      nrow(analysis_enriched_df)
    ),
    mean(
      !(is.finite(analysis_enriched_df$market_cap) &
          analysis_enriched_df$market_cap > 0)
    ),
    mean(
      !(is.finite(analysis_enriched_df$reference_price) &
          analysis_enriched_df$reference_price > 0)
    ),
    mean(is.na(analysis_enriched_df$sector_clean)),
    mean(is.na(analysis_enriched_df$industry_clean)),
    safe_rate(
      sum(
        analysis_enriched_df$is_sp500 &
          analysis_enriched_df$valid_yahoo_data,
        na.rm = TRUE
      ),
      sum(analysis_enriched_df$is_sp500, na.rm = TRUE)
    )
  ),
  display_value = c(
    format_count(nrow(analysis_enriched_df)),
    format_percent_1(
      safe_rate(
        nrow(analysis_enriched_df) -
          dplyr::n_distinct(
            analysis_enriched_df$symbol_yahoo,
            na.rm = TRUE
          ),
        nrow(analysis_enriched_df)
      )
    ),
    format_percent_1(
      safe_rate(
        sum(analysis_enriched_df$valid_yahoo_data, na.rm = TRUE),
        nrow(analysis_enriched_df)
      )
    ),
    format_percent_1(
      mean(
        !(is.finite(analysis_enriched_df$market_cap) &
            analysis_enriched_df$market_cap > 0)
      )
    ),
    format_percent_1(
      mean(
        !(is.finite(analysis_enriched_df$reference_price) &
            analysis_enriched_df$reference_price > 0)
      )
    ),
    format_percent_1(mean(is.na(analysis_enriched_df$sector_clean))),
    format_percent_1(mean(is.na(analysis_enriched_df$industry_clean))),
    format_percent_1(
      safe_rate(
        sum(
          analysis_enriched_df$is_sp500 &
            analysis_enriched_df$valid_yahoo_data,
          na.rm = TRUE
        ),
        sum(analysis_enriched_df$is_sp500, na.rm = TRUE)
      )
    )
  )
)

equity_filter_result <- apply_equity_filters(analysis_enriched_df)

equity_universe_df <- equity_filter_result$data |>
  dplyr::arrange(symbol_yahoo) |>
  dplyr::distinct(symbol_yahoo, .keep_all = TRUE) |>
  dplyr::mutate(
    benchmark_group = factor(benchmark_group, levels = benchmark_levels)
  )

equity_attrition_table <- equity_filter_result$log |>
  dplyr::mutate(
    removal_rate_display = format_percent_1(removal_rate_from_previous)
  )

final_universe_group_counts <- equity_universe_df |>
  dplyr::count(benchmark_group, name = "ticker_count", .drop = FALSE)

quote_log_file <- file.path(
  cache_directory,
  paste0("quote_log_", analysis_date, "_", download_symbol_hash, ".csv")
)
price_log_file <- file.path(
  cache_directory,
  paste0("price_log_", analysis_date, "_", download_symbol_hash, ".csv")
)

write_csv_safely <- function(data, path) {
  write_attempt <- capture_call(function() {
    readr::write_csv(data, path, na = "")
    invisible(path)
  })

  tibble::tibble(
    output_file = path,
    status = if (isTRUE(write_attempt$ok)) "written" else "write_failed",
    message = if (isTRUE(write_attempt$ok)) {
      if (length(write_attempt$warnings) == 0L) {
        NA_character_
      } else {
        paste(write_attempt$warnings, collapse = " | ")
      }
    } else {
      write_attempt$error
    }
  )
}

quote_log_write <- write_csv_safely(quote_log, quote_log_file)
price_log_write <- write_csv_safely(price_log, price_log_file)
```

```{r}
#| label: final-universe-tables
#| results: asis

show_table(
  post_retrieval_data_quality |>
    dplyr::select(metric, display_value),
  "Post-retrieval data quality for operating-company candidates"
)
show_table(
  equity_attrition_table |>
    dplyr::select(
      step,
      starting_observations,
      removed_at_step,
      remaining_observations,
      removal_rate_display
    ),
  "Sequential construction of equity_universe_df"
)
show_table(
  final_universe_group_counts,
  "Final operating-company equity universe by benchmark group"
)

retrieval_log_files <- dplyr::bind_rows(
  quote_log_write |>
    dplyr::mutate(log = "Snapshot quote", .before = 1),
  price_log_write |>
    dplyr::mutate(log = "Historical price", .before = 1)
)
show_table(retrieval_log_files, "Retrieval-log output status")
```

# Module 1: Universe scale and security-type breakdown

The full listed-security universe is used in this module. Counts therefore include securities that are intentionally excluded from later operating-company analyses.

```{r}
#| label: module1-summaries
#| results: asis

module1_exchange_summary <- universe_df |>
  dplyr::count(exchange, name = "ticker_count") |>
  dplyr::arrange(dplyr::desc(ticker_count))

module1_security_summary <- universe_df |>
  dplyr::count(security_type, name = "ticker_count") |>
  dplyr::arrange(dplyr::desc(ticker_count))

module1_detail_summary <- universe_df |>
  dplyr::count(
    exchange,
    security_type,
    benchmark_group,
    name = "ticker_count",
    .drop = FALSE
  ) |>
  dplyr::arrange(exchange, security_type, benchmark_group)

module1_coverage <- tibble::tibble(
  metric = "Inferred security-type classification",
  initial_observations = nrow(universe_df),
  valid_observations = sum(!is.na(universe_df$security_type)),
  missing_observations = sum(is.na(universe_df$security_type)),
  missingness_rate = format_percent_1(mean(is.na(universe_df$security_type))),
  sp500_valid = sum(universe_df$is_sp500 & !is.na(universe_df$security_type)),
  broader_market_valid = sum(
    !universe_df$is_sp500 & !is.na(universe_df$security_type)
  )
)

module1_benchmark_match_summary <- tibble::tibble(
  measure = c(
    "S&P 500 constituent symbols",
    "Matched to exchange universe",
    "Not matched to exchange universe",
    "Match rate",
    "Missing security-type classification rate"
  ),
  value = c(
    dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE),
    dplyr::n_distinct(
      universe_df$symbol_yahoo[universe_df$is_sp500],
      na.rm = TRUE
    ),
    nrow(sp500_universe_match),
    safe_rate(
      dplyr::n_distinct(
        universe_df$symbol_yahoo[universe_df$is_sp500],
        na.rm = TRUE
      ),
      dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE)
    ),
    mean(is.na(universe_df$security_type))
  ),
  display_value = c(
    format_count(dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE)),
    format_count(
      dplyr::n_distinct(
        universe_df$symbol_yahoo[universe_df$is_sp500],
        na.rm = TRUE
      )
    ),
    format_count(nrow(sp500_universe_match)),
    format_percent_1(
      safe_rate(
        dplyr::n_distinct(
          universe_df$symbol_yahoo[universe_df$is_sp500],
          na.rm = TRUE
        ),
        dplyr::n_distinct(sp500_df$symbol_yahoo, na.rm = TRUE)
      )
    ),
    format_percent_1(mean(is.na(universe_df$security_type)))
  )
)

show_table(module1_coverage, "Module 1 data coverage")
show_table(module1_exchange_summary, "Ticker records by exchange")
show_table(module1_security_summary, "Ticker records by inferred security type")
show_table(
  module1_detail_summary,
  "Ticker records by exchange, inferred security type, and benchmark membership"
)
show_table(
  module1_benchmark_match_summary |>
    dplyr::select(measure, display_value),
  "Module 1 benchmark matching and classification coverage"
)
```

```{r}
#| label: fig-module1-universe-scale
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Listed-security counts by exchange, inferred security type, and S&P 500 membership. Facets use independent vertical scales to keep smaller security classes readable."

module1_plot_data <- module1_detail_summary |>
  dplyr::mutate(
    benchmark_group = factor(benchmark_group, levels = benchmark_levels),
    exchange = factor(exchange, levels = c("NASDAQ", "NYSE", "AMEX"))
  )

module1_label_data <- module1_plot_data |>
  dplyr::group_by(exchange, security_type) |>
  dplyr::summarise(
    total_tickers = sum(ticker_count, na.rm = TRUE),
    sp500_tickers = sum(
      ticker_count[benchmark_group == "S&P 500"],
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    label = dplyr::if_else(
      sp500_tickers > 0,
      paste0("S&P 500: ", scales::comma(sp500_tickers)),
      ""
    )
  )

module1_plot <- ggplot(
  module1_plot_data,
  aes(x = exchange, y = ticker_count, fill = benchmark_group)
) +
  geom_col(width = 0.72) +
  geom_text(
    data = module1_label_data,
    aes(x = exchange, y = total_tickers, label = label),
    inherit.aes = FALSE,
    vjust = -0.35,
    size = 3.1,
    color = "#123B66"
  ) +
  facet_wrap(
    vars(security_type),
    scales = "free_y",
    ncol = 3
  ) +
  scale_fill_manual(values = benchmark_colors, drop = FALSE) +
  scale_y_continuous(
    labels = scales::label_number(big.mark = ","),
    expand = expansion(mult = c(0, 0.16))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "The listed universe extends far beyond operating-company common stocks",
    subtitle = paste0(
      "Counts include all usable NASDAQ, NYSE, and AMEX directory records; ",
      "security type is inferred from names and ticker patterns."
    ),
    x = NULL,
    y = "Ticker records",
    caption = paste0(
      "Sources: tidyquant::tq_exchange() and tidyquant::tq_index('SP500'). ",
      "Analysis date: ", analysis_date, "."
    )
  ) +
  theme(
    axis.text.x = element_text(face = "bold"),
    panel.spacing = grid::unit(1.1, "lines")
  )

module1_plot
```

# Module 2: Market capitalization and liquidity distributions

Market capitalization uses valid exchange-directory market capitalization first and a Yahoo quote value only as a fallback. Liquidity is the mean of unadjusted daily close multiplied by daily share volume over the most recent 30 valid trading observations. The metric is set to `NA` when fewer than `r minimum_valid_trading_days` valid observations are available.

```{r}
#| label: module2-summaries
#| results: asis

module2_coverage <- metric_coverage_table(
  equity_universe_df,
  metrics = c(
    "market_cap",
    "avg_daily_share_volume_30d",
    "avg_daily_dollar_volume_30d"
  ),
  labels = c(
    "Market capitalization",
    "30-day average daily share volume",
    "30-day average daily dollar volume"
  )
) |>
  dplyr::mutate(
    missingness_rate = format_percent_1(missingness_rate)
  )

module2_summary <- equity_universe_df |>
  dplyr::group_by(benchmark_group, .drop = FALSE) |>
  dplyr::summarise(
    ticker_count = dplyr::n(),
    valid_market_cap_count = sum(is.finite(market_cap)),
    median_market_cap = safe_median(market_cap),
    market_cap_p25 = safe_quantile(market_cap, 0.25),
    market_cap_p75 = safe_quantile(market_cap, 0.75),
    market_cap_missing_rate = mean(!is.finite(market_cap)),
    valid_liquidity_count = sum(is.finite(avg_daily_dollar_volume_30d)),
    median_share_volume_30d = safe_median(avg_daily_share_volume_30d),
    median_dollar_volume_30d = safe_median(avg_daily_dollar_volume_30d),
    share_volume_missing_rate = mean(!is.finite(avg_daily_share_volume_30d)),
    dollar_volume_missing_rate = mean(!is.finite(avg_daily_dollar_volume_30d)),
    .groups = "drop"
  )

module2_summary_display <- module2_summary |>
  dplyr::transmute(
    benchmark_group,
    ticker_count = format_count(ticker_count),
    valid_market_cap_count = format_count(valid_market_cap_count),
    median_market_cap = format_dollar_short(median_market_cap),
    market_cap_p25 = format_dollar_short(market_cap_p25),
    market_cap_p75 = format_dollar_short(market_cap_p75),
    market_cap_missing_rate = format_percent_1(market_cap_missing_rate),
    valid_liquidity_count = format_count(valid_liquidity_count),
    median_share_volume_30d = format_number_short(median_share_volume_30d),
    median_dollar_volume_30d = format_dollar_short(median_dollar_volume_30d),
    share_volume_missing_rate = format_percent_1(share_volume_missing_rate),
    dollar_volume_missing_rate = format_percent_1(dollar_volume_missing_rate)
  )

show_table(module2_coverage, "Module 2 data coverage")
show_table(module2_summary_display, "Module 2 distribution summary by benchmark group")
```

```{r}
#| label: fig-module2-distributions
#| fig-width: 13
#| fig-height: 6.2
#| fig-cap: "Log-scale distributions of market capitalization and 30-day average daily dollar volume."

market_cap_plot_data <- equity_universe_df |>
  dplyr::filter(is.finite(market_cap), market_cap > 0)

liquidity_plot_data <- equity_universe_df |>
  dplyr::filter(
    is.finite(avg_daily_dollar_volume_30d),
    avg_daily_dollar_volume_30d > 0
  )

market_cap_subtitle <- paste0(
  "Log10 scale; ",
  scales::comma(nrow(market_cap_plot_data)),
  " valid observations. Exchange-directory market cap is primary; Yahoo is the fallback."
)

liquidity_subtitle <- paste0(
  "Log10 scale; ",
  scales::comma(nrow(liquidity_plot_data)),
  " valid observations. Minimum history: ",
  minimum_valid_trading_days,
  " of the latest ",
  liquidity_trading_days,
  " trading days."
)

common_module2_caption <- paste0(
  "Sources: NASDAQ exchange directories and Yahoo Finance via tidyquant/quantmod. ",
  "Analysis date: ", analysis_date, "."
)

market_cap_plot <- make_density_plot(
  data = market_cap_plot_data,
  variable = "market_cap",
  title = "Market capitalization",
  subtitle = market_cap_subtitle,
  x_label = "Market capitalization (USD, log10 scale)",
  caption = common_module2_caption,
  log10_axis = TRUE
)

liquidity_plot <- make_density_plot(
  data = liquidity_plot_data,
  variable = "avg_daily_dollar_volume_30d",
  title = "30-day average daily dollar volume",
  subtitle = liquidity_subtitle,
  x_label = "Average daily dollar volume (USD, log10 scale)",
  caption = common_module2_caption,
  log10_axis = TRUE
)

module2_combined_plot <-
  (market_cap_plot | liquidity_plot) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

module2_combined_plot
```

# Module 3: Fundamental and profitability profiling

The snapshot quote route is used for trailing P/E and price-to-book. P/E plotting retains positive finite observations from 0 to 100. Price-to-book plotting retains positive finite observations from 0 to 20. Values outside these display bounds remain in the stored data and are counted below; the underlying fields are not winsorized.

```{r}
#| label: module3-summaries
#| results: asis

module3_summary <- equity_universe_df |>
  dplyr::group_by(benchmark_group, .drop = FALSE) |>
  dplyr::summarise(
    ticker_count = dplyr::n(),
    valid_pe_count = sum(is.finite(trailing_pe) & trailing_pe > 0),
    pe_coverage_rate = mean(is.finite(trailing_pe) & trailing_pe > 0),
    median_pe = safe_median(trailing_pe[is.finite(trailing_pe) & trailing_pe > 0]),
    pe_p25 = safe_quantile(
      trailing_pe[is.finite(trailing_pe) & trailing_pe > 0],
      0.25
    ),
    pe_p75 = safe_quantile(
      trailing_pe[is.finite(trailing_pe) & trailing_pe > 0],
      0.75
    ),
    pe_above_100 = sum(is.finite(trailing_pe) & trailing_pe > 100),
    valid_price_to_book_count = sum(
      is.finite(price_to_book) & price_to_book > 0
    ),
    price_to_book_coverage_rate = mean(
      is.finite(price_to_book) & price_to_book > 0
    ),
    median_price_to_book = safe_median(
      price_to_book[is.finite(price_to_book) & price_to_book > 0]
    ),
    price_to_book_p25 = safe_quantile(
      price_to_book[is.finite(price_to_book) & price_to_book > 0],
      0.25
    ),
    price_to_book_p75 = safe_quantile(
      price_to_book[is.finite(price_to_book) & price_to_book > 0],
      0.75
    ),
    price_to_book_above_20 = sum(
      is.finite(price_to_book) & price_to_book > 20
    ),
    operating_margin_coverage_rate = mean(is.finite(operating_margin)),
    .groups = "drop"
  )

module3_summary_display <- module3_summary |>
  dplyr::transmute(
    benchmark_group,
    ticker_count = format_count(ticker_count),
    valid_pe_count = format_count(valid_pe_count),
    pe_coverage_rate = format_percent_1(pe_coverage_rate),
    median_pe = dplyr::if_else(
      is.na(median_pe), "NA", scales::number(median_pe, accuracy = 0.1)
    ),
    pe_iqr = dplyr::if_else(
      is.na(pe_p25) | is.na(pe_p75),
      "NA",
      paste0(
        scales::number(pe_p25, accuracy = 0.1),
        " to ",
        scales::number(pe_p75, accuracy = 0.1)
      )
    ),
    pe_above_100 = format_count(pe_above_100),
    valid_price_to_book_count = format_count(valid_price_to_book_count),
    price_to_book_coverage_rate = format_percent_1(
      price_to_book_coverage_rate
    ),
    median_price_to_book = dplyr::if_else(
      is.na(median_price_to_book),
      "NA",
      scales::number(median_price_to_book, accuracy = 0.1)
    ),
    price_to_book_iqr = dplyr::if_else(
      is.na(price_to_book_p25) | is.na(price_to_book_p75),
      "NA",
      paste0(
        scales::number(price_to_book_p25, accuracy = 0.1),
        " to ",
        scales::number(price_to_book_p75, accuracy = 0.1)
      )
    ),
    price_to_book_above_20 = format_count(price_to_book_above_20),
    operating_margin_coverage_rate = format_percent_1(
      operating_margin_coverage_rate
    )
  )

module3_coverage <- dplyr::bind_rows(
  tibble::tibble(
    metric = "Trailing P/E: positive finite observations",
    initial_observations = nrow(equity_universe_df),
    valid_observations = sum(
      is.finite(equity_universe_df$trailing_pe) &
        equity_universe_df$trailing_pe > 0
    ),
    missing_observations = sum(
      !(is.finite(equity_universe_df$trailing_pe) &
          equity_universe_df$trailing_pe > 0)
    ),
    missingness_rate = mean(
      !(is.finite(equity_universe_df$trailing_pe) &
          equity_universe_df$trailing_pe > 0)
    ),
    sp500_valid = sum(
      equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$trailing_pe) &
        equity_universe_df$trailing_pe > 0
    ),
    broader_market_valid = sum(
      !equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$trailing_pe) &
        equity_universe_df$trailing_pe > 0
    )
  ),
  tibble::tibble(
    metric = "Price-to-book: positive finite observations",
    initial_observations = nrow(equity_universe_df),
    valid_observations = sum(
      is.finite(equity_universe_df$price_to_book) &
        equity_universe_df$price_to_book > 0
    ),
    missing_observations = sum(
      !(is.finite(equity_universe_df$price_to_book) &
          equity_universe_df$price_to_book > 0)
    ),
    missingness_rate = mean(
      !(is.finite(equity_universe_df$price_to_book) &
          equity_universe_df$price_to_book > 0)
    ),
    sp500_valid = sum(
      equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$price_to_book) &
        equity_universe_df$price_to_book > 0
    ),
    broader_market_valid = sum(
      !equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$price_to_book) &
        equity_universe_df$price_to_book > 0
    )
  ),
  tibble::tibble(
    metric = "Operating margin",
    initial_observations = nrow(equity_universe_df),
    valid_observations = sum(is.finite(equity_universe_df$operating_margin)),
    missing_observations = sum(!is.finite(equity_universe_df$operating_margin)),
    missingness_rate = mean(!is.finite(equity_universe_df$operating_margin)),
    sp500_valid = sum(
      equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$operating_margin)
    ),
    broader_market_valid = sum(
      !equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$operating_margin)
    )
  )
) |>
  dplyr::mutate(missingness_rate = format_percent_1(missingness_rate))

show_table(module3_coverage, "Module 3 data coverage")
show_table(module3_summary_display, "Module 3 valuation summary by benchmark group")
```

```{r}
#| label: fig-module3-valuations
#| fig-width: 13
#| fig-height: 6.2
#| fig-cap: "Trailing P/E and price-to-book distributions using disclosed plotting bounds."

pe_plot_data <- equity_universe_df |>
  dplyr::filter(
    is.finite(trailing_pe),
    trailing_pe > 0,
    trailing_pe <= 100
  )

price_to_book_plot_data <- equity_universe_df |>
  dplyr::filter(
    is.finite(price_to_book),
    price_to_book > 0,
    price_to_book <= 20
  )

pe_positive_count <- sum(
  is.finite(equity_universe_df$trailing_pe) &
    equity_universe_df$trailing_pe > 0
)
price_to_book_positive_count <- sum(
  is.finite(equity_universe_df$price_to_book) &
    equity_universe_df$price_to_book > 0
)
pe_coverage_overall <- safe_rate(pe_positive_count, nrow(equity_universe_df))
price_to_book_coverage_overall <- safe_rate(
  price_to_book_positive_count,
  nrow(equity_universe_df)
)
pe_above_bound <- sum(
  is.finite(equity_universe_df$trailing_pe) &
    equity_universe_df$trailing_pe > 100
)
price_to_book_above_bound <- sum(
  is.finite(equity_universe_df$price_to_book) &
    equity_universe_df$price_to_book > 20
)

module3_caption <- paste0(
  "Source: Yahoo Finance snapshot fields via quantmod::getQuote(). ",
  "Analysis date: ", analysis_date, ". Values outside the display bounds remain in the stored data."
)

pe_plot <- make_density_plot(
  data = pe_plot_data,
  variable = "trailing_pe",
  title = "Trailing P/E ratio",
  subtitle = paste0(
    "Positive-finite coverage: ",
    format_percent_1(pe_coverage_overall),
    "; display range: 0 to 100; included: ",
    scales::comma(nrow(pe_plot_data)),
    "; above 100: ",
    scales::comma(pe_above_bound),
    "."
  ),
  x_label = "Trailing P/E ratio",
  caption = module3_caption,
  lower_bound = 0,
  upper_bound = 100
) +
  scale_x_continuous(breaks = seq(0, 100, by = 20))

price_to_book_plot <- make_density_plot(
  data = price_to_book_plot_data,
  variable = "price_to_book",
  title = "Price-to-book ratio",
  subtitle = paste0(
    "Positive-finite coverage: ",
    format_percent_1(price_to_book_coverage_overall),
    "; display range: 0 to 20; included: ",
    scales::comma(nrow(price_to_book_plot_data)),
    "; above 20: ",
    scales::comma(price_to_book_above_bound),
    "."
  ),
  x_label = "Price-to-book ratio",
  caption = module3_caption,
  lower_bound = 0,
  upper_bound = 20
) +
  scale_x_continuous(breaks = seq(0, 20, by = 5))

module3_combined_plot <-
  (pe_plot | price_to_book_plot) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

module3_combined_plot
```

```{r}
#| label: operating-margin-note
#| results: asis

operating_margin_coverage <- if (nrow(equity_universe_df) == 0L) {
  NA_real_
} else {
  mean(is.finite(equity_universe_df$operating_margin))
}

if (is.na(operating_margin_coverage) || operating_margin_coverage < 0.20) {
  cat(
    paste0(
      "::: {.callout-note}\n",
      "## Operating margin omitted\n\n",
      "The selected documented `quantmod::getQuote()` Yahoo snapshot fields do not provide operating margin. ",
      "No API key or undocumented scraping endpoint is used, so operating-margin coverage is ",
      format_percent_1(operating_margin_coverage),
      " and no operating-margin plot is produced. The report continues without substituting values.\n",
      ":::\n"
    )
  )
}
```

# Module 4: Sector concentration and skew

For this module, the **broad equity universe** means the complete final operating-company equity universe, including current S&P 500 constituents. Ticker-count share is the equal-weighted listing-breadth view. Market-cap share weights that same broad universe by valid market capitalization. S&P 500 weights come directly from the current constituent table and are re-normalized over positive constituent weights.

```{r}
#| label: module4-calculations

sp500_positive_weight_total <- sum(
  sp500_df$sp500_weight[
    is.finite(sp500_df$sp500_weight) &
      sp500_df$sp500_weight > 0
  ],
  na.rm = TRUE
)

sp500_missing_sector_weight <- sum(
  sp500_df$sp500_weight[
    is.finite(sp500_df$sp500_weight) &
      sp500_df$sp500_weight > 0 &
      is.na(sp500_df$sp500_sector_clean)
  ],
  na.rm = TRUE
)

sp500_sector_weights <- sp500_df |>
  dplyr::filter(
    !is.na(sp500_sector_clean),
    is.finite(sp500_weight),
    sp500_weight > 0
  ) |>
  dplyr::group_by(sector = sp500_sector_clean) |>
  dplyr::summarise(
    sp500_weight_pct = sum(sp500_weight, na.rm = TRUE),
    .groups = "drop"
  )

sp500_sector_weight_sum <- sum(
  sp500_sector_weights$sp500_weight_pct,
  na.rm = TRUE
)

if (sp500_sector_weight_sum > 0) {
  sp500_sector_weights <- sp500_sector_weights |>
    dplyr::mutate(
      sp500_weight_pct = sp500_weight_pct / sp500_sector_weight_sum
    )
}

broader_market_sector_base <- equity_universe_df

broader_market_total_count <- nrow(broader_market_sector_base)
broader_market_classified_count <- sum(
  !is.na(broader_market_sector_base$sector_clean)
)
broader_market_missing_sector_count <- sum(
  is.na(broader_market_sector_base$sector_clean)
)

broader_market_total_cap <- sum(
  broader_market_sector_base$market_cap[
    is.finite(broader_market_sector_base$market_cap) &
      broader_market_sector_base$market_cap > 0
  ],
  na.rm = TRUE
)

broader_market_missing_sector_cap <- sum(
  broader_market_sector_base$market_cap[
    is.finite(broader_market_sector_base$market_cap) &
      broader_market_sector_base$market_cap > 0 &
      is.na(broader_market_sector_base$sector_clean)
  ],
  na.rm = TRUE
)

broader_sector_counts <- broader_market_sector_base |>
  dplyr::filter(!is.na(sector_clean)) |>
  dplyr::count(sector = sector_clean, name = "ticker_count")

if (broader_market_classified_count > 0L) {
  broader_sector_counts <- broader_sector_counts |>
    dplyr::mutate(
      broad_ticker_count_pct = ticker_count / broader_market_classified_count
    )
} else {
  broader_sector_counts <- broader_sector_counts |>
    dplyr::mutate(broad_ticker_count_pct = NA_real_)
}

broader_sector_market_cap <- broader_market_sector_base |>
  dplyr::filter(
    !is.na(sector_clean),
    is.finite(market_cap),
    market_cap > 0
  ) |>
  dplyr::group_by(sector = sector_clean) |>
  dplyr::summarise(
    sector_market_cap = sum(market_cap, na.rm = TRUE),
    .groups = "drop"
  )

classified_broader_market_cap <- sum(
  broader_sector_market_cap$sector_market_cap,
  na.rm = TRUE
)

if (classified_broader_market_cap > 0) {
  broader_sector_market_cap <- broader_sector_market_cap |>
    dplyr::mutate(
      broad_market_cap_pct = sector_market_cap / classified_broader_market_cap
    )
} else {
  broader_sector_market_cap <- broader_sector_market_cap |>
    dplyr::mutate(broad_market_cap_pct = NA_real_)
}

sector_comparison_table <- sp500_sector_weights |>
  dplyr::full_join(
    broader_sector_counts |>
      dplyr::select(sector, broad_ticker_count_pct),
    by = "sector"
  ) |>
  dplyr::full_join(
    broader_sector_market_cap |>
      dplyr::select(sector, broad_market_cap_pct),
    by = "sector"
  )

if (sp500_sector_weight_sum > 0) {
  sector_comparison_table$sp500_weight_pct <- tidyr::replace_na(
    sector_comparison_table$sp500_weight_pct,
    0
  )
}
if (broader_market_classified_count > 0L) {
  sector_comparison_table$broad_ticker_count_pct <- tidyr::replace_na(
    sector_comparison_table$broad_ticker_count_pct,
    0
  )
}
if (classified_broader_market_cap > 0) {
  sector_comparison_table$broad_market_cap_pct <- tidyr::replace_na(
    sector_comparison_table$broad_market_cap_pct,
    0
  )
}

sector_comparison_table <- sector_comparison_table |>
  dplyr::mutate(
    sp500_minus_broad_market_cap =
      sp500_weight_pct - broad_market_cap_pct,
    sp500_minus_broad_ticker_count =
      sp500_weight_pct - broad_ticker_count_pct
  ) |>
  dplyr::arrange(
    dplyr::desc(sp500_weight_pct),
    dplyr::desc(broad_market_cap_pct)
  )

module4_metric_coverage <- dplyr::bind_rows(
  tibble::tibble(
    metric = "Standardized sector classification",
    initial_observations = nrow(equity_universe_df),
    valid_observations = sum(!is.na(equity_universe_df$sector_clean)),
    missing_observations = sum(is.na(equity_universe_df$sector_clean)),
    missingness_rate = mean(is.na(equity_universe_df$sector_clean)),
    sp500_valid = sum(
      equity_universe_df$is_sp500 & !is.na(equity_universe_df$sector_clean)
    ),
    broader_market_valid = sum(
      !equity_universe_df$is_sp500 & !is.na(equity_universe_df$sector_clean)
    )
  ),
  tibble::tibble(
    metric = "Positive market capitalization",
    initial_observations = nrow(equity_universe_df),
    valid_observations = sum(
      is.finite(equity_universe_df$market_cap) &
        equity_universe_df$market_cap > 0
    ),
    missing_observations = sum(
      !(is.finite(equity_universe_df$market_cap) &
          equity_universe_df$market_cap > 0)
    ),
    missingness_rate = mean(
      !(is.finite(equity_universe_df$market_cap) &
          equity_universe_df$market_cap > 0)
    ),
    sp500_valid = sum(
      equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$market_cap) &
        equity_universe_df$market_cap > 0
    ),
    broader_market_valid = sum(
      !equity_universe_df$is_sp500 &
        is.finite(equity_universe_df$market_cap) &
        equity_universe_df$market_cap > 0
    )
  )
) |>
  dplyr::mutate(missingness_rate = format_percent_1(missingness_rate))

sector_coverage_table <- tibble::tibble(
  measure = c(
    "S&P 500 positive constituent weight represented by classified sectors",
    "S&P 500 weight omitted for missing sector",
    "Broad-universe ticker records represented by classified sectors",
    "Broad-universe ticker records omitted for missing sector",
    "Broad-universe market capitalization represented by classified sectors",
    "Broad-universe market capitalization omitted for missing sector"
  ),
  value = c(
    sp500_sector_weight_sum,
    sp500_missing_sector_weight,
    broader_market_classified_count,
    broader_market_missing_sector_count,
    classified_broader_market_cap,
    broader_market_missing_sector_cap
  ),
  display_value = c(
    format_percent_1(sp500_sector_weight_sum),
    format_percent_1(sp500_missing_sector_weight),
    format_count(broader_market_classified_count),
    paste0(
      format_count(broader_market_missing_sector_count),
      " (",
      format_percent_1(
        safe_rate(
          broader_market_missing_sector_count,
          broader_market_total_count
        )
      ),
      ")"
    ),
    format_dollar_short(classified_broader_market_cap),
    paste0(
      format_dollar_short(broader_market_missing_sector_cap),
      " (",
      format_percent_1(
        safe_rate(
          broader_market_missing_sector_cap,
          broader_market_total_cap
        )
      ),
      ")"
    )
  )
)

sector_comparison_display <- sector_comparison_table |>
  dplyr::transmute(
    sector,
    sp500_weight_pct = format_percent_1(sp500_weight_pct),
    broad_ticker_count_pct = format_percent_1(broad_ticker_count_pct),
    broad_market_cap_pct = format_percent_1(broad_market_cap_pct),
    sp500_minus_broad_market_cap = format_percent_1(
      sp500_minus_broad_market_cap
    ),
    sp500_minus_broad_ticker_count = format_percent_1(
      sp500_minus_broad_ticker_count
    )
  )
```

```{r}
#| label: module4-summaries
#| results: asis

show_table(module4_metric_coverage, "Module 4 data coverage")
show_table(sector_coverage_table, "Module 4 sector coverage and omitted weight")
show_table(
  sector_comparison_display,
  "Sector concentration: S&P 500 versus the broad equity universe"
)
```

```{r}
#| label: fig-module4-sector-concentration
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Sector weights for the S&P 500 compared with equal-weighted ticker breadth and market-cap-weighted concentration in the broad operating-company universe."

sector_plot_data <- sector_comparison_table |>
  dplyr::select(
    sector,
    sp500_weight_pct,
    broad_ticker_count_pct,
    broad_market_cap_pct
  ) |>
  tidyr::pivot_longer(
    cols = -sector,
    names_to = "measure",
    values_to = "sector_share"
  ) |>
  dplyr::mutate(
    measure = dplyr::recode(
      measure,
      sp500_weight_pct = "S&P 500 constituent weight",
      broad_ticker_count_pct = "Broad-universe ticker-count share",
      broad_market_cap_pct = "Broad-universe market-cap share"
    ),
    measure = factor(
      measure,
      levels = names(sector_measure_colors)
    )
  ) |>
  dplyr::filter(is.finite(sector_share))

sector_order <- sector_comparison_table |>
  dplyr::arrange(
    dplyr::desc(sp500_weight_pct),
    dplyr::desc(broad_market_cap_pct)
  ) |>
  dplyr::pull(sector)

sector_plot_data <- sector_plot_data |>
  dplyr::mutate(
    sector = factor(sector, levels = rev(unique(sector_order)))
  )

if (nrow(sector_plot_data) == 0L) {
  module4_plot <- empty_plot(
    "Sector concentration and skew",
    "No valid sector-share series are available.",
    paste0("Analysis date: ", analysis_date, ".")
  )
} else {
  module4_plot <- ggplot(
    sector_plot_data,
    aes(x = sector, y = sector_share, fill = measure)
  ) +
    geom_col(
      position = position_dodge(width = 0.78),
      width = 0.70
    ) +
    coord_flip() +
    scale_fill_manual(values = sector_measure_colors, drop = FALSE) +
    scale_y_continuous(
      labels = scales::label_percent(accuracy = 1),
      expand = expansion(mult = c(0, 0.05))
    ) +
    labs(
      title = "Benchmark weights emphasize a different sector mix than listing breadth",
      subtitle = paste0(
        "Ticker-count share is equal weighted; market-cap share weights the full operating-company universe by valid market cap."
      ),
      x = NULL,
      y = "Sector share",
      caption = paste0(
        "Sources: tidyquant::tq_index('SP500') and the cleaned exchange universe. ",
        "Broad-universe measures include current S&P 500 constituents. Analysis date: ",
        analysis_date,
        "."
      )
    )
}

module4_plot
```

# Consolidated data coverage

```{r}
#| label: consolidated-missingness

missingness_row <- function(scope, data, variable, missing_condition = NULL) {
  if (!is.data.frame(data)) {
    return(tibble::tibble())
  }

  total <- nrow(data)

  if (is.null(missing_condition)) {
    values <- data[[variable]]
    missing <- is.na(values)
  } else {
    missing <- missing_condition(data)
    missing[is.na(missing)] <- TRUE
  }

  tibble::tibble(
    scope = scope,
    variable = variable,
    total_observations = total,
    missing_observations = sum(missing),
    valid_observations = total - sum(missing),
    missingness_rate = safe_rate(sum(missing), total)
  )
}

consolidated_missingness <- dplyr::bind_rows(
  missingness_row("Full listed-security universe", universe_df, "exchange"),
  missingness_row("Full listed-security universe", universe_df, "security_type"),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "market_cap",
    function(x) !is.finite(x$market_cap) | x$market_cap <= 0
  ),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "sector_clean"
  ),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "industry_clean"
  ),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "historical_price_data",
    function(x) !x$historical_data_available
  ),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "avg_daily_share_volume_30d",
    function(x) !is.finite(x$avg_daily_share_volume_30d)
  ),
  missingness_row(
    "Download-candidate equity universe",
    analysis_enriched_df,
    "avg_daily_dollar_volume_30d",
    function(x) !is.finite(x$avg_daily_dollar_volume_30d)
  ),
  missingness_row(
    "Final operating-company equity universe",
    equity_universe_df,
    "trailing_pe",
    function(x) !is.finite(x$trailing_pe)
  ),
  missingness_row(
    "Final operating-company equity universe",
    equity_universe_df,
    "price_to_book",
    function(x) !is.finite(x$price_to_book)
  ),
  missingness_row(
    "Final operating-company equity universe",
    equity_universe_df,
    "operating_margin",
    function(x) !is.finite(x$operating_margin)
  )
)

consolidated_missingness_display <- consolidated_missingness |>
  dplyr::mutate(
    total_observations = format_count(total_observations),
    missing_observations = format_count(missing_observations),
    valid_observations = format_count(valid_observations),
    missingness_rate = format_percent_1(missingness_rate)
  )
```

```{r}
#| label: consolidated-missingness-table
#| results: asis

show_table(
  consolidated_missingness_display,
  "Consolidated data coverage and missingness"
)
```

# Analytical summary

```{r}
#| label: final-summary-calculations

get_group_value <- function(data, group_name, column_name) {
  matching_rows <- as.character(data$benchmark_group) == group_name
  if (!any(matching_rows)) {
    return(NA_real_)
  }
  data[[column_name]][which(matching_rows)[1L]]
}

full_listed_count <- nrow(universe_df)
final_equity_count <- nrow(equity_universe_df)
final_sp500_count <- sum(equity_universe_df$is_sp500, na.rm = TRUE)
final_broader_count <- sum(!equity_universe_df$is_sp500, na.rm = TRUE)

sp500_median_market_cap <- get_group_value(
  module2_summary,
  "S&P 500",
  "median_market_cap"
)
broader_median_market_cap <- get_group_value(
  module2_summary,
  "Broader Market",
  "median_market_cap"
)
sp500_median_dollar_volume <- get_group_value(
  module2_summary,
  "S&P 500",
  "median_dollar_volume_30d"
)
broader_median_dollar_volume <- get_group_value(
  module2_summary,
  "Broader Market",
  "median_dollar_volume_30d"
)

sp500_pe_coverage <- get_group_value(
  module3_summary,
  "S&P 500",
  "pe_coverage_rate"
)
broader_pe_coverage <- get_group_value(
  module3_summary,
  "Broader Market",
  "pe_coverage_rate"
)

sp500_pb_coverage <- get_group_value(
  module3_summary,
  "S&P 500",
  "price_to_book_coverage_rate"
)
broader_pb_coverage <- get_group_value(
  module3_summary,
  "Broader Market",
  "price_to_book_coverage_rate"
)

top_sp500_sector <- sector_comparison_table |>
  dplyr::filter(is.finite(sp500_weight_pct)) |>
  dplyr::slice_max(sp500_weight_pct, n = 1L, with_ties = FALSE)

top_broad_market_cap_sector <- sector_comparison_table |>
  dplyr::filter(is.finite(broad_market_cap_pct)) |>
  dplyr::slice_max(broad_market_cap_pct, n = 1L, with_ties = FALSE)

top_sp500_sector_name <- if (nrow(top_sp500_sector) == 0L) {
  "unavailable"
} else {
  top_sp500_sector$sector[[1]]
}
top_sp500_sector_weight <- if (nrow(top_sp500_sector) == 0L) {
  NA_real_
} else {
  top_sp500_sector$sp500_weight_pct[[1]]
}

top_broad_sector_name <- if (nrow(top_broad_market_cap_sector) == 0L) {
  "unavailable"
} else {
  top_broad_market_cap_sector$sector[[1]]
}
top_broad_sector_weight <- if (nrow(top_broad_market_cap_sector) == 0L) {
  NA_real_
} else {
  top_broad_market_cap_sector$broad_market_cap_pct[[1]]
}

final_summary_table <- tibble::tibble(
  measure = c(
    "Full listed-security records",
    "Final operating-company equity symbols",
    "Final S&P 500 symbols",
    "Final broader-market symbols",
    "S&P 500 median market capitalization",
    "Broader-market median market capitalization",
    "S&P 500 median 30-day dollar volume",
    "Broader-market median 30-day dollar volume",
    "S&P 500 positive P/E coverage",
    "Broader-market positive P/E coverage",
    "S&P 500 positive price-to-book coverage",
    "Broader-market positive price-to-book coverage",
    "Largest S&P 500 sector by current weight",
    "Largest broad-universe sector by market cap"
  ),
  value = c(
    format_count(full_listed_count),
    format_count(final_equity_count),
    format_count(final_sp500_count),
    format_count(final_broader_count),
    format_dollar_short(sp500_median_market_cap),
    format_dollar_short(broader_median_market_cap),
    format_dollar_short(sp500_median_dollar_volume),
    format_dollar_short(broader_median_dollar_volume),
    format_percent_1(sp500_pe_coverage),
    format_percent_1(broader_pe_coverage),
    format_percent_1(sp500_pb_coverage),
    format_percent_1(broader_pb_coverage),
    paste0(
      top_sp500_sector_name,
      " (",
      format_percent_1(top_sp500_sector_weight),
      ")"
    ),
    paste0(
      top_broad_sector_name,
      " (",
      format_percent_1(top_broad_sector_weight),
      ")"
    )
  )
)
```

```{r}
#| label: final-summary-table
#| results: asis

show_table(final_summary_table, "Key findings from the rendered data")
```

The exchange directories contain `r format_count(full_listed_count)` usable listed-security records, while the stricter operating-company analysis contains `r format_count(final_equity_count)` unique symbols. The difference reflects both the presence of non-operating-company securities and the explicit data-quality requirements applied before Modules 2 through 4.

Within the final analytical sample, the median market capitalization is `r format_dollar_short(sp500_median_market_cap)` for S&P 500 constituents and `r format_dollar_short(broader_median_market_cap)` for non-S&P equities. Median 30-day average daily dollar volume is `r format_dollar_short(sp500_median_dollar_volume)` for S&P 500 constituents and `r format_dollar_short(broader_median_dollar_volume)` for the broader market. These are descriptive associations with current index membership; they are not causal estimates of an index-inclusion effect.

Positive trailing P/E coverage is `r format_percent_1(sp500_pe_coverage)` for the S&P 500 group and `r format_percent_1(broader_pe_coverage)` for the broader-market group. Positive price-to-book coverage is `r format_percent_1(sp500_pb_coverage)` and `r format_percent_1(broader_pb_coverage)`, respectively. Differences in the plotted valuation distributions should therefore be interpreted together with these coverage rates.

The largest current S&P 500 sector by constituent weight is **`r top_sp500_sector_name`** at `r format_percent_1(top_sp500_sector_weight)`. The largest sector in the full operating-company universe by market capitalization is **`r top_broad_sector_name`** at `r format_percent_1(top_broad_sector_weight)`. Module 4 separates equal-weighted listing breadth from market-cap concentration so that these distinct concepts are not conflated.

## Limitations

- **Current-constituent and survivorship bias:** The benchmark table represents current constituents, not historical membership. Results do not describe firms that previously left the index or disappeared from the exchange directories.
- **Point-in-time limitations:** Exchange metadata, index weights, quotes, and recent prices are retrieved from current endpoints and may not share exactly the same timestamp.
- **Current rather than historical index weights:** S&P 500 sector weights are based on the current constituent snapshot and should not be interpreted as historical weights.
- **Yahoo Finance availability:** Symbols, quote fields, and historical prices may be missing, delayed, stale, rate limited, or temporarily unavailable. Failed symbols are logged rather than silently discarded.
- **Heuristic security-type classification:** The exchange feed does not supply a single standardized security-type field. Name and ticker-pattern rules may misclassify unusual securities.
- **Symbol normalization:** Class-share and preferred-share translations are practical Yahoo mappings, not a universal symbology standard. The original exchange symbol is retained for auditability.
- **Exchange metadata quality:** Market capitalization, sector, industry, and last-sale fields can be stale or absent. Exchange market capitalization is used first, with Yahoo market capitalization as a documented fallback.
- **Missing fundamentals:** P/E and price-to-book are available only for a subset of securities. Operating margin is omitted because the selected documented snapshot route does not provide it reliably.
- **Universe definition:** Raw listed-security counts include funds, preferreds, warrants, rights, units, SPACs, and other instruments. Modules 2 through 4 instead use the stricter common-equity/ADR universe defined in the attrition table.
- **Descriptive analysis:** Comparisons describe securities associated with current S&P 500 membership. They do not establish that membership causes differences in size, liquidity, valuation, or sector composition.

## Reproducibility notes

- Set `refresh_data <- TRUE` to force new exchange, index, quote, and historical-price requests.
- With `refresh_data <- FALSE`, exact caches are reused. Exchange and index retrieval may use the most recent prior cache if the current request fails; the retrieval manifest discloses that fallback.
- Quote and historical-price cache names incorporate the analysis date and an MD5 signature of the requested symbol set, preventing silent reuse for a materially different universe.
- A quote or historical-price result with no usable market data is returned for diagnostics but deliberately not stored as a reusable cache.
- Full condition and failed-symbol logs are written as CSV files in `r cache_directory` when that directory is writable; the report displays each write status.
- `tq_exchange()` documentation: <https://business-science.github.io/tidyquant/reference/tq_index.html>
- `tq_get()` documentation: <https://business-science.github.io/tidyquant/reference/tq_get.html>
- `quantmod::getQuote()` source and field map: <https://github.com/joshuaulrich/quantmod/blob/main/R/getQuote.R>

```{r}
#| label: session-information
#| code-fold: true
#| code-summary: "Show session information"

sessionInfo()
```
